---
title: "Metastasis-Associated Gene Ontology Query and Gene Set Construction"
author: "Gleison M. Azevedo"
format: html
execute:
  cache: true
  echo: true
  message: false
  warning: false
---

## Package Management

```{r pkg-install}
#| cache: false

# Automatically detect and install any missing CRAN packages before loading.
required_packages <- c(
  "tidyr", "dplyr", "httr", "vroom", "here", "UpSetR", "tools", "svglite"
)

missing <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing) > 0) {
  install.packages(missing, repos = "https://cloud.r-project.org")
}

library(tidyr)
library(dplyr)
library(httr)
library(vroom)
library(here)
library(UpSetR)
library(tools)
library(svglite)
```

## Input Data Integrity Verification

```{r checksum}
#| cache: false

# Verify MD5 checksums for all input files loaded from assets/ to ensure
# data integrity and guard against unintended modifications or corruption.

expected_checksums <- c(
  "assets/Detatch_and_Dissemination.csv"   = "834eb00d5fd05ebc8b4340ca8e630b0e",
  "assets/metastasis_supressors_nature.csv" = "6f1717e97a277fb0fed475e92647d214"
)

for (file_path in names(expected_checksums)) {
  abs_path <- here(file_path)
  observed <- unname(md5sum(abs_path))
  if (observed != expected_checksums[file_path]) {
    stop(
      sprintf(
        "Checksum mismatch for '%s': expected %s, got %s",
        file_path, expected_checksums[file_path], observed
      )
    )
  }
}

message("All input file checksums verified successfully.")
```

## Helper Functions

```{r functions}
#' Query Gene Ontology for genes annotated to a specific GO term.
#'
#' Retrieves human protein-coding gene symbols annotated under a given GO ID
#' using the AmiGO 2 SOLR API, restricted to experimental evidence codes.
#'
#' @param GO_id Character. A Gene Ontology identifier (e.g., "GO:0045123").
#' @return Character vector of unique gene symbols, or character(0) on failure.
query_GO <- function(GO_id) {
  url <- paste0(
    "https://golr-aux.geneontology.io/solr/select?defType=edismax",
    "&qt=standard&indent=on&wt=csv&rows=100000&start=0",
    "&fl=bioentity_label&facet=true&facet.mincount=1&facet.sort=count",
    "&json.nl=arrarr&facet.limit=25&hl=true",
    "&hl.simple.pre=%3Cem%20class%3D%22hilite%22%3E&hl.snippets=1000",
    "&csv.encapsulator=&csv.separator=%09&csv.header=false",
    "&csv.mv.separator=%7C",
    "&fq=document_category:%22annotation%22",
    "&fq=isa_partof_closure:%22", GO_id, "%22",
    "&fq=taxon_subset_closure_label:%22Homo%20sapiens%22",
    "&fq=type:%22protein%22",
    "&fq=evidence_subset_closure_label:%22experimental%20evidence%22",
    "&facet.field=aspect&facet.field=taxon_subset_closure_label",
    "&facet.field=type&facet.field=evidence_subset_closure_label",
    "&facet.field=regulates_closure_label",
    "&facet.field=isa_partof_closure_label",
    "&facet.field=annotation_class_label&facet.field=qualifier",
    "&facet.field=annotation_extension_class_closure_label",
    "&facet.field=assigned_by&facet.field=panther_family_label",
    "&q=*%3A*"
  )

  response <- tryCatch(
    {
      vroom(url, delim = "\t", col_names = FALSE, show_col_types = FALSE)
    },
    error = function(e) {
      return(data.frame())
    }
  )

  if (ncol(response) == 0) {
    return(character(0))
  }

  genes <- response[[1]]
  return(unique(as.vector(genes)))
}


#' Batch-query Gene Ontology for a vector of GO IDs.
#'
#' Iterates over a set of GO identifiers and returns a named list mapping
#' each GO ID to its associated gene symbols.
#'
#' @param go_ids Character vector of GO identifiers.
#' @return Named list; names are GO IDs, values are character vectors of genes.
get_genes_by_GOid <- function(go_ids) {
  genes_per_term <- list()
  for (go_id in go_ids) {
    genes_per_term[[go_id]] <- query_GO(go_id)
  }
  return(genes_per_term)
}


#' Convert a GO-to-genes list into a tidy data frame.
#'
#' Maps each gene to its corresponding metastasis signature label using the
#' GO ID as the join key, producing a two-column data frame.
#'
#' @param genesPerTerm Named list from get_genes_by_GOid().
#' @param metastasis_ids Data frame with columns GO_id and Signature.
#' @return Data frame with columns: external_gene_name, Signature.
list_to_df <- function(genesPerTerm, metastasis_ids) {
  result <- data.frame(
    external_gene_name = character(),
    Signature = character(),
    stringsAsFactors = FALSE
  )

  for (go_id in names(genesPerTerm)) {
    genes <- genesPerTerm[[go_id]]
    signature <- metastasis_ids$Signature[metastasis_ids$GO_id == go_id]

    if (length(signature) > 0 && length(genes) > 0) {
      temp <- data.frame(
        external_gene_name = genes,
        Signature = rep(signature, length(genes)),
        stringsAsFactors = FALSE
      )
      result <- rbind(result, temp)
    }
  }

  return(result)
}
```

## Data Loading

The metastasis-associated biological processes are sourced from a curated table
mapping GO identifiers to hallmark signatures of the metastatic cascade. The
signature labels are cleaned by removing a fixed-length prefix, and the GO ID
column is split into the identifier and its textual description.

A second dataset provides a list of known metastasis suppressor genes (MSGs)
compiled from the literature (Rodrigues et al., *Nature Reviews Cancer*, 2024).

```{r data-load}
# Curated GO biological process identifiers for metastasis hallmarks
metastasis_ids <- vroom::vroom(
  here("assets/Detatch_and_Dissemination.csv"),
  show_col_types = FALSE
)
metastasis_ids$Signature <- substr(
  metastasis_ids$Signature, 12, nchar(metastasis_ids$Signature)
)
metastasis_ids <- separate(
  metastasis_ids,
  GO_id,
  into = c("GO_id", "Description"),
  sep = 10,
  extra = "merge"
)

# Metastasis suppressor genes from literature
metastasis_suppressors <- vroom::vroom(
  here("assets/metastasis_supressors_nature.csv"),
  delim = "\t",
  show_col_types = FALSE
)
metastasis_suppressors <- metastasis_suppressors %>%
  dplyr::select(external_gene_name = `GeneSymb`) %>%
  dplyr::mutate(Signature = "metastasis suppressor")

head(metastasis_ids)
```

## GO API Query and Gene Set Assembly

Each GO identifier from the metastasis hallmark table is queried against the
Gene Ontology AmiGO 2 SOLR endpoint to retrieve experimentally supported
human gene annotations. The resulting gene lists are then merged with the
curated metastasis suppressor gene set to form a unified gene–signature table.

```{r go-query}
#| cache: true

# Query the GO database for genes associated with each GO term
genesPerTerm <- get_genes_by_GOid(metastasis_ids$GO_id)

# Assemble gene-to-signature data frame and merge with suppressor genes
result <- list_to_df(genesPerTerm, metastasis_ids) %>%
  bind_rows(metastasis_suppressors) %>%
  unique()

# Prepare Table 1: unique GO annotations (without Hallmark grouping)
table1 <- dplyr::select(metastasis_ids, -Hallmark) %>%
  unique()

# --- Save outputs to results/metastasis_genes/ ---
output_dir <- here("results/metastasis_genes")
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

result %>%
  vroom::vroom_write(
    file = file.path(output_dir, "metastasis_genes.csv"),
    delim = ","
  )

table1 %>%
  vroom::vroom_write(
    file = file.path(output_dir, "table1.csv"),
    delim = ","
  )

head(result)

# --- Memory cleanup ---
rm(genesPerTerm)
gc()
```

## Equivalence Checks

Temporary validation chunk to verify that the refactored script produces
outputs identical to the original pipeline. Set `eval: true` to activate.

```{r equivalence}
#| eval: false
#| cache: false

# Helper: trim whitespace in all character columns for robust comparison
trim_char_cols <- function(df) {
  df[] <- lapply(df, function(col) {
    if (is.character(col)) trimws(col) else col
  })
  df
}

# --- Check metastasis_genes.csv equivalence ---
# Compare the newly written CSV against the original to confirm identical output
new_genes <- vroom::vroom(
  here("results/metastasis_genes/metastasis_genes.csv"),
  show_col_types = FALSE
)

new_genes_sorted <- trim_char_cols(as.data.frame(
  new_genes[order(new_genes$external_gene_name, new_genes$Signature), ]
))
orig_genes_sorted <- trim_char_cols(as.data.frame(
  result[order(result$external_gene_name, result$Signature), ]
))

genes_match <- all.equal(orig_genes_sorted, new_genes_sorted, check.attributes = FALSE)
message("metastasis_genes.csv equivalence: ", genes_match)

# --- Check table1.csv equivalence ---
# Compare new output against original paper_data/table1.csv
original_table1 <- vroom::vroom(
  here("results/paper_data/table1.csv"),
  show_col_types = FALSE
)
new_table1 <- vroom::vroom(
  here("results/metastasis_genes/table1.csv"),
  show_col_types = FALSE
)

orig_t1_sorted <- trim_char_cols(as.data.frame(
  original_table1[order(original_table1$GO_id), ]
))
new_t1_sorted <- trim_char_cols(as.data.frame(
  new_table1[order(new_table1$GO_id), ]
))

table1_match <- all.equal(orig_t1_sorted, new_t1_sorted, check.attributes = FALSE)
message("table1.csv equivalence: ", table1_match)

# Cleanup equivalence objects
rm(new_genes, new_genes_sorted, orig_genes_sorted,
   original_table1, new_table1, orig_t1_sorted, new_t1_sorted)
```

## Intersection of Gene Sets Across Biological Processes

An UpSet plot is generated to visualize the overlap structure among gene sets
assigned to each biological process signature. This provides an overview of
shared and unique genes across the metastatic hallmarks and suppressor categories.

```{r upset-plot}
#| fig-width: 12
#| fig-height: 7

# Pivot gene-signature table to a binary membership matrix for UpSetR
pivotada <- result %>%
  dplyr::select(external_gene_name, Signature) %>%
  dplyr::mutate(n = 1) %>%
  tidyr::pivot_wider(
    id_cols = external_gene_name,
    names_from = Signature,
    values_from = n,
    values_fn = list(n = length),
    values_fill = list(n = 0)
  )

# Display UpSet plot
UpSetR::upset(as.data.frame(pivotada), nsets = 50, nintersects = NA)

# --- Save UpSet plot as SVG and PNG ---
plot_dir <- here("results/metastasis_genes")
if (!dir.exists(plot_dir)) dir.create(plot_dir, recursive = TRUE)

# SVG
svglite::svglite(file.path(plot_dir, "upset_gene_sets.svg"), width = 12, height = 7)
UpSetR::upset(as.data.frame(pivotada), nsets = 50, nintersects = NA)
dev.off()

# PNG
png(file.path(plot_dir, "upset_gene_sets.png"), width = 12, height = 7, units = "in", res = 300)
UpSetR::upset(as.data.frame(pivotada), nsets = 50, nintersects = NA)
dev.off()
```
