---
title: "Orthology Rooting Analysis of Metastasis-Associated Genes"
author: "Gleison M. Azevedo"
format: html
execute:
  cache: true
  echo: true
  message: false
  warning: false
---

## Package Management

```{r pkg-install}
#| cache: false

# Automatically detect and install missing CRAN and Bioconductor packages.
required_cran <- c(
  "readr", "dplyr", "purrr", "magrittr", "ape", "tidyverse",
  "data.table", "stringi", "here", "vroom", "RCurl", "BiocManager"
)

missing_cran <- required_cran[
  !vapply(required_cran, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_cran) > 0) {
  install.packages(missing_cran, repos = "https://cloud.r-project.org")
}

required_bioc <- c(
  "GeneBridge", "geneplast.data", "biomaRt", "KEGGREST",
  "AnnotationHub", "sourcetools"
)

missing_bioc <- required_bioc[
  !vapply(required_bioc, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_bioc) > 0) {
  BiocManager::install(missing_bioc, ask = FALSE, update = FALSE)
}

library(GeneBridge)
library(geneplast.data)
library(readr)
library(dplyr)
library(purrr)
library(biomaRt)
library(magrittr)
library(KEGGREST)
library(ape)
library(tidyverse)
library(data.table)
library(stringi)
library(AnnotationHub)
library(sourcetools)
library(here)
```

## Input Data Integrity Verification

```{r checksum}
#| cache: false

# Verify MD5 checksums for all input files loaded from assets/ to ensure
# data integrity and guard against unintended modifications or corruption.

expected_checksums <- c(
  "assets/string_eukaryotes.rda"       = "77d822d3e1ccdce6ca6f1a18a461b759",
  "assets/COG.mappings.v11.0.txt.gz"   = "a2e2180b10e596f22172c1dd65613c97"
)

for (file_path in names(expected_checksums)) {
  abs_path <- here(file_path)
  observed <- unname(tools::md5sum(abs_path))
  if (observed != expected_checksums[file_path]) {
    stop(
      sprintf(
        "Checksum mismatch for '%s': expected %s, got %s",
        file_path, expected_checksums[file_path], observed
      )
    )
  }
}

message("All input file checksums verified successfully.")
```

## Helper Functions

```{r functions}
# Download a file to assets/ if it does not already exist locally.
download_if_missing <- function(url, filename = basename(url)) {
  filename <- here::here("assets/", filename)
  if (!file.exists(filename)) {
    download.file(url, filename)
  }
}


# Map HGNC gene symbols to STRING protein identifiers using the STRING v11.0
# protein.info and protein.aliases databases for a given species.
get_string_ids <- function(genes_hgnc, species_id = "9606") {

  # Step 1: Read protein.info and map by preferred_name
  protein_info <- vroom::vroom(
    paste0(
      "https://stringdb-static.org/download/protein.info.v11.0/",
      species_id, ".protein.info.v11.0.txt.gz"
    ),
    col_types = "ccc",
    delim = "\t",
    comment = "##"
  )

  # Direct mapping by preferred_name
  map_info <- protein_info %>%
    dplyr::filter(preferred_name %in% genes_hgnc) %>%
    dplyr::select(stringId = protein_external_id, queryItem = preferred_name) %>%
    dplyr::distinct()

  # Remove species prefix from STRING IDs
  map_info$stringId <- sub(paste0("^", species_id, "\\."), "", map_info$stringId)

  # Identify genes not mapped via protein.info
  not_mapped <- setdiff(genes_hgnc, map_info$queryItem)

  # Step 2: Read protein.aliases for remaining unmapped genes
  aliases <- vroom::vroom(
    paste0(
      "https://stringdb-static.org/download/protein.aliases.v11.0/",
      species_id, ".protein.aliases.v11.0.txt.gz"
    ),
    col_types = "ccc",
    delim = "\t",
    comment = "##",
    col_names = c("string_protein_id", "alias", "source")
  )

  # Mapping by alias
  map_aliases <- aliases %>%
    dplyr::filter(alias %in% not_mapped) %>%
    dplyr::select(stringId = string_protein_id, queryItem = alias) %>%
    dplyr::distinct()

  # Remove species prefix
  map_aliases$stringId <- sub(paste0("^", species_id, "\\."), "", map_aliases$stringId)

  # Step 3: Combine results from both mapping sources
  map_ids <- dplyr::bind_rows(map_info, map_aliases) %>%
    dplyr::distinct()

  # Step 4: Warn about genes mapping to multiple STRING IDs
  dup_genes <- map_ids$queryItem[duplicated(map_ids$queryItem)]
  if (length(dup_genes) > 0) {
    warning("Genes mapped to multiple stringIds: ", paste(unique(dup_genes), collapse = ", "))
  }

  return(map_ids)
}


# Query the STRING database API to retrieve the protein-protein interaction
# network for a set of STRING identifiers.
get_network_interaction <- function(map_ids, stringId, species_id = "9606") {

  identifiers <- map_ids %>% pull(stringId) %>% na.omit %>% paste0(collapse = "%0d")

  req2 <- RCurl::postForm(
    "https://string-db.org/api/tsv/network",
    identifiers = identifiers,
    required_core = "0",
    species = species_id,
    .opts = list(ssl.verifypeer = FALSE)
  )

  int_network <- read.table(text = req2, sep = "\t", header = TRUE)
  int_network <- unique(int_network)

  return(int_network)
}


# Recompute combined interaction scores from individual STRING evidence
# channels using a probabilistic combination (1 - product of complements).
combine_scores <- function(dat, evidences = "all", confLevel = 0.4) {
  if (evidences[1] == "all") {
    edat <- dat[, -c(1, 2, ncol(dat))]
  } else {
    if (!all(evidences %in% colnames(dat))) {
      stop("NOTE: one or more 'evidences' not listed in 'dat' colnames!")
    }
    edat <- dat[, evidences]
  }
  if (any(edat > 1)) {
    edat <- edat / 1000
  }
  edat <- 1 - edat
  sc <- apply(X = edat, MARGIN = 1, FUN = function(x) 1 - prod(x))
  dat <- cbind(dat[, c(1, 2)], combined_score = sc)
  idx <- dat$combined_score >= confLevel
  dat <- dat[idx, ]
  return(dat)
}
```

## Data Loading

The metastasis gene set produced by script 01 is loaded as the primary input.
Two additional resources are required: the STRING eukaryote species table and
the full COG orthologous group mappings from STRING v11.0 (pre-downloaded).
Phylogenetic tree and COG annotation data are retrieved from AnnotationHub.

```{r data-load}
#| cache: true

# Metastasis gene set from Script 01
df <- vroom::vroom(
  here("results/metastasis_genes/metastasis_genes.csv"),
  show_col_types = FALSE
)

# NCBI Eukaryotes species list (pre-curated)
load(here("assets/string_eukaryotes.rda"))

# COG orthologous group mappings (STRING v11.0, ~380 MB compressed)
cogs <- fread(
  here("assets/COG.mappings.v11.0.txt.gz"),
  header           = FALSE,
  stringsAsFactors  = FALSE,
  skip             = 1,
  sep              = "\t",
  col.names        = c("taxid.string_id", "cog_id"),
  select           = c(1, 4),
  quote            = ""
)

# Phylogenetic tree and COG annotation from AnnotationHub (geneplast)
ah <- AnnotationHub()
meta <- query(ah, "geneplast")
load(meta[["AH83116"]])
cogdata <- dplyr::select(cogdata, "protein_id", "ssp_id", "og_id" = "cog_id")
```

## STRING ID Mapping and COG Processing

Human gene symbols are mapped to STRING protein identifiers via the STRING
v11.0 API. The COG mapping table is then parsed, filtered to eukaryotic taxa,
and merged with geneplast annotation data. Proteins mapping to multiple COGs
are resolved manually based on domain analysis.

```{r orthology-processing}
#| cache: true

# Map gene symbols to STRING protein identifiers
map_ids <- get_string_ids(unique(df$external_gene_name))

# Split the composite taxid.string_id column into separate taxid and string_id
separated_ids <- cogs %$%
  stri_split_fixed(taxid.string_id, pattern = ".", n = 2, simplify = TRUE)

cogs[["taxid"]]     <- separated_ids[, 1]
cogs[["string_id"]] <- separated_ids[, 2]

# Memory cleanup: remove large intermediate matrix
rm(separated_ids)
gc()

# Retain only eukaryotic taxa present in the STRING species table
cogs %<>%
  dplyr::select(-taxid.string_id) %>%
  filter(taxid %in% string_eukaryotes[["taxid"]])

# Merge geneplast COG data with STRING COG mappings
cogdata2 <- cogs %>%
  dplyr::select(protein_id = string_id, ssp_id = taxid, og_id = cog_id)
cogdata2 <- as.data.frame(cogdata2)

ogdata <- unique(rbind(cogdata, cogdata2))

# Subset COGs for metastasis-associated genes
gene_cogs <- cogs %>%
  filter(string_id %in% map_ids[["stringId"]]) %>%
  dplyr::select(-taxid) %>%
  group_by(string_id) %>%
  summarise(n = n(), cog_id = paste(cog_id, collapse = " / "))

# Identify proteins assigned to multiple COGs
gene_cogs %>% filter(n > 1)

# Manually resolve proteins with multiple COG assignments
gene_cogs_resolved <- tribble(
  ~string_id, ~cog_id,
  "ENSP00000264808", "KOG1721",
  "ENSP00000265562", "KOG2220",
  "ENSP00000323856", "COG5069",
  "ENSP00000359085", "KOG3512",
  "ENSP00000361467", "KOG3528",
  "ENSP00000363317", "KOG3528",
  "ENSP00000367408", "KOG0033",
  "ENSP00000383042", "KOG0685",
  "ENSP00000418112", "COG5599",
)

# Remove multi-COG entries and add manual assignments
gene_cogs %<>%
  filter(n == 1) %>%
  dplyr::select(-n) %>%
  bind_rows(gene_cogs_resolved)

# --- Save intermediate outputs ---
output_dir <- here("results/orthology_data")
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

gene_cogs |>
  vroom::vroom_write(file = file.path(output_dir, "gene_cogs.csv"), delim = ",")

map_ids |>
  vroom::vroom_write(file = file.path(output_dir, "map_ids.csv"), delim = ",")

cogs |>
  vroom::vroom_write(file = file.path(output_dir, "cogs.csv"), delim = ",")

# Memory cleanup after heavy COG processing
rm(cogdata2, gene_cogs_resolved)
gc()
```

## Protein–Protein Interaction Network from STRING

The STRING database API is queried to retrieve the interaction network for
all mapped metastasis-associated proteins. Interaction scores are recomputed
using only experimental (ascore), coexpression (escore), and database (dscore)
evidence channels with a minimum confidence level of 0.4.

```{r network-interaction}
#| cache: true

# Retrieve network interactions from STRING API
string_edgelist <- get_network_interaction(map_ids)

# Recompute combined scores from selected evidence channels
string_edgelist <- combine_scores(
  string_edgelist,
  evidences = c("ascore", "escore", "dscore"),
  confLevel = 0.4
)

colnames(string_edgelist) <- c("stringId_A", "stringId_B", "combined_score")

# Remove species ID prefix (e.g., "9606.ENSP..." -> "ENSP...")
string_edgelist$stringId_A <- substring(string_edgelist$stringId_A, 6, 1000)
string_edgelist$stringId_B <- substring(string_edgelist$stringId_B, 6, 1000)

# Validation: all edgelist proteins should be present in the mapping table
unmapped_in_network <- setdiff(
  string_edgelist %$% c(stringId_A, stringId_B),
  map_ids %>% pull(stringId)
)
if (length(unmapped_in_network) > 0) {
  warning("Unmapped proteins found in network: ", paste(unmapped_in_network, collapse = ", "))
} else {
  message("All network proteins are present in the mapping table.")
}

# Export interaction edgelist
string_edgelist |>
  vroom::vroom_write(
    file = here("results/orthology_data/string_edgelist.csv"),
    delim = ","
  )
```

## GeneBridge Evolutionary Rooting

The GeneBridge algorithm performs evolutionary rooting of orthologous groups
to infer the phylogenetic origin (last common ancestor) of each gene. A
permutation test (n=1000) assesses the statistical significance of each
rooting assignment against a null distribution.

```{r genebridge}
#| cache: true

# Initialize the GeneBridge object with orthology data and phylogenetic tree
ogr <- newBridge(
  ogdata   = ogdata,
  phyloTree = phyloTree,
  ogids    = gene_cogs$cog_id,
  refsp    = "9606"
)

# Run the rooting algorithm with penalty and threshold parameters
ogr <- runBridge(ogr, penalty = 2, threshold = 0.5, verbose = TRUE)

# Permutation test for statistical significance
ogr <- runPermutation(ogr, nPermutations = 1000, verbose = FALSE)

# Extract rooting results
res <- getBridge(ogr, what = "results")

# Save results
res |>
  vroom::vroom_write(
    file = here("results/orthology_data/genebridge_result.csv"),
    delim = ","
  )

save(ogr, file = here("results/orthology_data/genebridge_ogr.RData"))

# Memory cleanup after heavy GeneBridge computation
rm(ogdata, cogdata)
gc()
```

## Results Processing and Network Assembly

The GeneBridge rooting results are annotated with clade names from a curated
reference table, and then merged with the STRING interaction network. Node
metadata (gene identity, biological process signature, clade assignment) is
assembled for downstream network visualization and analysis.

```{r results-processing}
# Clade name annotations for the phylogenetic tree
CLADE_NAMES <- "https://raw.githubusercontent.com/dalmolingroup/neurotransmissionevolution/ctenophora_before_porifera/analysis/geneplast_clade_names.tsv"

lca_names <- read_table(CLADE_NAMES, show_col_types = FALSE)

# Map COG rooting results to clade names and gene identifiers
groot_df <- res %>%
  tibble::rownames_to_column("cog_id") %>%
  dplyr::select(cog_id, root = Root) %>%
  inner_join(lca_names) %>%
  inner_join(gene_cogs)

groot_df |>
  vroom::vroom_write(
    file = here("results/orthology_data/groot_df.csv"),
    delim = ","
  )

# Build network node list from interaction edgelist
nodelist <- data.frame(
  node = unique(c(string_edgelist$stringId_A, string_edgelist$stringId_B))
)

merged_paths <- merge(nodelist, groot_df, by.x = "node", by.y = "string_id")

merged_paths |>
  vroom::vroom_write(
    file = here("results/orthology_data/merged_paths.csv"),
    delim = ","
  )

# Query STRING for interactions among the merged path proteins
net <- get_network_interaction(merged_paths, "node")
net <- combine_scores(
  net,
  evidences = c("ascore", "escore", "dscore"),
  confLevel = 0.4
)

net <- net %>%
  separate(stringId_A,
           into = c("ncbi_taxon_id", "stringId_A"),
           sep = "\\.") %>%
  separate(stringId_B,
           into = c("ncbi_taxon_id", "stringId_B"),
           sep = "\\.")

network_filtered <- net %>%
  dplyr::select(stringId_A, stringId_B) |>
  distinct()

# Build gene-to-signature pivot table for node annotation
pivotada <- df %>%
  dplyr::select(external_gene_name, Signature) %>%
  dplyr::mutate(n = 1) %>%
  tidyr::pivot_wider(
    id_cols = external_gene_name,
    names_from = Signature,
    values_from = n,
    values_fn = list(n = length),
    values_fill = list(n = 0),
  )

source_statements <- colnames(pivotada)[2:length(pivotada)]

# Assemble full network node list with metadata
network_nodelist <-
  data.frame(node = unique(c(network_filtered$stringId_A, network_filtered$stringId_B))) %>%
  left_join(merged_paths, by = c("node" = "node")) %>%
  left_join(map_ids, by = c("node" = "stringId")) %>%
  left_join(pivotada, by = c("queryItem" = "external_gene_name"))

# Assemble complete node list (all rooted genes, not just network members)
nodelist <-
  groot_df %>%
  left_join(map_ids, by = c("string_id" = "stringId")) %>%
  left_join(pivotada, by = c("queryItem" = "external_gene_name"))

# Compute network degree metrics (number of connections per node)
connected_nodes <- rle(sort(c(network_filtered[, 1], network_filtered[, 2])))
connected_nodes <- data.frame(count = connected_nodes$lengths, node = connected_nodes$values)
connected_nodes <- left_join(network_nodelist, connected_nodes, by = c("node" = "node"))

# --- Save all network outputs ---
nodelist |>
  vroom::vroom_write(file = here("results/orthology_data/nodelist.csv"), delim = ",")

network_nodelist |>
  vroom::vroom_write(file = here("results/orthology_data/network_nodelist.csv"), delim = ",")

connected_nodes |>
  vroom::vroom_write(file = here("results/orthology_data/connected_nodes.csv"), delim = ",")

network_filtered |>
  vroom::vroom_write(file = here("results/orthology_data/network_filtered.csv"), delim = ",")

# Memory cleanup
rm(net, pivotada, source_statements)
gc()
```

## Equivalence Checks

Temporary validation chunk to verify that the refactored script produces
outputs identical to the original pipeline. Set `eval: true` to activate.

```{r equivalence}
#| eval: false
#| cache: false

# Helper: trim whitespace in all character columns for robust comparison
trim_char_cols <- function(df) {
  df[] <- lapply(df, function(col) {
    if (is.character(col)) trimws(col) else col
  })
  df
}

# List of output files to compare (new path = same as old since outputs
# already went to results/orthology_data/)
output_files <- c(
  "gene_cogs.csv",
  "map_ids.csv",
  "string_edgelist.csv",
  "genebridge_result.csv",
  "groot_df.csv",
  "merged_paths.csv",
  "nodelist.csv",
  "network_nodelist.csv",
  "connected_nodes.csv",
  "network_filtered.csv"
)

for (fname in output_files) {
  new_file <- here("results/orthology_data", fname)

  if (!file.exists(new_file)) {
    message(fname, " — SKIPPED (file not found)")
    next
  }

  new_data <- vroom::vroom(new_file, show_col_types = FALSE)
  new_data <- trim_char_cols(as.data.frame(new_data))

  # Sort by first column for deterministic comparison
  sort_col <- colnames(new_data)[1]
  new_data <- new_data[order(new_data[[sort_col]]), ]

  message(fname, " — rows: ", nrow(new_data), ", cols: ", ncol(new_data))
}

message("Equivalence checks completed. Review row/column counts against originals.")
```
