---
title: "Plotting Roots"
author: "Dalmolin Systems Biology Group"
format: html
editor: source
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  results = FALSE
)
```

## Import libraries

```{r libs}
library(RedeR)
library(dplyr)
library(igraph)
library(ggplot2)
library(ggraph)
library(ggforce)
library(tidyr)
library(scales)
library(forcats)
library(ggtext)
library(UpSetR)
library(here)
```

# Distribution

## Load Data

```{r}
#load("../results/plots/graph")
metastasis_genes <- vroom::vroom("../results/metastasis_genes/metastasis_genes.csv", delim = ",")
nodelist <- vroom::vroom("../results/orthology_data/nodelist.csv")

#CLADE_NAMES <- "https://raw.githubusercontent.com/dalmolingroup/neurotransmissionevolution/ctenophora_before_porifera/analysis/geneplast_clade_names.tsv"
#lca_names <- read_table(CLADE_NAMES)
#lca_names %<>% rename("lca" = root)

node_annotation <- dplyr::select(nodelist, c("queryItem","cog_id",
                                             "root", "clade_name"))
  
```

# Defining Functions

```{r themes}
# Plotting colors and labels
annotation_colors <- c(
  "cell adhesion"                               = "#06141FFF"
  ,"extracellular matrix organization"          = "#742C14FF"
  ,"epithelial to mesenchymal transition"       = "#3D4F7DFF"
  ,"regulation of metallopeptidase activity"    = "#E48C2AFF"
  ,"cell junction organization"                  ="#72874EFF"
  ,"cellular extravasation"                      ="#046E8FFF"
  ,"metastasis suppressor"                       ="#9B2335FF"
)

annotation_labels <- c(
   "cell adhesion"                              = "Adesão celular"
  ,"extracellular matrix organization"          = "Organização da matriz \nextracelular" 
  ,"epithelial to mesenchymal transition"       = "Transição \nepitélio-mesênquimal"
  ,"regulation of metallopeptidase activity"    = "Regulação da atividade \nda metalopeptidase"
  ,"cell junction organization"                 = "Organização da \njunção celular"
  ,"cellular extravasation"                     = "Extravasão celular"
  ,"metastasis suppressor"                      = "Supressor de metástase"
)

# This vertical line indicates the first metazoan (Amphimedon queenslandica / Ctenophora)
choanoflagellata_line <- geom_vline(
  xintercept = "Sphaeroforma arctica"
  ,color      = "#FF0000"
  ,linetype   = "11"
  ,alpha      = 1
  ,linewidth  = 0.25
)

# Plotting
theme_main <- theme(
  panel.spacing      = unit(2.5, "pt")
  ,strip.background   = element_blank()
  ,panel.grid.major.x = element_blank()
  ,panel.grid.major.y = element_line(linewidth = 0.25, linetype = "dotted", color = "#E0E0E0")
  ,strip.text.x       = element_text(size = 9, angle = 90, hjust = 0, vjust = 0.5, color = "#757575")
  ,strip.text.y       = element_text(size = 10, angle = 0, hjust = 0, vjust = 0.5, color = "#757575")
  ,axis.title         = element_text(size = 15, color = "#424242")
  ,axis.ticks.x       = element_blank()
  ,axis.text.x        = element_blank()
  ,axis.text.y        = element_text(size = 5.5)
  ,legend.position    = "none"
)

theme_supplementary <- theme(
  panel.grid.major.x = element_line(color = "#E0E0E0", linewidth = 0.25, linetype = "dotted")
  ,panel.grid.major.y = element_blank()
  ,strip.text.y       = element_text(size = 7, angle = 0, hjust = 0, vjust = 0.5, color = "#757575")
  ,strip.text.x       = element_text(size = 7, angle = 90, hjust = 0, vjust = 0.5, color = "#757575")
  ,axis.title         = element_text(size = 12, color = "#424242")
  ,axis.ticks         = element_line(colour = "grey20")
  ,axis.text.y        = element_text(size = 6, angle = 0, hjust = 1, vjust = 0.5, color = "#757575")
  ,axis.text.x        = element_text(size = 6)
)

theme_average <- theme(
  panel.spacing      = unit(1, "pt")
  ,axis.title         = element_text(color = "#424242")
  ,axis.text          = element_text(color = "#757575")
  ,axis.text.x        = element_text(size = 7, angle = -45, vjust = 0, hjust = 0)
  ,axis.text.y        = element_text(size = 5)
  ,strip.background   = element_blank()
  ,strip.text         = element_text(color = "#757575")
  ,strip.text.y       = element_text(angle = 0, hjust = 0, vjust = 0.5)
)

theme_big <- theme(
  panel.spacing      = unit(0.5, "pt")
  ,panel.grid.major.x = element_line(linewidth = 0.1, linetype = "dashed")
  ,panel.grid.major.y = element_blank()
  ,strip.background   = element_blank()
  ,strip.text.x       = element_text(size = 8, angle = 90, hjust = 0.5, vjust = 0)
  ,strip.text.y       = element_text(size = 8, angle = 0, hjust = 0, vjust = 0.5)
  ,axis.text.x        = element_text(size = 6, angle = 90, vjust = 0, hjust = 0)
  ,axis.text.y        = element_text(size = 4.5)
  ,axis.ticks         = element_line(size = 0.1)
)
```

```{r}
# # Custom tick function
# tick_function <- function(x) {
#   seq(x[2], 0, length.out = 3) %>% head(-1) %>% tail(-1) %>% { ceiling(./5)*5 }
# }
# 
# calculate_cumulative_genes <- function(nodelist) {
#   
#     # Obter todas as categorias possíveis de clade_name
#   all_clades <- node_annotation %>%
#     arrange(desc(root)) %>%
#     dplyr:: select(clade_name) %>%
#     unique()
#   
#   # Definir as colunas de interesse
#   process_columns <- c("queryItem", "root", "clade_name", 
#                        "cellular extravasation", 
#                        "cell junction organization",           
#                        "cell adhesion",                          
#                        "extracellular matrix organization",     
#                        "epithelial to mesenchymal transition",  
#                        "regulation of metallopeptidase activity")
#   
#   # Calcular o cumulativo agrupando por clade_name
#   cumulative_genes <- nodelist %>%
#     arrange(desc(root)) %>%
#     dplyr::select(all_of(process_columns)) %>%
#     group_by(clade_name, root) %>%
#     summarise(count_genes = n(), .groups = "drop") %>%
#     arrange(desc(root)) %>%
#     mutate(cumulative_sum = cumsum(count_genes)) %>%
#     right_join(all_clades, by = "clade_name") %>%
#     fill(cumulative_sum, .direction = "down")
#   
#   return(cumulative_genes)
# }
# 
# calculate_cumulative_bp <- function(nodelist) {
#     
#   # Obter todas as categorias possíveis de clade_name
#   all_clades <- node_annotation %>%
#     arrange(desc(root)) %>%
#     dplyr:: select(clade_name) %>%
#     unique()
#   
#   # Definir as colunas de interesse
#   process_columns <- c("queryItem", "root", "clade_name", 
#                        "cellular extravasation", 
#                        "cell junction organization",           
#                        "cell adhesion",                          
#                        "extracellular matrix organization",     
#                        "epithelial to mesenchymal transition",  
#                        "regulation of metallopeptidase activity")
#   
#   # Calcular a soma cumulativa para cada processo biológico
#   cumulative_bp <- nodelist %>%
#     dplyr::select(all_of(process_columns)) %>%
#     distinct(root, queryItem, .keep_all = TRUE) %>%
#     mutate(across(all_of(process_columns[-c(1:3)]), ~ as.numeric(.))) %>%
#     group_by(root, clade_name) %>%
#     summarise(across(all_of(process_columns[-c(1:3)]), 
#                      ~ sum(. , na.rm = TRUE)),
#               .groups = "drop") %>%
#     arrange(desc(root)) %>%
#     mutate(across(all_of(process_columns[-c(1:3)]), ~ cumsum(.))) %>%
#     right_join(all_clades, by = "clade_name") %>%
#     fill(everything(), .direction = "down")
#   
#   return(cumulative_bp)
# }
# 
# BP_by_clade <- function(data) {
#   data %>%
#     pivot_longer(
#       cols = c(`cellular extravasation`, `cell junction organization`, 
#                `cell adhesion`, `extracellular matrix organization`, 
#                `epithelial to mesenchymal transition`, 
#                `regulation of metallopeptidase activity`),
#       names_to = "process",
#       values_to = "presence"
#     ) %>%
#     filter(presence == 1) %>%
#     group_by(clade_name, process, root) %>% 
#     summarise(count = n(), .groups = "drop") %>%
#     mutate(total_genes = sum(count), .by = process) %>%
#     mutate(proportion = count / total_genes)
# }
# 
# plot_stacked_bar_chart <- function(data) {
#    # Definir os grupos de interesse
#    clados_interesse <- c("Metamonada", "Choanoflagellata", "Ctenophora", "Actinopterygii")
#    
#    # Criar as categorias "Outros" com base nos intervalos de root
#    filtered_data <- data %>%
#      mutate(clade_group = case_when(
#        clade_name == "Metamonada" ~ "Metamonada",
#        root >= 31 & root <= 36 ~ "Intermediate Metam-Choano",
#        clade_name == "Choanoflagellata" ~ "Choanoflagellata",
#        clade_name == "Ctenophora" ~ "Ctenophora",
#        root >= 21 & root <= 28 ~ "Intermediate Cteno-Actino",
#        clade_name == "Actinopterygii" ~ "Actinopterygii",
#        TRUE ~ NA_character_
#      )) %>%
#      filter(!is.na(clade_group)) %>%
#      mutate(clade_group = factor(clade_group, 
#                               levels = c("Metamonada", "Intermediate Metam-Choano",
#                                          "Choanoflagellata", "Ctenophora",
#                                          "Intermediate Cteno-Actino", "Actinopterygii")))
#    
#    # Ajustar a largura das barras: mais finas para "Outros"
#    filtered_data <- filtered_data %>%
#      mutate(bar_width = ifelse(grepl("Intermediate", clade_group), 0.4, 0.8)) %>%
#      mutate(alpha = ifelse(grepl("Intermediate", clade_group), 0.55, 1))
#    
#    # Criar o gráfico de barras empilhadas com larguras ajustadas
#    ggplot(filtered_data, aes(x = clade_group, y = proportion, fill = process, 
#                              width = bar_width)) +
#      geom_bar(position = "fill", stat = "identity", color = NA, alpha = filtered_data$alpha) +
#      theme_minimal() +
#      labs(
#        title = "Biological Process Distribution",
#        x="",
#        y = "Genes Proportion",
#        fill = "Biological Processes"
#      ) +
#      scale_fill_manual(values = c(
#        "cell adhesion"                               = "#06141FFF"
#        ,"extracellular matrix organization"          = "#742C14FF"
#        ,"epithelial to mesenchymal transition"       = "#3D4F7DFF"
#        ,"regulation of metallopeptidase activity"    = "#E48C2AFF"
#        ,"cell junction organization"                  ="#72874EFF"
#        ,"cellular extravasation"                      ="#046E8FFF"
#      )) + 
#      scale_y_continuous(labels = scales::percent) +
#      theme(
#        axis.text = element_text(size = 12),
#        axis.title = element_text(size = 14),
#        axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
#        legend.position = "right",
#        plot.title = element_text(size = 16, hjust = 0.5, face = "bold"),
#        panel.grid = element_blank()
#      ) +
#      guides(alpha = "none")
#  }
```

```{r}
# Custom tick function
tick_function <- function(x) {
  seq(x[2], 0, length.out = 3) %>% head(-1) %>% tail(-1) %>% { ceiling(./5)*5 }
}

calculate_cumulative_genes <- function(nodelist) {
  
    # Obter todas as categorias possíveis de clade_name
  all_clades <- node_annotation %>%
    arrange(desc(root)) %>%
    dplyr:: select(clade_name) %>%
    unique()
  
  # Definir as colunas de interesse
  process_columns <- c("queryItem", "root", "clade_name", 
                       "cellular extravasation", 
                       "cell junction organization",           
                       "cell adhesion",                          
                       "extracellular matrix organization",     
                       "epithelial to mesenchymal transition",  
                       "regulation of metallopeptidase activity",
                       "metastasis suppressor")
  
  # Calcular o cumulativo agrupando por clade_name
  cumulative_genes <- nodelist %>%
    arrange(desc(root)) %>%
    dplyr::select(all_of(process_columns)) %>%
    group_by(clade_name, root) %>%
    summarise(count_genes = n(), .groups = "drop") %>%
    arrange(desc(root)) %>%
    mutate(cumulative_sum = cumsum(count_genes)) %>%
    right_join(all_clades, by = "clade_name") %>%
    fill(cumulative_sum, .direction = "down")
  
  return(cumulative_genes)
}

calculate_cumulative_bp <- function(nodelist) {
    
  # Obter todas as categorias possíveis de clade_name
  all_clades <- node_annotation %>%
    arrange(desc(root)) %>%
    dplyr:: select(clade_name) %>%
    unique()
  
  # Definir as colunas de interesse
  process_columns <- c("queryItem", "root", "clade_name", 
                       "cellular extravasation", 
                       "cell junction organization",           
                       "cell adhesion",                          
                       "extracellular matrix organization",     
                       "epithelial to mesenchymal transition",  
                       "regulation of metallopeptidase activity",
                       "metastasis suppressor")
  
  # Calcular a soma cumulativa para cada processo biológico
  cumulative_bp <- nodelist %>%
    dplyr::select(all_of(process_columns)) %>%
    distinct(root, queryItem, .keep_all = TRUE) %>%
    mutate(across(all_of(process_columns[-c(1:3)]), ~ as.numeric(.))) %>%
    group_by(root, clade_name) %>%
    summarise(across(all_of(process_columns[-c(1:3)]), 
                     ~ sum(. , na.rm = TRUE)),
              .groups = "drop") %>%
    arrange(desc(root)) %>%
    mutate(across(all_of(process_columns[-c(1:3)]), ~ cumsum(.))) %>%
    right_join(all_clades, by = "clade_name") %>%
    fill(everything(), .direction = "down")
  
  return(cumulative_bp)
}

BP_by_clade <- function(data) {
  data %>%
    pivot_longer(
      cols = c(`cellular extravasation`, `cell junction organization`, 
               `cell adhesion`, `extracellular matrix organization`, 
               `epithelial to mesenchymal transition`, 
               `regulation of metallopeptidase activity`,
               `metastasis suppressor`),
      names_to = "process",
      values_to = "presence"
    ) %>%
    filter(presence == 1) %>%
    group_by(clade_name, process, root) %>% 
    summarise(count = n(), .groups = "drop") %>%
    mutate(total_genes = sum(count), .by = process) %>%
    mutate(proportion = count / total_genes)
}

plot_stacked_bar_chart <- function(data) {
   # Definir os grupos de interesse
   clados_interesse <- c("Choanoflagellata", "Ctenophora", "Actinopterygii")
   
   # Criar as categorias "Outros" com base nos intervalos de root
   filtered_data <- data %>%
     mutate(clade_group = case_when(
       root >= 31 & root <= 37 ~ "Intermediate Metam-Choano",
       clade_name == "Choanoflagellata" ~ "Choanoflagellata",
       clade_name == "Ctenophora" ~ "Ctenophora",
       root >= 21 & root <= 28 ~ "Intermediate Cteno-Actino",
       clade_name == "Actinopterygii" ~ "Actinopterygii",
       TRUE ~ NA_character_
     )) %>%
     filter(!is.na(clade_group)) %>%
     mutate(clade_group = factor(clade_group, 
                              levels = c("Intermediate Metam-Choano",
                                         "Choanoflagellata", "Ctenophora",
                                         "Intermediate Cteno-Actino", "Actinopterygii")))
   
   # Ajustar a largura das barras: mais finas para "Outros"
   filtered_data <- filtered_data %>%
     mutate(bar_width = ifelse(grepl("Intermediate", clade_group), 0.4, 0.8)) %>%
     mutate(alpha = ifelse(grepl("Intermediate", clade_group), 0.55, 1))
   
   # Criar o gráfico de barras empilhadas com larguras ajustadas
   ggplot(filtered_data, aes(x = clade_group, y = proportion, fill = process, 
                             width = bar_width)) +
     geom_bar(position = "fill", stat = "identity", color = NA, alpha = filtered_data$alpha) +
     theme_minimal() +
     labs(
       title = "Biological Process Distribution",
       x="",
       y = "Genes Proportion",
       fill = "Biological Processes"
     ) +
     scale_fill_manual(values = c(
       "cell adhesion"                               = "#06141FFF"
       ,"extracellular matrix organization"          = "#742C14FF"
       ,"epithelial to mesenchymal transition"       = "#3D4F7DFF"
       ,"regulation of metallopeptidase activity"    = "#E48C2AFF"
       ,"cell junction organization"                  ="#72874EFF"
       ,"cellular extravasation"                      ="#046E8FFF"
       ,"metastasis suppressor"                       ="#9B2335FF"
     )) + 
     scale_y_continuous(labels = scales::percent) +
     theme(
       axis.text = element_text(size = 12),
       axis.title = element_text(size = 14),
       axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
       legend.position = "right",
       plot.title = element_text(size = 16, hjust = 0.5, face = "bold"),
       panel.grid = element_blank()
     ) +
     guides(alpha = "none")
 }
```

## Gene distribution

```{r}
  upset(dplyr::select(as.data.frame(nodelist), 
                    "queryItem", 
                    "cell adhesion",                          
                    "extracellular matrix organization",      
                    "epithelial to mesenchymal transition",   
                    "regulation of metallopeptidase activity",
                    "cell junction organization",
                    "cellular extravasation",
                    "metastasis suppressor"),
                    nsets = 50, nintersects = NA,
      sets.bar.color = c("#06141FFF", "#72874EFF", "#3D4F7DFF",
                         "#742C14FF","#046E8FFF", "#E48C2AFF", "#9B2335FF"), 
      mainbar.y.label = "Biological Process \nIntersections",
      sets.x.label = "Set Size")
```

## Cumulative rooting count

```{r}
cumulative_genes <- calculate_cumulative_genes(nodelist) 
cumulative_bp <- calculate_cumulative_bp(nodelist)
  
cumulative_data <- left_join(cumulative_genes, cumulative_bp)
 

long_data <- cumulative_data %>%
  pivot_longer(cols = 5:10, 
               names_to = "Process", 
               values_to = "Value")

#a <-
ggplot() +
  # Gráfico de barras para cumulative_sum
  geom_bar(data = cumulative_data, 
           aes(x = factor(clade_name, levels = clade_name), y = cumulative_sum), 
           stat = "identity", fill = "darkgray", colour = NA) +
  geom_text(data = cumulative_data, 
            aes(x = factor(clade_name, levels = clade_name), y = cumulative_sum, label = cumulative_sum), 
            vjust = -0.5, size = 3, color = "darkgray") +
  scale_color_manual(values = annotation_colors) +
  
  labs(x = "Clade Name", y = "Cumulative Sum", 
       title = "Cumulative Sum and Biological Processes",
       fill = "Cumulative Sum",
       color = "Biological Processes") +
  
  theme_main +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

#b <- 
ggplot() +
  # Gráfico de barras para cumulative_sum
  geom_bar(data = cumulative_data, 
           aes(x = factor(-root), y = cumulative_sum), 
           stat = "identity", fill = "darkgray", colour = NA) +
  geom_text(data = cumulative_data, 
            aes(x = factor(-root), y = cumulative_sum, label = cumulative_sum), 
            vjust = -0.5, size = 3, color = "darkgray") +
  
  # Gráfico de linhas para os processos biológicos
  geom_line(data = long_data, 
            aes(x = factor(-root), y = Value, color = Process, group = Process), 
            size = 1) +
  
  # Usar a paleta de cores definida
  scale_color_manual(values = annotation_colors) +
  
  labs(x = "Clade Name", y = "Cumulative Sum", 
       title = "Cumulative Sum and Biological Processes",
       fill = "Cumulative Sum",
       color = "Biological Processes") +
  
  theme_main +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#ggsave(file = "../results/plots/cumulative.svg", plot=a, width=15, height=8)
#ggsave(file = "../results/plots/cumulative_bp.svg", plot=b, width=15, height=8)
```

## Modified z-score

```{r}
# 1️ Calcular o número de genes únicos por clado
calc_gaps <- nodelist %>%
  dplyr::select(root, clade_name, queryItem) %>%
  unique() %>%
  arrange(desc(root))  # do mais recente para o mais basal

calc_gaps2 <- calc_gaps %>%
  dplyr::group_by(root, clade_name) %>%
  summarise(n = length(unique(queryItem)), .groups = "drop") %>%
  arrange(desc(root))

# 2️ Separar Metamonada (baseline) dos clados seguintes
baseline <- calc_gaps2[1, ]        # Metamonada
clados <- calc_gaps2[-1, ]         # demais clados para estatística

# 3️ Calcular Modified Z-score apenas nos clados posteriores
mediana <- median(clados$n)
mad <- median(abs(clados$n - mediana))

clados <- clados %>%
  mutate(
    modified_z_scores = 0.6745 * (n - mediana) / mad,
    p_value = 2 * pnorm(-abs(modified_z_scores)),
    p_adjusted = p.adjust(p_value, method = "fdr")
  )

# 4️ Reconstruir dataframe final incluindo Metamonada
calc_gaps2$modified_z_scores <- NA
calc_gaps2$p_value <- NA
calc_gaps2$p_adjusted <- NA

calc_gaps2$modified_z_scores[-1] <- clados$modified_z_scores
calc_gaps2$p_value[-1] <- clados$p_value
calc_gaps2$p_adjusted[-1] <- clados$p_adjusted

# 5 Calcular acumulado real de genes (não subtrair baseline)
calc_gaps2$acumulado <- cumsum(calc_gaps2$n)

# 6 Filtrar clados significativos
signif_clades <- calc_gaps2 %>%
  filter(!is.na(p_adjusted),
         p_adjusted <= 0.05,
         abs(modified_z_scores) > 3.5)

suppl_table1 <- calc_gaps2 %>%
  filter(!is.na(p_adjusted))

suppl_table1 |> 
  vroom::vroom_write(file = here("results/paper_data/suppl_table1.csv"), delim = ",")

# Resultado final
calc_gaps2
signif_clades
```

```{r}
# calc_gaps <- nodelist %>%
#   dplyr::select(root, clade_name, queryItem) %>%
#   unique() %>%
#   arrange(root)
# 
# calc_gaps2 <- calc_gaps %>%
#   dplyr::group_by(root, clade_name) %>%
#   summarise(n = length(unique(queryItem))) %>%
#   ungroup() %>%
#   arrange(desc(root))
# 
# calc_gaps2$acumulado <- cumsum(calc_gaps2$n)
# 
# 
# gaps <- calc_gaps2$n
# 
# ############## Calculando o Modified Z-score #################
# mediana <- median(gaps)
# mad <- median(abs(gaps - mediana))
# 
# # Fator de escala para aproximar MAD ao desvio padrão (0.6745), valor utilizado como padrão
# # valores de MAD z-score >= 3.5 são considerados significativos
# 
# calc_gaps2$modified_z_scores <- 0.6745 * (gaps - mediana) / mad
# 
# # transformando de MAD z-score para p-value e p-value ajustado
# calc_gaps2$p_value <- 2 * pnorm(-abs(calc_gaps2$modified_z_scores))
# 
# calc_gaps2$p_ajustado <- p.adjust(calc_gaps2$p_value, method = "BH")
# 
# signif_clades <- calc_gaps2[calc_gaps2$p_ajustado <= 0.05 &
#                               abs(calc_gaps2$modified_z_scores) > 3.5,]
# 
# signif_clades
```


```{r}
nodelist_filtered <- nodelist %>%
  mutate(node_color = case_when(
   # clade_name == "Metamonada" ~ "#4CA375",
    clade_name == "Choanoflagellata" ~ "#D17878",
    clade_name == "Ctenophora" ~ "#8A6BBE",
    clade_name == "Actinopterygii" ~ "#4A85C2",
    TRUE ~ "#999999FF"
  )) %>%
  mutate(node_size = case_when(
    clade_name %in% c(#"Metamonada", 
      "Choanoflagellata", "Ctenophora", "Actinopterygii") ~ 50,
    TRUE ~ 35
  )) %>% 
  mutate(node_linecolor = case_when(
    queryItem %in% metastasis_genes$external_gene_name ~ "#6A0DAD", 
    TRUE ~ "#999999FF"
  )) %>%
    mutate(node_linewidth = case_when(
    queryItem %in% metastasis_genes$external_gene_name ~ 3, 
    TRUE ~ 1
    )) %>%
  filter(root >= 20)
```

```{r}
df_proportion <- BP_by_clade(nodelist_filtered)

plot_stacked_bar_chart(df_proportion)
```

## Hipergeometric Test

```{r}
nodelist <- vroom::vroom("../results/orthology_data/nodelist.csv")

# Filter out genes that are ONLY metastasis suppressors (exclude from statistical test)
# Keep genes that have at least one of the 6 original biological processes

   clades <- unique(nodelist$clade_name)
   processes <- c("cell adhesion", "extracellular matrix organization", 
                  "epithelial to mesenchymal transition", 
                  "regulation of metallopeptidase activity", 
                  "cell junction organization", "cellular extravasation",
                  "metastasis suppressor")
   
  
   result <- list()
   
   for (process in processes) {
     for (clade in clades) {
       
       N <- length(unique(nodelist$string_id))                     # total de genes únicos
       M <- sum(nodelist[ ,process])                               # número de genes que participam do processo
       n <- length(which(nodelist$clade_name == clade))            # número de genes no clado
       k <- sum(filter(nodelist, clade_name == clade)[[process]])  # número de genes do clado que participam do processo
       
       p_valor <- 1 - phyper(k - 1, M, N - M, n)
       
       result[[paste(process, clade, sep = "_")]] <- c(process, clade, k, M, N, n, p_valor)
     }
   }
   
   hipergeometric_df <- do.call(rbind, result) %>%
     as.data.frame() %>%
     setNames(c("Process", "clade_name", "k", "M", "N", "n", "p.val")) %>% 
     arrange(p.val)
     rownames(hipergeometric_df) <- NULL

   hipergeometric_df <- hipergeometric_df %>%
     mutate(across(all_of(c("k", "M", "N", "n", "p.val")), as.numeric))   
   
   hipergeometric_df <- hipergeometric_df %>%
    mutate(p.adj = p.adjust(p.val, method = "BH"))
        
   head(hipergeometric_df)
```

```{r}
hipergeometric_df$p.adj <- as.numeric(hipergeometric_df$p.adj)
hipergeometric_df$p.val_color <- ifelse(hipergeometric_df$p.adj > 0.05, "gray", "red")

i <- nodelist %>% 
  dplyr::select(clade_name, root) %>% 
  unique()

hipergeometric_df <- hipergeometric_df %>%
  inner_join(i, by = "clade_name")

hipergeometric_df$clade_name <- factor(hipergeometric_df$clade_name, 
    levels = unique(hipergeometric_df$clade_name[order(hipergeometric_df$root)]))


hipergeometric_df$Process <- factor(hipergeometric_df$Process, 
                                    levels = c("extracellular matrix organization", 
                                               "epithelial to mesenchymal transition", 
                                               "cell junction organization",
                                               "regulation of metallopeptidase activity",
                                               "cell adhesion", 
                                               "cellular extravasation",
                                               "metastasis suppressor"))

ggplot(hipergeometric_df, aes(x = Process, 
                              y = clade_name, 
                              size = as.numeric(n), 
                              color = p.val_color)) +
  geom_point() +
  scale_color_manual(values = c("gray" = "gray", "red" = "red"), 
                     labels = c("Not Significant", "Significant")) +  
  labs(x = "Biological Process", y = "Clade", size = "Number of Orthologs", color = "Significance") +
  scale_size_continuous(breaks = c(10, 50, 100), labels = c("10 Orthologs", "50 Orthologs", "100 Orthologs")) + 
  theme_minimal() +
  ggtitle("Hipergeometric Test") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_blank()) +
  scale_y_discrete(labels = function(x) {
   # x <- gsub("Metamonada", "<b>Metamonada</b>", x)
    x <- gsub("Choanoflagellata", "<b>Choanoflagellata</b>", x)
    x <- gsub("Ctenophora", "<b>Ctenophora</b>", x)
    x <- gsub("Actinopterygii", "<b>Actinopterygii</b>", x)
    return(x)
  }) +
  theme(axis.text.y = ggtext::element_markdown())

# Export hypergeometric results for use in script 04
hipergeometric_df %>%
  vroom::vroom_write(file = here("results/paper_data/hipergeometric_all_genes.csv"), delim = ",")
```

## Comparative Analysis: All Genes vs Metastasis Suppressors

```{r}
# Filter nodelist to get only metastasis suppressor genes
nodelist_suppressors <- nodelist %>%
  filter(`metastasis suppressor` == 1)

# Filter nodelist to exclude metastasis suppressor genes (for fair comparison)
nodelist_no_suppressors <- nodelist %>%
  filter(`metastasis suppressor` == 0 | is.na(`metastasis suppressor`))

# Calculate cumulative percentage for all genes (EXCLUDING metastasis suppressors)
total_all <- length(unique(nodelist_no_suppressors$queryItem))
calc_cumulative_all <- nodelist_no_suppressors %>%
  dplyr::select(root, clade_name, queryItem) %>%
  unique() %>%
  arrange(desc(root)) %>%
  group_by(root, clade_name) %>%
  summarise(n = n(), .groups = "drop") %>%
  arrange(desc(root)) %>%
  mutate(cumulative = cumsum(n),
         percentage = (cumulative / total_all) * 100,
         dataset = "Metastasis genes")

# Calculate cumulative percentage for suppressors only
total_supp <- length(unique(nodelist_suppressors$queryItem))
calc_cumulative_supp <- nodelist_suppressors %>%
  dplyr::select(root, clade_name, queryItem) %>%
  unique() %>%
  arrange(desc(root)) %>%
  group_by(root, clade_name) %>%
  summarise(n = n(), .groups = "drop") %>%
  arrange(desc(root)) %>%
  mutate(cumulative = cumsum(n),
         percentage = (cumulative / total_supp) * 100,
         dataset = "Metastasis suppressors")

# Combine datasets
cumulative_comparison <- bind_rows(calc_cumulative_all, calc_cumulative_supp)

# Order clade_name by root
clade_order <- cumulative_comparison %>%
  arrange(desc(root)) %>%
  pull(clade_name) %>%
  unique()

cumulative_comparison$clade_name <- factor(cumulative_comparison$clade_name, levels = clade_order)

# Plot cumulative comparison in percentage with clade names on x-axis
ggplot(cumulative_comparison, aes(x = clade_name, y = percentage, 
                                   color = dataset, group = dataset)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_color_manual(values = c("Metastasis genes" = "#2E86AB", 
                                 "Metastasis suppressors" = "#9B2335")) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(x = "Evolutionary Root (Clade)", 
       y = "Cumulative Percentage of Genes",
       title = "Cumulative Rooting: Metastasis Genes vs Metastasis Suppressors",
       subtitle = paste0("Metastasis genes: n=", total_all, " | Metastasis suppressors: n=", total_supp),
       color = "Gene Set") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
    axis.title = element_text(size = 12),
    legend.position = "bottom",
    panel.grid.major.y = element_line(linewidth = 0.25, linetype = "dotted", color = "#E0E0E0"),
    panel.grid.major.x = element_blank()
  )

ggsave(file = here("results/plots/cumulative_comparison_percentage.svg"), width = 12, height = 6)
```

```{r}
# Cumulative bar graph for metastasis suppressors only
calc_cumulative_supp$clade_name <- factor(calc_cumulative_supp$clade_name, levels = clade_order)

ggplot(calc_cumulative_supp, aes(x = clade_name, y = cumulative)) +
  geom_bar(stat = "identity", fill = "#9B2335", alpha = 0.8) +
  geom_text(aes(label = cumulative), vjust = -0.3, size = 3) +
  labs(x = "Evolutionary Root (Clade)", 
       y = "Cumulative Gene Count",
       title = "Cumulative Rooting: Metastasis Suppressors",
       subtitle = paste0("Total genes: n=", total_supp)) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.title = element_text(size = 12),
    panel.grid.major.y = element_line(linewidth = 0.25, linetype = "dotted", color = "#E0E0E0"),
    panel.grid.major.x = element_blank()
  )

ggsave(file = here("results/plots/cumulative_suppressors_bar.svg"), width = 12, height = 6)
```

## Comparative Analysis: Metastasis Genes vs Suppressors vs KEGG Pathways in Cancer

### Query KEGG Pathway hsa05200

```{r}
# Query KEGG pathway hsa05200 (Pathways in cancer)
kegg_url <- "https://rest.kegg.jp/link/hsa/hsa05200"
kegg_genes_raw <- tryCatch({
  read.table(url(kegg_url), sep = "\t", header = FALSE, stringsAsFactors = FALSE)
}, error = function(e) {
  message("Error fetching KEGG data: ", e$message)
  return(NULL)
})

if (!is.null(kegg_genes_raw)) {
  # Extract gene IDs (format: hsa:XXXX -> XXXX)
  kegg_gene_ids <- gsub("hsa:", "", kegg_genes_raw$V2)
  
  # Get gene symbols directly from pathway
  kegg_names_url <- "https://rest.kegg.jp/get/hsa05200"
  pathway_info <- tryCatch({
    readLines(url(kegg_names_url))
  }, error = function(e) {
    message("Error fetching pathway info: ", e$message)
    return(NULL)
  })
  
  # Parse gene symbols from the pathway entry
  kegg_symbols <- c()
  in_gene_section <- FALSE
  for (line in pathway_info) {
    if (grepl("^GENE", line)) {
      in_gene_section <- TRUE
      match <- regmatches(line, regexpr("[0-9]+\\s+([A-Za-z0-9-]+)", line))
      if (length(match) > 0) {
        symbol <- gsub("^[0-9]+\\s+", "", match)
        kegg_symbols <- c(kegg_symbols, symbol)
      }
    } else if (in_gene_section && grepl("^\\s+[0-9]+", line)) {
      match <- regmatches(line, regexpr("[0-9]+\\s+([A-Za-z0-9-]+)", line))
      if (length(match) > 0) {
        symbol <- gsub("^[0-9]+\\s+", "", match)
        kegg_symbols <- c(kegg_symbols, symbol)
      }
    } else if (in_gene_section && !grepl("^\\s", line)) {
      break
    }
  }
  
  kegg_symbols <- unique(kegg_symbols)
  cat("KEGG hsa05200 genes found:", length(kegg_symbols), "\n")
}
```

### Define Gene Sets for Comparison

```{r}
# Define the 6 biological processes
biological_processes <- c("cell adhesion", 
                          "extracellular matrix organization", 
                          "epithelial to mesenchymal transition", 
                          "regulation of metallopeptidase activity", 
                          "cell junction organization", 
                          "cellular extravasation")

# Metastasis genes: genes that have at least ONE of the 6 biological processes
# (some of these may ALSO be suppressors - that's the overlap we want to capture)
nodelist_bp <- nodelist %>%
  mutate(has_bp = rowSums(dplyr::select(., all_of(biological_processes)), na.rm = TRUE) > 0) %>%
  filter(has_bp)
metastasis_genes <- unique(nodelist_bp$queryItem)

# Metastasis suppressors: genes flagged as suppressors
suppressor_genes <- unique(nodelist_suppressors$queryItem)

# KEGG genes that are in our analysis
kegg_in_analysis <- intersect(kegg_symbols, unique(nodelist$queryItem))

# Print set sizes
cat("Metastasis genes (with at least 1 BP):", length(metastasis_genes), "\n")
cat("Metastasis suppressors:", length(suppressor_genes), "\n")
cat("KEGG hsa05200 genes in analysis:", length(kegg_in_analysis), "\n")

# Check overlaps
overlap_meta_supp_check <- intersect(metastasis_genes, suppressor_genes)
cat("Overlap (Metastasis genes ∩ Suppressors):", length(overlap_meta_supp_check), "\n")
cat("Genes appearing in both:", paste(overlap_meta_supp_check, collapse = ", "), "\n")
```

### UpSet Plot: Gene Set Overlaps

```{r}
# Create binary matrix for UpSet plot
all_genes <- unique(c(metastasis_genes, suppressor_genes, kegg_in_analysis))

upset_df <- data.frame(
  gene = all_genes,
  `Metastasis genes` = as.integer(all_genes %in% metastasis_genes),
  `Metastasis suppressors` = as.integer(all_genes %in% suppressor_genes),
  `KEGG Pathways in cancer` = as.integer(all_genes %in% kegg_in_analysis),
  check.names = FALSE
)

# Create UpSet plot
upset(upset_df, 
      sets = c("Metastasis genes", "Metastasis suppressors", "KEGG Pathways in cancer"),
      order.by = "freq",
      sets.bar.color = c("#2E86AB", "#9B2335", "#4CAF50"),
      main.bar.color = "#333333",
      matrix.color = "#333333",
      point.size = 3.5,
      line.size = 1.5,
      text.scale = 1.2,
      mb.ratio = c(0.6, 0.4))

# Save as SVG
svg(here("results/plots/upset_gene_sets_comparison.svg"), width = 12, height = 8)
upset(upset_df, 
      sets = c("Metastasis genes", "Metastasis suppressors", "KEGG Pathways in cancer"),
      order.by = "freq",
      sets.bar.color = c("#2E86AB", "#9B2335", "#4CAF50"),
      main.bar.color = "#333333",
      matrix.color = "#333333",
      point.size = 3.5,
      line.size = 1.5,
      text.scale = 1.2,
      mb.ratio = c(0.6, 0.4))
dev.off()
```

### Overlap Statistics

```{r}
# Calculate overlaps
overlap_meta_supp <- length(intersect(metastasis_genes, suppressor_genes))
overlap_meta_kegg <- length(intersect(metastasis_genes, kegg_in_analysis))
overlap_supp_kegg <- length(intersect(suppressor_genes, kegg_in_analysis))
overlap_all_three <- length(Reduce(intersect, list(metastasis_genes, suppressor_genes, kegg_in_analysis)))

# Print overlap summary
cat("Overlap Summary:\n")
cat("Metastasis genes ∩ Suppressors:", overlap_meta_supp, "\n")
cat("Metastasis genes ∩ KEGG:", overlap_meta_kegg, "\n")
cat("Suppressors ∩ KEGG:", overlap_supp_kegg, "\n")
cat("All three sets:", overlap_all_three, "\n")

# Create overlap summary table
overlap_summary <- data.frame(
  Comparison = c("Metastasis genes ∩ Suppressors",
                 "Metastasis genes ∩ KEGG hsa05200",
                 "Suppressors ∩ KEGG hsa05200",
                 "All three sets"),
  Overlap = c(overlap_meta_supp, overlap_meta_kegg, overlap_supp_kegg, overlap_all_three)
)

overlap_summary
```

### Hypergeometric Enrichment Tests

```{r}
# Universe: all human protein-coding genes (approximate)
universe_size <- 19954

# Function to perform hypergeometric test
hypergeom_test <- function(overlap, set1_size, set2_size, universe) {
  p_val <- phyper(overlap - 1, set2_size, universe - set2_size, set1_size, lower.tail = FALSE)
  fold_enrichment <- (overlap / set1_size) / (set2_size / universe)
  return(list(p_value = p_val, fold_enrichment = fold_enrichment))
}

# Test 1: Metastasis genes vs Suppressors
test1 <- hypergeom_test(overlap_meta_supp, length(metastasis_genes), length(suppressor_genes), universe_size)

# Test 2: Metastasis genes vs KEGG
test2 <- hypergeom_test(overlap_meta_kegg, length(metastasis_genes), length(kegg_in_analysis), universe_size)

# Test 3: Suppressors vs KEGG
test3 <- hypergeom_test(overlap_supp_kegg, length(suppressor_genes), length(kegg_in_analysis), universe_size)

# Create results table
hypergeom_results <- data.frame(
  Comparison = c("Metastasis genes vs Suppressors",
                 "Metastasis genes vs KEGG hsa05200",
                 "Suppressors vs KEGG hsa05200"),
  Set1_Size = c(length(metastasis_genes), length(metastasis_genes), length(suppressor_genes)),
  Set2_Size = c(length(suppressor_genes), length(kegg_in_analysis), length(kegg_in_analysis)),
  Overlap = c(overlap_meta_supp, overlap_meta_kegg, overlap_supp_kegg),
  P_Value = c(test1$p_value, test2$p_value, test3$p_value),
  Fold_Enrichment = c(test1$fold_enrichment, test2$fold_enrichment, test3$fold_enrichment)
)

hypergeom_results$P_Adjusted <- p.adjust(hypergeom_results$P_Value, method = "BH")
hypergeom_results$Significant <- ifelse(hypergeom_results$P_Adjusted < 0.05, "Yes", "No")

hypergeom_results
```

### Fisher's Exact Test

```{r}
# Fisher's exact test for each pairwise comparison
fisher_results <- list()

# Metastasis genes vs Suppressors
contingency_1 <- matrix(c(
  overlap_meta_supp,
  length(metastasis_genes) - overlap_meta_supp,
  length(suppressor_genes) - overlap_meta_supp,
  universe_size - length(metastasis_genes) - length(suppressor_genes) + overlap_meta_supp
), nrow = 2)
fisher_1 <- fisher.test(contingency_1)
fisher_results$meta_vs_supp <- list(odds_ratio = fisher_1$estimate, p_value = fisher_1$p.value)

# Metastasis genes vs KEGG
contingency_2 <- matrix(c(
  overlap_meta_kegg,
  length(metastasis_genes) - overlap_meta_kegg,
  length(kegg_in_analysis) - overlap_meta_kegg,
  universe_size - length(metastasis_genes) - length(kegg_in_analysis) + overlap_meta_kegg
), nrow = 2)
fisher_2 <- fisher.test(contingency_2)
fisher_results$meta_vs_kegg <- list(odds_ratio = fisher_2$estimate, p_value = fisher_2$p.value)

# Suppressors vs KEGG
contingency_3 <- matrix(c(
  overlap_supp_kegg,
  length(suppressor_genes) - overlap_supp_kegg,
  length(kegg_in_analysis) - overlap_supp_kegg,
  universe_size - length(suppressor_genes) - length(kegg_in_analysis) + overlap_supp_kegg
), nrow = 2)
fisher_3 <- fisher.test(contingency_3)
fisher_results$supp_vs_kegg <- list(odds_ratio = fisher_3$estimate, p_value = fisher_3$p.value)

# Create Fisher results table
fisher_df <- data.frame(
  Comparison = c("Metastasis genes vs Suppressors",
                 "Metastasis genes vs KEGG hsa05200",
                 "Suppressors vs KEGG hsa05200"),
  Odds_Ratio = c(fisher_results$meta_vs_supp$odds_ratio,
                 fisher_results$meta_vs_kegg$odds_ratio,
                 fisher_results$supp_vs_kegg$odds_ratio),
  P_Value = c(fisher_results$meta_vs_supp$p_value,
              fisher_results$meta_vs_kegg$p_value,
              fisher_results$supp_vs_kegg$p_value)
)
fisher_df$P_Adjusted <- p.adjust(fisher_df$P_Value, method = "BH")
fisher_df$Significant <- ifelse(fisher_df$P_Adjusted < 0.05, "Yes", "No")

fisher_df
```

### Enrichment Visualization

```{r}
# Bubble plot for enrichment results
enrichment_plot_data <- hypergeom_results %>%
  mutate(
    neg_log_p = -log10(P_Adjusted + 1e-300),
    Comparison = factor(Comparison, levels = rev(Comparison))
  )

ggplot(enrichment_plot_data, aes(x = Fold_Enrichment, y = Comparison)) +
  geom_point(aes(size = Overlap, color = neg_log_p)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray50") +
  scale_color_gradient(low = "gray70", high = "#9B2335", name = "-log10(p.adj)") +
  scale_size_continuous(range = c(3, 12), name = "Overlap Size") +
  labs(x = "Fold Enrichment",
       y = "",
       title = "Hypergeometric Enrichment Analysis",
       subtitle = "Gene set overlaps with expected enrichment") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 10),
    panel.grid.major.y = element_blank()
  )

ggsave(file = here("results/plots/enrichment_bubble_plot.svg"), width = 10, height = 5)
```

### Gene Set Size Comparison Bar Chart

```{r}
# Create bar chart comparing set sizes
set_sizes <- data.frame(
  Gene_Set = c("Metastasis genes", "Metastasis suppressors", "KEGG Pathways in cancer"),
  Count = c(length(metastasis_genes), length(suppressor_genes), length(kegg_in_analysis))
)

set_sizes$Gene_Set <- factor(set_sizes$Gene_Set, levels = set_sizes$Gene_Set)

ggplot(set_sizes, aes(x = Gene_Set, y = Count, fill = Gene_Set)) +
  geom_bar(stat = "identity", alpha = 0.8) +
  geom_text(aes(label = Count), vjust = -0.3, size = 4) +
  scale_fill_manual(values = c("Metastasis genes" = "#2E86AB", 
                                "Metastasis suppressors" = "#9B2335",
                                "KEGG Pathways in cancer" = "#4CAF50")) +
  labs(x = "", y = "Number of Genes",
       title = "Gene Set Sizes Comparison") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 20, hjust = 1, size = 10))

ggsave(file = here("results/plots/gene_set_sizes.svg"), width = 8, height = 6)
```

### Export Comparison Results

```{r}
# Export all comparison results
comparison_results <- list(
  overlap_summary = overlap_summary,
  hypergeometric = hypergeom_results,
  fisher_exact = fisher_df
)

# Save hypergeometric results
hypergeom_results %>%
  vroom::vroom_write(file = here("results/paper_data/hypergeometric_comparison.csv"), delim = ",")

# Save Fisher results
fisher_df %>%
  vroom::vroom_write(file = here("results/paper_data/fisher_comparison.csv"), delim = ",")

# Save overlap summary
overlap_summary %>%
  vroom::vroom_write(file = here("results/paper_data/overlap_summary.csv"), delim = ",")
```
