---
title: "Evolutionary Rooting Analysis and Visualization of Metastasis-Associated Genes"
author: "Gleison M. Azevedo"
format: html
execute:
  cache: true
  echo: true
  message: false
  warning: false
---

## Package Management

```{r pkg-install}
#| cache: false

# Automatically detect and install any missing CRAN packages before loading.
required_packages <- c(
  "dplyr", "tidyr", "ggplot2", "ggraph", "ggforce", "ggtext",
  "scales", "forcats", "igraph", "UpSetR", "here", "vroom", "svglite"
)

missing <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing) > 0) {
  install.packages(missing, repos = "https://cloud.r-project.org")
}

library(dplyr)
library(tidyr)
library(ggplot2)
library(ggraph)
library(ggforce)
library(ggtext)
library(scales)
library(forcats)
library(igraph)
library(UpSetR)
library(here)
library(vroom)
library(svglite)
```

## Helper Functions and Theme Definitions

```{r functions}
# ---------------------------------------------------------------------------
# Color palette mapping each biological process signature to a distinct color,
# used consistently across all plots in the analysis.
# ---------------------------------------------------------------------------
annotation_colors <- c(
  "cell adhesion"                            = "#06141FFF",
  "extracellular matrix organization"        = "#742C14FF",
  "epithelial to mesenchymal transition"     = "#3D4F7DFF",
  "regulation of metallopeptidase activity"  = "#E48C2AFF",
  "cell junction organization"               = "#72874EFF",
  "cellular extravasation"                   = "#046E8FFF",
  "metastasis suppressor"                    = "#9B2335FF",
  "tumor suppressor"                         = "#8B008BFF",
  "oncogene"                                 = "#FF6347FF"
)

# Annotation labels for biological process categories (English)
annotation_labels <- c(
  "cell adhesion"                            = "Cell adhesion",
  "extracellular matrix organization"        = "ECM\norganization",
  "epithelial to mesenchymal transition"     = "Epithelial-mesenchymal\ntransition",
  "regulation of metallopeptidase activity"  = "Metallopeptidase\nactivity regulation",
  "cell junction organization"               = "Cell junction\norganization",
  "cellular extravasation"                   = "Cellular\nextravasation",
  "metastasis suppressor"                    = "Metastasis\nsuppressor",
  "tumor suppressor"                         = "Tumor\nsuppressor",
  "oncogene"                                 = "Oncogene"
)

# Vertical reference line marking the transition to Metazoa
choanoflagellata_line <- geom_vline(
  xintercept = "Sphaeroforma arctica",
  color      = "#FF0000",
  linetype   = "11",
  alpha      = 1,
  linewidth  = 0.25
)

# --- ggplot2 Themes ---

# Main theme for rooting distribution plots
theme_main <- theme(
  panel.spacing      = unit(2.5, "pt"),
  strip.background   = element_blank(),
  panel.grid.major.x = element_blank(),
  panel.grid.major.y = element_line(linewidth = 0.25, linetype = "dotted", color = "#E0E0E0"),
  strip.text.x       = element_text(size = 9, angle = 90, hjust = 0, vjust = 0.5, color = "#757575"),
  strip.text.y       = element_text(size = 10, angle = 0, hjust = 0, vjust = 0.5, color = "#757575"),
  axis.title         = element_text(size = 15, color = "#424242"),
  axis.ticks.x       = element_blank(),
  axis.text.x        = element_blank(),
  axis.text.y        = element_text(size = 5.5),
  legend.position    = "none"
)

# Theme for supplementary figures
theme_supplementary <- theme(
  panel.grid.major.x = element_line(color = "#E0E0E0", linewidth = 0.25, linetype = "dotted"),
  panel.grid.major.y = element_blank(),
  strip.text.y       = element_text(size = 7, angle = 0, hjust = 0, vjust = 0.5, color = "#757575"),
  strip.text.x       = element_text(size = 7, angle = 90, hjust = 0, vjust = 0.5, color = "#757575"),
  axis.title         = element_text(size = 12, color = "#424242"),
  axis.ticks         = element_line(colour = "grey20"),
  axis.text.y        = element_text(size = 6, angle = 0, hjust = 1, vjust = 0.5, color = "#757575"),
  axis.text.x        = element_text(size = 6)
)

# Theme for average-based figures
theme_average <- theme(
  panel.spacing      = unit(1, "pt"),
  axis.title         = element_text(color = "#424242"),
  axis.text          = element_text(color = "#757575"),
  axis.text.x        = element_text(size = 7, angle = -45, vjust = 0, hjust = 0),
  axis.text.y        = element_text(size = 5),
  strip.background   = element_blank(),
  strip.text         = element_text(color = "#757575"),
  strip.text.y       = element_text(angle = 0, hjust = 0, vjust = 0.5)
)

# Theme for large multi-panel figures
theme_big <- theme(
  panel.spacing      = unit(0.5, "pt"),
  panel.grid.major.x = element_line(linewidth = 0.1, linetype = "dashed"),
  panel.grid.major.y = element_blank(),
  strip.background   = element_blank(),
  strip.text.x       = element_text(size = 8, angle = 90, hjust = 0.5, vjust = 0),
  strip.text.y       = element_text(size = 8, angle = 0, hjust = 0, vjust = 0.5),
  axis.text.x        = element_text(size = 6, angle = 90, vjust = 0, hjust = 0),
  axis.text.y        = element_text(size = 4.5),
  axis.ticks         = element_line(size = 0.1)
)

# Custom tick helper for axis breaks
tick_function <- function(x) {
  seq(x[2], 0, length.out = 3) %>% head(-1) %>% tail(-1) %>% { ceiling(./5)*5 }
}

# Compute cumulative gene count across evolutionary clades, ordered from
# the most ancestral to the most recent.
calculate_cumulative_genes <- function(nodelist, node_annotation) {
  all_clades <- node_annotation %>%
    arrange(desc(root)) %>%
    dplyr::select(clade_name) %>%
    unique()

  process_columns <- c("queryItem", "root", "clade_name",
                        "cellular extravasation",
                        "cell junction organization",
                        "cell adhesion",
                        "extracellular matrix organization",
                        "epithelial to mesenchymal transition",
                        "regulation of metallopeptidase activity",
                        "metastasis suppressor",
                        "tumor suppressor",
                        "oncogene")

  cumulative_genes <- nodelist %>%
    arrange(desc(root)) %>%
    dplyr::select(all_of(process_columns)) %>%
    group_by(clade_name, root) %>%
    summarise(count_genes = n(), .groups = "drop") %>%
    arrange(desc(root)) %>%
    mutate(cumulative_sum = cumsum(count_genes)) %>%
    right_join(all_clades, by = "clade_name") %>%
    fill(cumulative_sum, .direction = "down")

  return(cumulative_genes)
}

# Compute cumulative counts per biological process across clades, enabling
# visualization of how each process accumulates over evolutionary time.
calculate_cumulative_bp <- function(nodelist, node_annotation) {
  all_clades <- node_annotation %>%
    arrange(desc(root)) %>%
    dplyr::select(clade_name) %>%
    unique()

  process_columns <- c("queryItem", "root", "clade_name",
                        "cellular extravasation",
                        "cell junction organization",
                        "cell adhesion",
                        "extracellular matrix organization",
                        "epithelial to mesenchymal transition",
                        "regulation of metallopeptidase activity",
                        "metastasis suppressor",
                        "tumor suppressor",
                        "oncogene")

  cumulative_bp <- nodelist %>%
    dplyr::select(all_of(process_columns)) %>%
    distinct(root, queryItem, .keep_all = TRUE) %>%
    mutate(across(all_of(process_columns[-c(1:3)]), ~ as.numeric(.))) %>%
    group_by(root, clade_name) %>%
    summarise(across(all_of(process_columns[-c(1:3)]),
                     ~ sum(. , na.rm = TRUE)),
              .groups = "drop") %>%
    arrange(desc(root)) %>%
    mutate(across(all_of(process_columns[-c(1:3)]), ~ cumsum(.))) %>%
    right_join(all_clades, by = "clade_name") %>%
    fill(everything(), .direction = "down")

  return(cumulative_bp)
}

# Summarize gene counts per biological process within each clade, producing
# proportions suitable for stacked bar chart visualization.
BP_by_clade <- function(data) {
  data %>%
    pivot_longer(
      cols = c(`cellular extravasation`, `cell junction organization`,
               `cell adhesion`, `extracellular matrix organization`,
               `epithelial to mesenchymal transition`,
               `regulation of metallopeptidase activity`,
               `metastasis suppressor`,
               `tumor suppressor`,
               `oncogene`),
      names_to = "process",
      values_to = "presence"
    ) %>%
    filter(presence == 1) %>%
    group_by(clade_name, process, root) %>%
    summarise(count = n(), .groups = "drop") %>%
    mutate(total_genes = sum(count), .by = process) %>%
    mutate(proportion = count / total_genes)
}

# Create a stacked bar chart comparing biological process distributions
# across key evolutionary transition points (Choanoflagellata, Ctenophora,
# Actinopterygii) with intermediate clades aggregated.
plot_stacked_bar_chart <- function(data) {
  clades_of_interest <- c("Choanoflagellata", "Ctenophora", "Actinopterygii")

  filtered_data <- data %>%
    mutate(clade_group = case_when(
      root >= 31 & root <= 37 ~ "Intermediate Metam-Choano",
      clade_name == "Choanoflagellata" ~ "Choanoflagellata",
      clade_name == "Ctenophora" ~ "Ctenophora",
      root >= 21 & root <= 28 ~ "Intermediate Cteno-Actino",
      clade_name == "Actinopterygii" ~ "Actinopterygii",
      TRUE ~ NA_character_
    )) %>%
    filter(!is.na(clade_group)) %>%
    mutate(clade_group = factor(clade_group,
                              levels = c("Intermediate Metam-Choano",
                                         "Choanoflagellata", "Ctenophora",
                                         "Intermediate Cteno-Actino", "Actinopterygii")))

  filtered_data <- filtered_data %>%
    mutate(bar_width = ifelse(grepl("Intermediate", clade_group), 0.4, 0.8)) %>%
    mutate(alpha = ifelse(grepl("Intermediate", clade_group), 0.55, 1))

  ggplot(filtered_data, aes(x = clade_group, y = proportion, fill = process,
                             width = bar_width)) +
    geom_bar(position = "fill", stat = "identity", color = NA, alpha = filtered_data$alpha) +
    theme_minimal() +
    labs(
      title = "Biological Process Distribution",
      x = "",
      y = "Genes Proportion",
      fill = "Biological Processes"
    ) +
    scale_fill_manual(values = annotation_colors) +
    scale_y_continuous(labels = scales::percent) +
    theme(
      axis.text = element_text(size = 12),
      axis.title = element_text(size = 14),
      axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
      legend.position = "right",
      plot.title = element_text(size = 16, hjust = 0.5, face = "bold"),
      panel.grid = element_blank()
    ) +
    guides(alpha = "none")
}

# Perform a hypergeometric enrichment test to assess whether the overlap
# between two gene sets is greater than expected by chance.
hypergeom_test <- function(overlap, set1_size, set2_size, universe) {
  p_val <- phyper(overlap - 1, set2_size, universe - set2_size, set1_size, lower.tail = FALSE)
  fold_enrichment <- (overlap / set1_size) / (set2_size / universe)
  return(list(p_value = p_val, fold_enrichment = fold_enrichment))
}
```

## Data Loading

Gene rooting results and metastasis gene annotations produced by scripts 01
and 02 are loaded. The node list contains clade assignments and binary
indicators for each biological process signature.

```{r data-load}
# Load metastasis gene set (script 01) and rooted node list (script 02)
metastasis_genes <- vroom::vroom(
  here("results/metastasis_genes/metastasis_genes.csv"),
  delim = ",",
  show_col_types = FALSE
)

nodelist <- vroom::vroom(
  here("results/orthology_data/nodelist.csv"),
  show_col_types = FALSE
)

# Extract core annotation columns for reuse
node_annotation <- dplyr::select(
  nodelist, c("queryItem", "cog_id", "root", "clade_name")
)
```

## Gene Distribution Across Biological Processes

An UpSet plot visualizes the overlap structure among gene sets assigned to
each biological process signature, showing how metastasis-related genes
participate in multiple processes simultaneously.

```{r upset-distribution}
#| fig-width: 14
#| fig-height: 7

upset(dplyr::select(as.data.frame(nodelist),
                    "queryItem",
                    "cell adhesion",
                    "extracellular matrix organization",
                    "epithelial to mesenchymal transition",
                    "regulation of metallopeptidase activity",
                    "cell junction organization",
                    "cellular extravasation",
                    "metastasis suppressor",
                    "tumor suppressor",
                    "oncogene"),
      nsets = 50, nintersects = NA,
      sets.bar.color = c("#06141FFF", "#72874EFF", "#9B2335FF", "#3D4F7DFF",
                         "#742C14FF", "#046E8FFF", "#E48C2AFF", "#8B008BFF", "#FF6347FF"),
      mainbar.y.label = "Biological Process \nIntersections",
      sets.x.label = "Set Size")

# Save UpSet plot
plot_dir <- here("results/rooting_analysis/plots")
if (!dir.exists(plot_dir)) dir.create(plot_dir, recursive = TRUE)

svglite::svglite(file.path(plot_dir, "upset_distribution.svg"), width = 14, height = 7)
upset(dplyr::select(as.data.frame(nodelist),
                    "queryItem",
                    "cell adhesion",
                    "extracellular matrix organization",
                    "epithelial to mesenchymal transition",
                    "regulation of metallopeptidase activity",
                    "cell junction organization",
                    "cellular extravasation",
                    "metastasis suppressor",
                    "tumor suppressor",
                    "oncogene"),
      nsets = 50, nintersects = NA,
      sets.bar.color = c("#06141FFF", "#72874EFF", "#9B2335FF", "#3D4F7DFF",
                         "#742C14FF", "#046E8FFF", "#E48C2AFF", "#8B008BFF", "#FF6347FF"),
      mainbar.y.label = "Biological Process \nIntersections",
      sets.x.label = "Set Size")
dev.off()

png(file.path(plot_dir, "upset_distribution.png"), width = 14, height = 7, units = "in", res = 300)
upset(dplyr::select(as.data.frame(nodelist),
                    "queryItem",
                    "cell adhesion",
                    "extracellular matrix organization",
                    "epithelial to mesenchymal transition",
                    "regulation of metallopeptidase activity",
                    "cell junction organization",
                    "cellular extravasation",
                    "metastasis suppressor",
                    "tumor suppressor",
                    "oncogene"),
      nsets = 50, nintersects = NA,
      sets.bar.color = c("#06141FFF", "#72874EFF", "#9B2335FF", "#3D4F7DFF",
                         "#742C14FF", "#046E8FFF", "#E48C2AFF", "#8B008BFF", "#FF6347FF"),
      mainbar.y.label = "Biological Process \nIntersections",
      sets.x.label = "Set Size")
dev.off()
```

## Cumulative Rooting Count

Cumulative gene and biological process counts are computed across the
phylogenetic tree from the most ancestral to the most recent clade. This
reveals the tempo at which metastasis-related genes were acquired over
evolutionary time.

```{r cumulative-rooting}
#| fig-width: 15
#| fig-height: 8

cumulative_genes <- calculate_cumulative_genes(nodelist, node_annotation)
cumulative_bp <- calculate_cumulative_bp(nodelist, node_annotation)

cumulative_data <- left_join(cumulative_genes, cumulative_bp)

long_data <- cumulative_data %>%
  pivot_longer(cols = 5:10,
               names_to = "Process",
               values_to = "Value")

# Plot 1: Cumulative gene count by clade name
p_cumulative <- ggplot() +
  geom_bar(data = cumulative_data,
           aes(x = factor(clade_name, levels = clade_name), y = cumulative_sum),
           stat = "identity", fill = "darkgray", colour = NA) +
  geom_text(data = cumulative_data,
            aes(x = factor(clade_name, levels = clade_name), y = cumulative_sum, label = cumulative_sum),
            vjust = -0.5, size = 3, color = "darkgray") +
  scale_color_manual(values = annotation_colors) +
  labs(x = "Clade Name", y = "Cumulative Sum",
       title = "Cumulative Sum and Biological Processes",
       fill = "Cumulative Sum",
       color = "Biological Processes") +
  theme_main +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

print(p_cumulative)

# Plot 2: Cumulative count with biological process overlay
p_cumulative_bp <- ggplot() +
  geom_bar(data = cumulative_data,
           aes(x = factor(-root), y = cumulative_sum),
           stat = "identity", fill = "darkgray", colour = NA) +
  geom_text(data = cumulative_data,
            aes(x = factor(-root), y = cumulative_sum, label = cumulative_sum),
            vjust = -0.5, size = 3, color = "darkgray") +
  geom_line(data = long_data,
            aes(x = factor(-root), y = Value, color = Process, group = Process),
            size = 1) +
  scale_color_manual(values = annotation_colors) +
  labs(x = "Clade Name", y = "Cumulative Sum",
       title = "Cumulative Sum and Biological Processes",
       fill = "Cumulative Sum",
       color = "Biological Processes") +
  theme_main +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p_cumulative_bp)

# Save plots
ggsave(file.path(plot_dir, "cumulative.svg"), plot = p_cumulative, width = 15, height = 8)
ggsave(file.path(plot_dir, "cumulative.png"), plot = p_cumulative, width = 15, height = 8, dpi = 300)
ggsave(file.path(plot_dir, "cumulative_bp.svg"), plot = p_cumulative_bp, width = 15, height = 8)
ggsave(file.path(plot_dir, "cumulative_bp.png"), plot = p_cumulative_bp, width = 15, height = 8, dpi = 300)
```

## Modified Z-Score Analysis

A modified z-score analysis identifies clades with statistically significant
enrichment or depletion of newly acquired genes relative to the median rate.
The baseline (Metamonada, most ancestral) is excluded from the test. An FDR
threshold of q < 0.05 combined with |z| > 3.5 defines significance.

```{r modified-zscore}
# Count unique genes per clade
calc_gaps <- nodelist %>%
  dplyr::select(root, clade_name, queryItem) %>%
  unique() %>%
  arrange(desc(root))

calc_gaps2 <- calc_gaps %>%
  dplyr::group_by(root, clade_name) %>%
  summarise(n = length(unique(queryItem)), .groups = "drop") %>%
  arrange(desc(root))

# Separate Metamonada (baseline) from the remaining clades
baseline <- calc_gaps2[1, ]
clades <- calc_gaps2[-1, ]

# Compute modified z-scores for non-baseline clades
mediana <- median(clades$n)
mad <- median(abs(clades$n - mediana))

clades <- clades %>%
  mutate(
    modified_z_scores = 0.6745 * (n - mediana) / mad,
    p_value = 2 * pnorm(-abs(modified_z_scores)),
    p_adjusted = p.adjust(p_value, method = "fdr")
  )

# Reconstruct complete data frame including baseline
calc_gaps2$modified_z_scores <- NA
calc_gaps2$p_value <- NA
calc_gaps2$p_adjusted <- NA

calc_gaps2$modified_z_scores[-1] <- clades$modified_z_scores
calc_gaps2$p_value[-1] <- clades$p_value
calc_gaps2$p_adjusted[-1] <- clades$p_adjusted

# Compute cumulative gene count
calc_gaps2$acumulado <- cumsum(calc_gaps2$n)

# Identify significantly enriched/depleted clades
signif_clades <- calc_gaps2 %>%
  filter(!is.na(p_adjusted),
         p_adjusted <= 0.05,
         abs(modified_z_scores) > 3.5)

# Save supplementary table
output_dir <- here("results/rooting_analysis")
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

suppl_table1 <- calc_gaps2 %>%
  filter(!is.na(p_adjusted))

suppl_table1 |>
  vroom::vroom_write(file = file.path(output_dir, "suppl_table1.csv"), delim = ",")

calc_gaps2
signif_clades
```

## Biological Process Proportions by Clade

Genes rooted at or after the Actinopterygii divergence (root ≥ 20) are
classified by clade and colored by node significance. A stacked bar chart
shows the proportional distribution of biological processes at key
evolutionary transition points.

```{r clade-proportion}
#| fig-width: 12
#| fig-height: 7

nodelist_filtered <- nodelist %>%
  mutate(node_color = case_when(
    clade_name == "Choanoflagellata" ~ "#D17878",
    clade_name == "Ctenophora" ~ "#8A6BBE",
    clade_name == "Actinopterygii" ~ "#4A85C2",
    TRUE ~ "#999999FF"
  )) %>%
  mutate(node_size = case_when(
    clade_name %in% c("Choanoflagellata", "Ctenophora", "Actinopterygii") ~ 50,
    TRUE ~ 35
  )) %>%
  mutate(node_linecolor = case_when(
    queryItem %in% metastasis_genes$external_gene_name ~ "#6A0DAD",
    TRUE ~ "#999999FF"
  )) %>%
  mutate(node_linewidth = case_when(
    queryItem %in% metastasis_genes$external_gene_name ~ 3,
    TRUE ~ 1
  )) %>%
  filter(root >= 20)

df_proportion <- BP_by_clade(nodelist_filtered)
plot_stacked_bar_chart(df_proportion)
```

## Hypergeometric Enrichment Test by Clade

A hypergeometric test assesses whether each biological process is
over-represented within each evolutionary clade relative to the full
gene set. P-values are adjusted for multiple testing using the
Benjamini-Hochberg method (FDR < 0.05).

```{r hypergeometric-clade}
#| fig-width: 12
#| fig-height: 8

clades <- unique(nodelist$clade_name)
processes <- c("cell adhesion", "extracellular matrix organization",
               "epithelial to mesenchymal transition",
               "regulation of metallopeptidase activity",
               "cell junction organization", "cellular extravasation",
               "metastasis suppressor",
               "tumor suppressor", "oncogene")

result_list <- list()

for (process in processes) {
  for (clade in clades) {
    N <- length(unique(nodelist$string_id))
    M <- sum(nodelist[, process])
    n <- length(which(nodelist$clade_name == clade))
    k <- sum(filter(nodelist, clade_name == clade)[[process]])

    p_valor <- 1 - phyper(k - 1, M, N - M, n)

    result_list[[paste(process, clade, sep = "_")]] <- c(process, clade, k, M, N, n, p_valor)
  }
}

hipergeometric_df <- do.call(rbind, result_list) %>%
  as.data.frame() %>%
  setNames(c("Process", "clade_name", "k", "M", "N", "n", "p.val")) %>%
  arrange(p.val)
rownames(hipergeometric_df) <- NULL

hipergeometric_df <- hipergeometric_df %>%
  mutate(across(all_of(c("k", "M", "N", "n", "p.val")), as.numeric))

hipergeometric_df <- hipergeometric_df %>%
  mutate(p.adj = p.adjust(p.val, method = "BH"))

# Assign significance color
hipergeometric_df$p.adj <- as.numeric(hipergeometric_df$p.adj)
hipergeometric_df$p.val_color <- ifelse(hipergeometric_df$p.adj > 0.05, "gray", "red")

# Merge with clade root order
i <- nodelist %>%
  dplyr::select(clade_name, root) %>%
  unique()

hipergeometric_df <- hipergeometric_df %>%
  inner_join(i, by = "clade_name")

hipergeometric_df$clade_name <- factor(
  hipergeometric_df$clade_name,
  levels = unique(hipergeometric_df$clade_name[order(hipergeometric_df$root)])
)

hipergeometric_df$Process <- factor(hipergeometric_df$Process,
                                    levels = c("extracellular matrix organization",
                                               "epithelial to mesenchymal transition",
                                               "cell junction organization",
                                               "regulation of metallopeptidase activity",
                                               "cell adhesion",
                                               "cellular extravasation",
                                               "metastasis suppressor",
                                               "tumor suppressor",
                                               "oncogene"))

# Bubble plot: enrichment significance by clade and process
p_hyper <- ggplot(hipergeometric_df, aes(x = Process,
                              y = clade_name,
                              size = as.numeric(n),
                              color = p.val_color)) +
  geom_point() +
  scale_color_manual(values = c("gray" = "gray", "red" = "red"),
                     labels = c("Not Significant", "Significant")) +
  labs(x = "Biological Process", y = "Clade", size = "Number of Orthologs", color = "Significance") +
  scale_size_continuous(breaks = c(10, 50, 100), labels = c("10 Orthologs", "50 Orthologs", "100 Orthologs")) +
  theme_minimal() +
  ggtitle("Hypergeometric Enrichment Test") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_blank()) +
  scale_y_discrete(labels = function(x) {
    x <- gsub("Choanoflagellata", "<b>Choanoflagellata</b>", x)
    x <- gsub("Ctenophora", "<b>Ctenophora</b>", x)
    x <- gsub("Actinopterygii", "<b>Actinopterygii</b>", x)
    return(x)
  }) +
  theme(axis.text.y = ggtext::element_markdown())

print(p_hyper)

# Save hypergeometric results
hipergeometric_df %>%
  vroom::vroom_write(file = file.path(output_dir, "hypergeometric_all_genes.csv"), delim = ",")

# Save plot
ggsave(file.path(plot_dir, "hypergeometric_test.svg"), plot = p_hyper, width = 12, height = 8)
ggsave(file.path(plot_dir, "hypergeometric_test.png"), plot = p_hyper, width = 12, height = 8, dpi = 300)
```

## Comparative Analysis: All Genes vs Metastasis Suppressors

Cumulative rooting distributions are compared between the metastasis-
associated gene set (excluding suppressors) and the metastasis suppressor
gene set. A Kolmogorov-Smirnov test evaluates whether the two groups
exhibit statistically different evolutionary age distributions.

```{r comparative-analysis}
#| fig-width: 12
#| fig-height: 6

# Separate suppressors from other metastasis genes
nodelist_suppressors <- nodelist %>%
  filter(`metastasis suppressor` == 1)

nodelist_no_suppressors <- nodelist %>%
  filter(`metastasis suppressor` == 0 | is.na(`metastasis suppressor`))

# Calculate cumulative percentage for metastasis genes (excluding suppressors)
total_all <- length(unique(nodelist_no_suppressors$queryItem))
calc_cumulative_all <- nodelist_no_suppressors %>%
  dplyr::select(root, clade_name, queryItem) %>%
  unique() %>%
  arrange(desc(root)) %>%
  group_by(root, clade_name) %>%
  summarise(n = n(), .groups = "drop") %>%
  arrange(desc(root)) %>%
  mutate(cumulative = cumsum(n),
         percentage = (cumulative / total_all) * 100,
         dataset = "Metastasis genes")

# Calculate cumulative percentage for suppressors only
total_supp <- length(unique(nodelist_suppressors$queryItem))
calc_cumulative_supp <- nodelist_suppressors %>%
  dplyr::select(root, clade_name, queryItem) %>%
  unique() %>%
  arrange(desc(root)) %>%
  group_by(root, clade_name) %>%
  summarise(n = n(), .groups = "drop") %>%
  arrange(desc(root)) %>%
  mutate(cumulative = cumsum(n),
         percentage = (cumulative / total_supp) * 100,
         dataset = "Metastasis suppressors")

# Combine datasets
cumulative_comparison <- bind_rows(calc_cumulative_all, calc_cumulative_supp)

# Order clade names by evolutionary root
clade_order <- cumulative_comparison %>%
  arrange(desc(root)) %>%
  pull(clade_name) %>%
  unique()

cumulative_comparison$clade_name <- factor(cumulative_comparison$clade_name, levels = clade_order)

# Kolmogorov-Smirnov test comparing evolutionary distributions
nodelist_stats <- nodelist %>%
  mutate(
    group_label = case_when(
      `metastasis suppressor` == 1 ~ "Metastasis suppressors",
      TRUE ~ "Metastasis genes"
    ),
    clade_factor = factor(clade_name, levels = clade_order, ordered = TRUE),
    clade_rank = as.numeric(clade_factor)
  )

ranks_genes <- nodelist_stats %>%
  filter(group_label == "Metastasis genes") %>%
  pull(clade_rank)

ranks_suppressors <- nodelist_stats %>%
  filter(group_label == "Metastasis suppressors") %>%
  pull(clade_rank)

ks_res <- ks.test(ranks_genes, ranks_suppressors)

# Plot with KS test result
stats_label <- paste0(
  "KS Test p-value: ", format.pval(ks_res$p.value, digits = 2, eps = 0.001)
)

plot_final <- ggplot(cumulative_comparison, aes(x = clade_name, y = percentage,
                                  color = dataset, group = dataset)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_color_manual(values = c("Metastasis genes" = "#2E86AB",
                                "Metastasis suppressors" = "#9B2335")) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(x = "Evolutionary Root (Clade)",
       y = "Cumulative Percentage of Genes",
       title = "Cumulative Rooting: Metastasis Genes vs Metastasis Suppressors",
       subtitle = paste0("Metastasis genes: n=", total_all,
                         " | Metastasis suppressors: n=", total_supp,
                         "\n", stats_label),
       color = "Gene Set") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.title = element_text(size = 12),
    legend.position = "bottom",
    panel.grid.major.y = element_line(linewidth = 0.25, linetype = "dotted", color = "#E0E0E0"),
    panel.grid.major.x = element_blank(),
    plot.subtitle = element_text(size = 10, face = "italic", color = "#333333", margin = margin(b = 10))
  )

print(plot_final)

ggsave(file.path(plot_dir, "cumulative_comparison_percentage.svg"), plot = plot_final, width = 12, height = 6)
ggsave(file.path(plot_dir, "cumulative_comparison_percentage.png"), plot = plot_final, width = 12, height = 6, dpi = 300)

# Bar plot for metastasis suppressors only
calc_cumulative_supp$clade_name <- factor(calc_cumulative_supp$clade_name, levels = clade_order)

p_supp_bar <- ggplot(calc_cumulative_supp, aes(x = clade_name, y = cumulative)) +
  geom_bar(stat = "identity", fill = "#9B2335", alpha = 0.8) +
  geom_text(aes(label = cumulative), vjust = -0.3, size = 3) +
  labs(x = "Evolutionary Root (Clade)",
       y = "Cumulative Gene Count",
       title = "Cumulative Rooting: Metastasis Suppressors",
       subtitle = paste0("Total genes: n=", total_supp)) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.title = element_text(size = 12),
    panel.grid.major.y = element_line(linewidth = 0.25, linetype = "dotted", color = "#E0E0E0"),
    panel.grid.major.x = element_blank()
  )

print(p_supp_bar)

ggsave(file.path(plot_dir, "cumulative_suppressors_bar.svg"), plot = p_supp_bar, width = 12, height = 6)
ggsave(file.path(plot_dir, "cumulative_suppressors_bar.png"), plot = p_supp_bar, width = 12, height = 6, dpi = 300)

# Print KS test result
cat("\n--- Statistical Results ---\n")
cat("KS Test p-value:", ks_res$p.value, "\n")
cat("---------------------------\n")
```

## KEGG Pathway Query (hsa05200: Pathways in Cancer)

The KEGG REST API is queried to retrieve all genes annotated in the
"Pathways in Cancer" (hsa05200) pathway, enabling a three-way comparison
between metastasis genes, suppressor genes, and cancer pathway genes.

```{r kegg-query}
#| cache: true

kegg_url <- "https://rest.kegg.jp/link/hsa/hsa05200"
kegg_genes_raw <- tryCatch({
  read.table(url(kegg_url), sep = "\t", header = FALSE, stringsAsFactors = FALSE)
}, error = function(e) {
  message("Error fetching KEGG data: ", e$message)
  return(NULL)
})

if (!is.null(kegg_genes_raw)) {
  kegg_gene_ids <- gsub("hsa:", "", kegg_genes_raw$V2)

  kegg_names_url <- "https://rest.kegg.jp/get/hsa05200"
  pathway_info <- tryCatch({
    readLines(url(kegg_names_url))
  }, error = function(e) {
    message("Error fetching pathway info: ", e$message)
    return(NULL)
  })

  # Parse gene symbols from the KEGG pathway entry
  kegg_symbols <- c()
  in_gene_section <- FALSE
  for (line in pathway_info) {
    if (grepl("^GENE", line)) {
      in_gene_section <- TRUE
      match <- regmatches(line, regexpr("[0-9]+\\s+([A-Za-z0-9-]+)", line))
      if (length(match) > 0) {
        symbol <- gsub("^[0-9]+\\s+", "", match)
        kegg_symbols <- c(kegg_symbols, symbol)
      }
    } else if (in_gene_section && grepl("^\\s+[0-9]+", line)) {
      match <- regmatches(line, regexpr("[0-9]+\\s+([A-Za-z0-9-]+)", line))
      if (length(match) > 0) {
        symbol <- gsub("^[0-9]+\\s+", "", match)
        kegg_symbols <- c(kegg_symbols, symbol)
      }
    } else if (in_gene_section && !grepl("^\\s", line)) {
      break
    }
  }

  kegg_symbols <- unique(kegg_symbols)
  cat("KEGG hsa05200 genes found:", length(kegg_symbols), "\n")
}
```

## Gene Set Comparison: Metastasis vs Suppressors vs KEGG

Three gene sets are compared using an UpSet plot and overlap statistics:
metastasis-associated genes (having at least one of the six biological
processes), metastasis suppressor genes, and KEGG cancer pathway genes.

```{r gene-set-comparison}
#| fig-width: 12
#| fig-height: 8

# Define the 8 biological processes (including cancer gene sets)
biological_processes <- c("cell adhesion",
                          "extracellular matrix organization",
                          "epithelial to mesenchymal transition",
                          "regulation of metallopeptidase activity",
                          "cell junction organization",
                          "cellular extravasation",
                          "tumor suppressor",
                          "oncogene")

# Metastasis genes: at least one biological process
nodelist_bp <- nodelist %>%
  mutate(has_bp = rowSums(dplyr::select(., all_of(biological_processes)), na.rm = TRUE) > 0) %>%
  filter(has_bp)
metastasis_gene_names <- unique(nodelist_bp$queryItem)

# Metastasis suppressors
suppressor_genes <- unique(nodelist_suppressors$queryItem)

# KEGG genes present in the analysis
kegg_in_analysis <- intersect(kegg_symbols, unique(nodelist$queryItem))

# Print set sizes and overlaps
cat("Metastasis genes:", length(metastasis_gene_names), "\n")
cat("Metastasis suppressors:", length(suppressor_genes), "\n")
cat("KEGG hsa05200 genes in analysis:", length(kegg_in_analysis), "\n")

overlap_meta_supp_check <- intersect(metastasis_gene_names, suppressor_genes)
cat("Overlap (Metastasis genes ∩ Suppressors):", length(overlap_meta_supp_check), "\n")

# UpSet plot: three-way gene set comparison
all_genes <- unique(c(metastasis_gene_names, suppressor_genes, kegg_in_analysis))

upset_df <- data.frame(
  gene = all_genes,
  `Metastasis genes` = as.integer(all_genes %in% metastasis_gene_names),
  `Metastasis suppressors` = as.integer(all_genes %in% suppressor_genes),
  `KEGG Pathways in cancer` = as.integer(all_genes %in% kegg_in_analysis),
  check.names = FALSE
)

upset(upset_df,
      sets = c("Metastasis genes", "Metastasis suppressors", "KEGG Pathways in cancer"),
      order.by = "freq",
      sets.bar.color = c("#2E86AB", "#9B2335", "#4CAF50"),
      main.bar.color = "#333333",
      matrix.color = "#333333",
      point.size = 3.5,
      line.size = 1.5,
      text.scale = 1.2,
      mb.ratio = c(0.6, 0.4))

# Save UpSet plot
svglite::svglite(file.path(plot_dir, "upset_gene_sets_comparison.svg"), width = 12, height = 8)
upset(upset_df,
      sets = c("Metastasis genes", "Metastasis suppressors", "KEGG Pathways in cancer"),
      order.by = "freq",
      sets.bar.color = c("#2E86AB", "#9B2335", "#4CAF50"),
      main.bar.color = "#333333",
      matrix.color = "#333333",
      point.size = 3.5,
      line.size = 1.5,
      text.scale = 1.2,
      mb.ratio = c(0.6, 0.4))
dev.off()

png(file.path(plot_dir, "upset_gene_sets_comparison.png"), width = 12, height = 8, units = "in", res = 300)
upset(upset_df,
      sets = c("Metastasis genes", "Metastasis suppressors", "KEGG Pathways in cancer"),
      order.by = "freq",
      sets.bar.color = c("#2E86AB", "#9B2335", "#4CAF50"),
      main.bar.color = "#333333",
      matrix.color = "#333333",
      point.size = 3.5,
      line.size = 1.5,
      text.scale = 1.2,
      mb.ratio = c(0.6, 0.4))
dev.off()

# Calculate and print overlap statistics
overlap_meta_supp <- length(intersect(metastasis_gene_names, suppressor_genes))
overlap_meta_kegg <- length(intersect(metastasis_gene_names, kegg_in_analysis))
overlap_supp_kegg <- length(intersect(suppressor_genes, kegg_in_analysis))
overlap_all_three <- length(Reduce(intersect, list(metastasis_gene_names, suppressor_genes, kegg_in_analysis)))

cat("\nOverlap Summary:\n")
cat("Metastasis genes ∩ Suppressors:", overlap_meta_supp, "\n")
cat("Metastasis genes ∩ KEGG:", overlap_meta_kegg, "\n")
cat("Suppressors ∩ KEGG:", overlap_supp_kegg, "\n")
cat("All three sets:", overlap_all_three, "\n")

overlap_summary <- data.frame(
  Comparison = c("Metastasis genes ∩ Suppressors",
                 "Metastasis genes ∩ KEGG hsa05200",
                 "Suppressors ∩ KEGG hsa05200",
                 "All three sets"),
  Overlap = c(overlap_meta_supp, overlap_meta_kegg, overlap_supp_kegg, overlap_all_three)
)
```

## Hypergeometric Enrichment Test: Gene Set Comparisons

The universe of human protein-coding genes is obtained from the STRING v11.0
database by counting unique gene names for species 9606. Hypergeometric tests
evaluate whether the pairwise overlaps between the three gene sets are
significantly enriched relative to this genome-wide background.

```{r hypergeometric-comparison}
#| cache: true

# Dynamically obtain the universe size from STRING protein.info
protein_info <- vroom::vroom(
  "https://stringdb-static.org/download/protein.info.v11.0/9606.protein.info.v11.0.txt.gz",
  col_types = "ccc",
  delim = "\t",
  comment = "##"
)
universe_size <- length(unique(protein_info$preferred_name))
cat("STRING DB universe size (unique human genes):", universe_size, "\n")

# Memory cleanup
rm(protein_info)
gc()

# Test 1: Metastasis genes vs Suppressors
test1 <- hypergeom_test(overlap_meta_supp, length(metastasis_gene_names), length(suppressor_genes), universe_size)

# Test 2: Metastasis genes vs KEGG
test2 <- hypergeom_test(overlap_meta_kegg, length(metastasis_gene_names), length(kegg_in_analysis), universe_size)

# Test 3: Suppressors vs KEGG
test3 <- hypergeom_test(overlap_supp_kegg, length(suppressor_genes), length(kegg_in_analysis), universe_size)

# Assemble results table
hypergeom_results <- data.frame(
  Comparison = c("Metastasis genes vs Suppressors",
                 "Metastasis genes vs KEGG hsa05200",
                 "Suppressors vs KEGG hsa05200"),
  Set1_Size = c(length(metastasis_gene_names), length(metastasis_gene_names), length(suppressor_genes)),
  Set2_Size = c(length(suppressor_genes), length(kegg_in_analysis), length(kegg_in_analysis)),
  Overlap = c(overlap_meta_supp, overlap_meta_kegg, overlap_supp_kegg),
  Universe_Size = universe_size,
  P_Value = c(test1$p_value, test2$p_value, test3$p_value),
  Fold_Enrichment = c(test1$fold_enrichment, test2$fold_enrichment, test3$fold_enrichment)
)

hypergeom_results$P_Adjusted <- p.adjust(hypergeom_results$P_Value, method = "BH")
hypergeom_results$Significant <- ifelse(hypergeom_results$P_Adjusted < 0.05, "Yes", "No")

hypergeom_results
```

## Export Results

```{r save-outputs}
# Save all analysis outputs to results/rooting_analysis/
hypergeom_results %>%
  vroom::vroom_write(file = file.path(output_dir, "hypergeometric_comparison.csv"), delim = ",")

overlap_summary %>%
  vroom::vroom_write(file = file.path(output_dir, "overlap_summary.csv"), delim = ",")

# Memory cleanup
rm(upset_df, nodelist_bp, nodelist_stats,
   long_data, cumulative_data, cumulative_genes, cumulative_bp)
gc()
```

