---
title: "Exclusive Genes Analysis"
author: "BioME"
format: html
editor: source
---

# Analysis of Exclusive Genes by Biological Process

This script analyzes genes that belong exclusively to one biological process (not shared across multiple processes).

## Loading Libraries

```{r}
#| warning: false
library(tidyverse)
library(vroom)
library(ggplot2)
library(here)
library(ggtext)
library(UpSetR)
library(scales)
```

## Loading Data

```{r}
nodelist <- vroom::vroom(here("results/orthology_data/nodelist.csv"))
metastasis_genes <- vroom(here("results/metastasis_genes/metastasis_genes.csv")) %>%
  subset(Signature != "metastasis suppressor")

# Get clade order from nodelist (ordered by root descending - oldest first)
clade_order <- nodelist %>%
  subset(queryItem %in% unique(metastasis_genes$external_gene_name)) %>%
  dplyr::select(clade_name, root) %>%
  unique() %>%
  arrange(desc(root)) %>%
  pull(clade_name)
```

## Defining Colors and Labels

```{r themes}
# Plotting colors and labels
annotation_colors <- c(
  "cell adhesion"                               = "#06141FFF"
  ,"extracellular matrix organization"          = "#742C14FF"
  ,"epithelial to mesenchymal transition"       = "#3D4F7DFF"
  ,"regulation of metallopeptidase activity"    = "#E48C2AFF"
  ,"cell junction organization"                 = "#72874EFF"
  ,"cellular extravasation"                     = "#046E8FFF"
  ,"metastasis suppressor"                      = "#9B2335FF"
)

# First metazoan reference line
first_metazoan <- 29
```

## Identifying Exclusive Genes

Exclusive genes are those that belong to only ONE biological process (excluding metastasis suppressor for this categorization).

```{r}
# Define the 6 biological processes
biological_processes <- c("cell adhesion", 
                          "extracellular matrix organization", 
                          "epithelial to mesenchymal transition", 
                          "regulation of metallopeptidase activity", 
                          "cell junction organization", 
                          "cellular extravasation")

# Calculate the number of biological processes each gene belongs to
nodelist_bp_count <- nodelist %>%
  mutate(bp_count = rowSums(dplyr::select(., all_of(biological_processes)), na.rm = TRUE))

# Filter for exclusive genes (belong to exactly 1 biological process)
nodelist_exclusive <- nodelist_bp_count %>%
  filter(bp_count == 1)

# Summary of exclusive genes
cat("Total genes:", nrow(nodelist), "\n")
cat("Exclusive genes (belong to 1 BP only):", nrow(nodelist_exclusive), "\n")
cat("Shared genes (belong to 2+ BPs):", nrow(nodelist) - nrow(nodelist_exclusive), "\n")
```

## Distribution of Exclusive Genes by Biological Process

```{r}
# Count exclusive genes per process
exclusive_summary <- nodelist_exclusive %>%
  pivot_longer(
    cols = all_of(biological_processes),
    names_to = "process",
    values_to = "presence"
  ) %>%
  filter(presence == 1) %>%
  group_by(process) %>%
  summarise(n_exclusive = n_distinct(queryItem), .groups = "drop") %>%
  arrange(desc(n_exclusive))

exclusive_summary

# Bar plot of exclusive genes per process
ggplot(exclusive_summary, aes(x = reorder(process, n_exclusive), y = n_exclusive, fill = process)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = n_exclusive), hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_fill_manual(values = annotation_colors) +
  labs(x = "Biological Process",
       y = "Number of Exclusive Genes",
       title = "Exclusive Genes per Biological Process") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text = element_text(size = 10))

ggsave(file = here("results/plots/exclusive_genes_per_process.svg"), width = 10, height = 6)
```

## Modified Z-Score Analysis for Exclusive Genes

```{r}
# Calculate modified Z-score for exclusive genes
calc_modified_zscore_exclusive <- function(nodelist_excl) {
  
  # Get clade order (use the one defined at top level)
  clade_order_df <- nodelist_excl %>%
    dplyr::select(clade_name, root) %>%
    unique() %>%
    arrange(desc(root))
  
  # Count genes per clade for exclusive genes
  genes_per_clade <- nodelist_excl %>%
    dplyr::select(root, clade_name, queryItem) %>%
    unique() %>%
    group_by(root, clade_name) %>%
    summarise(n = n(), .groups = "drop") %>%
    right_join(clade_order_df %>% dplyr::select(clade_name), by = "clade_name") %>%
    replace_na(list(n = 0))
  
  # Calculate modified Z-score
  median_val <- median(genes_per_clade$n, na.rm = TRUE)
  mad_val <- mad(genes_per_clade$n, constant = 1, na.rm = TRUE)
  
  genes_per_clade <- genes_per_clade %>%
    mutate(
      modified_zscore = (n - median_val) / (1.4826 * mad_val),
      significant = abs(modified_zscore) > 3.5
    )
  
  return(genes_per_clade)
}

zscore_exclusive <- calc_modified_zscore_exclusive(nodelist_exclusive)

# Display significant clades
zscore_exclusive %>%
  filter(significant) %>%
  arrange(desc(abs(modified_zscore)))
```

## Modified Z-Score Plot for Exclusive Genes

```{r}
# Order clades by root (use nodelist for root info since we already have it from the z-score calc)
zscore_exclusive$clade_name <- factor(zscore_exclusive$clade_name, levels = clade_order)



ggplot(zscore_exclusive, aes(x = clade_name, y = modified_zscore)) +
  geom_bar(stat = "identity", aes(fill = significant)) +
  geom_hline(yintercept = c(-3.5, 3.5), linetype = "dashed", color = "red") +
  scale_fill_manual(values = c("FALSE" = "gray60", "TRUE" = "#9B2335")) +
  labs(x = "Clade",
       y = "Modified Z-Score",
       title = "Modified Z-Score Analysis: Exclusive Genes",
       fill = "Significant") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
    legend.position = "bottom"
  )

ggsave(file = here("results/plots/zscore_exclusive_genes.svg"), width = 14, height = 6)
```

## Cumulative Rooting Count for Exclusive Genes

```{r}
# clade_order is already defined at file load

# Calculate cumulative for all exclusive genes
cumulative_exclusive <- nodelist_exclusive %>%
  dplyr::select(root, clade_name, queryItem) %>%
  unique() %>%
  arrange(desc(root)) %>%
  group_by(root, clade_name) %>%
  summarise(n = n(), .groups = "drop") %>%
  arrange(desc(root)) %>%
  mutate(cumulative = cumsum(n))

cumulative_exclusive$clade_name <- factor(cumulative_exclusive$clade_name, levels = clade_order)

# Plot cumulative rooting
ggplot(cumulative_exclusive, aes(x = clade_name, y = cumulative)) +
  geom_line(group = 1, color = "#2E86AB", linewidth = 1.2) +
  geom_point(color = "#2E86AB", size = 2) +
  labs(x = "Evolutionary Root (Clade)",
       y = "Cumulative Gene Count",
       title = "Cumulative Rooting: Exclusive Genes",
       subtitle = paste0("Total exclusive genes: n=", nrow(nodelist_exclusive %>% dplyr::select(queryItem) %>% unique()))) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
    panel.grid.major.y = element_line(linewidth = 0.25, linetype = "dotted", color = "#E0E0E0"),
    panel.grid.major.x = element_blank()
  )

ggsave(file = here("results/plots/cumulative_exclusive_genes.svg"), width = 12, height = 6)
```

## Cumulative Rooting by Biological Process (Exclusive Genes Only)

```{r}
# Calculate cumulative per biological process for exclusive genes
cumulative_bp_exclusive <- nodelist_exclusive %>%
  pivot_longer(
    cols = all_of(biological_processes),
    names_to = "process",
    values_to = "presence"
  ) %>%
  filter(presence == 1) %>%
  dplyr::select(root, clade_name, queryItem, process) %>%
  unique() %>%
  arrange(desc(root)) %>%
  group_by(root, clade_name, process) %>%
  summarise(n = n(), .groups = "drop") %>%
  arrange(desc(root)) %>%
  group_by(process) %>%
  mutate(cumulative = cumsum(n))

cumulative_bp_exclusive$clade_name <- factor(cumulative_bp_exclusive$clade_name, levels = clade_order)

# Plot cumulative by BP
ggplot(cumulative_bp_exclusive, aes(x = clade_name, y = cumulative, 
                                     color = process, group = process)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  scale_color_manual(values = annotation_colors) +
  labs(x = "Evolutionary Root (Clade)",
       y = "Cumulative Gene Count",
       title = "Cumulative Rooting by Biological Process: Exclusive Genes Only",
       color = "Biological Process") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
    legend.position = "bottom",
    legend.text = element_text(size = 8),
    panel.grid.major.y = element_line(linewidth = 0.25, linetype = "dotted", color = "#E0E0E0"),
    panel.grid.major.x = element_blank()
  ) +
  guides(color = guide_legend(nrow = 2))

ggsave(file = here("results/plots/cumulative_bp_exclusive.svg"), width = 14, height = 8)
```

## Hypergeometric Test for Exclusive Genes

```{r}
clades <- unique(nodelist_exclusive$clade_name)
processes <- biological_processes

result <- list()

for (process in processes) {
  for (clade in clades) {
    
    N <- length(unique(nodelist_exclusive$string_id))                     # total de genes únicos
    M <- sum(nodelist_exclusive[ ,process])                               # número de genes que participam do processo
    n <- length(which(nodelist_exclusive$clade_name == clade))            # número de genes no clado
    k <- sum(filter(nodelist_exclusive, clade_name == clade)[[process]])  # número de genes do clado que participam do processo
    
    p_valor <- 1 - phyper(k - 1, M, N - M, n)
    
    result[[paste(process, clade, sep = "_")]] <- c(process, clade, k, M, N, n, p_valor)
  }
}

hipergeometric_exclusive_df <- do.call(rbind, result) %>%
  as.data.frame() %>%
  setNames(c("Process", "clade_name", "k", "M", "N", "n", "p.val")) %>% 
  arrange(p.val)

rownames(hipergeometric_exclusive_df) <- NULL

hipergeometric_exclusive_df <- hipergeometric_exclusive_df %>%
  mutate(across(all_of(c("k", "M", "N", "n", "p.val")), as.numeric))   

hipergeometric_exclusive_df <- hipergeometric_exclusive_df %>%
  mutate(p.adj = p.adjust(p.val, method = "BH"))

head(hipergeometric_exclusive_df)
```

## Hypergeometric Test Visualization

```{r}
hipergeometric_exclusive_df$p.adj <- as.numeric(hipergeometric_exclusive_df$p.adj)
hipergeometric_exclusive_df$p.val_color <- ifelse(hipergeometric_exclusive_df$p.adj > 0.05, "gray", "red")

i <- nodelist_exclusive %>% 
  dplyr::select(clade_name, root) %>% 
  unique()

hipergeometric_exclusive_df <- hipergeometric_exclusive_df %>%
  inner_join(i, by = "clade_name")

hipergeometric_exclusive_df$clade_name <- factor(hipergeometric_exclusive_df$clade_name, 
    levels = unique(hipergeometric_exclusive_df$clade_name[order(hipergeometric_exclusive_df$root)]))

hipergeometric_exclusive_df$Process <- factor(hipergeometric_exclusive_df$Process, 
                                    levels = c("extracellular matrix organization", 
                                               "epithelial to mesenchymal transition", 
                                               "cell junction organization",
                                               "regulation of metallopeptidase activity",
                                               "cell adhesion", 
                                               "cellular extravasation"))

ggplot(hipergeometric_exclusive_df, aes(x = Process, 
                              y = clade_name, 
                              size = as.numeric(n), 
                              color = p.val_color)) +
  geom_point() +
  scale_color_manual(values = c("gray" = "gray", "red" = "red"), 
                     labels = c("Not Significant", "Significant")) +  
  labs(x = "Biological Process", 
       y = "Clade", 
       size = "Number of Genes", 
       color = "Significance",
       title = "Hypergeometric Test: Exclusive Genes") +
  scale_size_continuous(breaks = c(5, 20, 50), labels = c("5 genes", "20 genes", "50 genes")) + 
  theme_minimal() +
  ggtitle("Hypergeometric Test: Exclusive Genes") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_blank()) +
  scale_y_discrete(labels = function(x) {
    x <- gsub("Choanoflagellata", "<b>Choanoflagellata</b>", x)
    x <- gsub("Ctenophora", "<b>Ctenophora</b>", x)
    x <- gsub("Actinopterygii", "<b>Actinopterygii</b>", x)
    return(x)
  }) +
  theme(axis.text.y = ggtext::element_markdown())

ggsave(file = here("results/plots/hipergeometric_exclusive.svg"), width = 12, height = 10)
```


## Export Exclusive Genes Data

```{r}
# Export exclusive genes list
nodelist_exclusive %>%
  vroom::vroom_write(file = here("results/orthology_data/nodelist_exclusive.csv"), delim = ",")

# Export hypergeometric results
hipergeometric_exclusive_df %>%
  vroom::vroom_write(file = here("results/paper_data/hipergeometric_exclusive.csv"), delim = ",")

# Export z-score results
zscore_exclusive %>%
  vroom::vroom_write(file = here("results/paper_data/zscore_exclusive.csv"), delim = ",")
```
