---
title: "Exclusive Genes Analysis: Evolutionary Rooting of Single-Process Metastasis Genes"
author: "Gleison M. Azevedo"
format: html
execute:
  cache: true
  echo: true
  message: false
  warning: false
---

## Package Management

```{r pkg-install}
#| cache: false

# Automatically detect and install any missing CRAN packages before loading.
required_packages <- c(
  "tidyverse", "vroom", "ggplot2", "ggtext", "UpSetR",
  "scales", "here", "svglite"
)

missing <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing) > 0) {
  install.packages(missing, repos = "https://cloud.r-project.org")
}

library(tidyverse)
library(vroom)
library(ggplot2)
library(ggtext)
library(UpSetR)
library(scales)
library(here)
library(svglite)
```

## Helper Functions and Theme Definitions

```{r functions}
# ---------------------------------------------------------------------------
# Color palette mapping each biological process to a distinct color,
# consistent with the other scripts in the analysis pipeline.
# ---------------------------------------------------------------------------
annotation_colors <- c(
  "cell adhesion"                            = "#06141FFF",
  "extracellular matrix organization"        = "#742C14FF",
  "epithelial to mesenchymal transition"     = "#3D4F7DFF",
  "regulation of metallopeptidase activity"  = "#E48C2AFF",
  "cell junction organization"               = "#72874EFF",
  "cellular extravasation"                   = "#046E8FFF",
  "metastasis suppressor"                    = "#9B2335FF"
)

# Reference root value for the first metazoan
first_metazoan <- 29

# Compute modified z-scores for gene counts per clade to identify clades
# with statistically significant enrichment or depletion of exclusive genes.
# Clades with zero genes are included via right join.
calc_modified_zscore_exclusive <- function(nodelist_excl, clade_order_df) {
  # Count genes per clade
  genes_per_clade <- nodelist_excl %>%
    dplyr::select(root, clade_name, queryItem) %>%
    unique() %>%
    group_by(root, clade_name) %>%
    summarise(n = n(), .groups = "drop") %>%
    right_join(clade_order_df %>% dplyr::select(clade_name), by = "clade_name") %>%
    replace_na(list(n = 0))

  # Modified z-score calculation
  median_val <- median(genes_per_clade$n, na.rm = TRUE)
  mad_val <- mad(genes_per_clade$n, constant = 1, na.rm = TRUE)

  genes_per_clade <- genes_per_clade %>%
    mutate(
      modified_zscore = (n - median_val) / (1.4826 * mad_val),
      significant = abs(modified_zscore) > 3.5
    )

  return(genes_per_clade)
}
```

## Data Loading

Rooted node list and metastasis gene annotations produced by scripts 01
and 02 are loaded. Metastasis suppressor genes are excluded from this
analysis, which focuses on the six core biological processes.

```{r data-load}
nodelist <- vroom::vroom(
  here("results/orthology_data/nodelist.csv"),
  show_col_types = FALSE
)

metastasis_genes <- vroom::vroom(
  here("results/metastasis_genes/metastasis_genes.csv"),
  show_col_types = FALSE
) %>%
  subset(Signature != "metastasis suppressor")

# Establish clade ordering (most ancestral first) restricted to metastasis genes
clade_order <- nodelist %>%
  subset(queryItem %in% unique(metastasis_genes$external_gene_name)) %>%
  dplyr::select(clade_name, root) %>%
  unique() %>%
  arrange(desc(root)) %>%
  pull(clade_name)

# Clade order as a data frame for z-score joins
clade_order_df <- nodelist %>%
  subset(queryItem %in% unique(metastasis_genes$external_gene_name)) %>%
  dplyr::select(clade_name, root) %>%
  unique() %>%
  arrange(desc(root))

# Create output directories
output_dir <- here("results/exclusive_genes")
plot_dir   <- file.path(output_dir, "plots")
if (!dir.exists(plot_dir)) dir.create(plot_dir, recursive = TRUE)
```

## Identifying Exclusive Genes

Exclusive genes are defined as those participating in exactly one of the
six core biological processes (cell adhesion, ECM organization, EMT,
metallopeptidase regulation, cell junction organization, or cellular
extravasation). This filtering removes multi-process genes to enable
process-specific evolutionary analysis.

```{r identify-exclusive}
# Define the 6 biological processes
biological_processes <- c("cell adhesion",
                          "extracellular matrix organization",
                          "epithelial to mesenchymal transition",
                          "regulation of metallopeptidase activity",
                          "cell junction organization",
                          "cellular extravasation")

# Count biological processes per gene
nodelist_bp_count <- nodelist %>%
  mutate(bp_count = rowSums(dplyr::select(., all_of(biological_processes)), na.rm = TRUE))

# Filter for exclusive genes (exactly 1 biological process)
nodelist_exclusive <- nodelist_bp_count %>%
  filter(bp_count == 1)

# Summary statistics
cat("Total genes:", nrow(nodelist), "\n")
cat("Exclusive genes (belong to 1 BP only):", nrow(nodelist_exclusive), "\n")
cat("Shared genes (belong to 2+ BPs):", nrow(nodelist) - nrow(nodelist_exclusive), "\n")
```

## Distribution of Exclusive Genes by Biological Process

A horizontal bar chart shows the number of exclusive genes assigned to each
biological process, revealing which processes have the most genes not shared
with other functional categories.

```{r exclusive-distribution}
#| fig-width: 10
#| fig-height: 6

# Count exclusive genes per process
exclusive_summary <- nodelist_exclusive %>%
  pivot_longer(
    cols = all_of(biological_processes),
    names_to = "process",
    values_to = "presence"
  ) %>%
  filter(presence == 1) %>%
  group_by(process) %>%
  summarise(n_exclusive = n_distinct(queryItem), .groups = "drop") %>%
  arrange(desc(n_exclusive))

exclusive_summary

# Bar plot
p_excl_bar <- ggplot(exclusive_summary, aes(x = reorder(process, n_exclusive),
                                             y = n_exclusive, fill = process)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = n_exclusive), hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_fill_manual(values = annotation_colors) +
  labs(x = "Biological Process",
       y = "Number of Exclusive Genes",
       title = "Exclusive Genes per Biological Process") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text = element_text(size = 10))

print(p_excl_bar)

ggsave(file.path(plot_dir, "exclusive_genes_per_process.svg"), plot = p_excl_bar, width = 10, height = 6)
ggsave(file.path(plot_dir, "exclusive_genes_per_process.png"), plot = p_excl_bar, width = 10, height = 6, dpi = 300)
```

## Modified Z-Score Analysis for Exclusive Genes

A modified z-score identifies clades where exclusive gene acquisition
departs significantly from the median rate. Clades with |z| > 3.5 are
flagged as significant, indicating bursts or gaps in the acquisition of
single-process genes.

```{r modified-zscore}
#| fig-width: 14
#| fig-height: 6

zscore_exclusive <- calc_modified_zscore_exclusive(nodelist_exclusive, clade_order_df)

# Display significant clades
zscore_exclusive %>%
  filter(significant) %>%
  arrange(desc(abs(modified_zscore)))

# Order clades for plotting
zscore_exclusive$clade_name <- factor(zscore_exclusive$clade_name, levels = clade_order)

# Z-score bar plot
p_zscore <- ggplot(zscore_exclusive, aes(x = clade_name, y = modified_zscore)) +
  geom_bar(stat = "identity", aes(fill = significant)) +
  geom_hline(yintercept = c(-3.5, 3.5), linetype = "dashed", color = "red") +
  scale_fill_manual(values = c("FALSE" = "gray60", "TRUE" = "#9B2335")) +
  labs(x = "Clade",
       y = "Modified Z-Score",
       title = "Modified Z-Score Analysis: Exclusive Genes",
       fill = "Significant") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
    legend.position = "bottom"
  )

print(p_zscore)

ggsave(file.path(plot_dir, "zscore_exclusive_genes.svg"), plot = p_zscore, width = 14, height = 6)
ggsave(file.path(plot_dir, "zscore_exclusive_genes.png"), plot = p_zscore, width = 14, height = 6, dpi = 300)
```

## Cumulative Rooting Count for Exclusive Genes

Cumulative gene count is computed from the most ancestral to the most recent
clade, showing the tempo of exclusive gene acquisition across evolutionary
time for the entire set.

```{r cumulative-rooting}
#| fig-width: 12
#| fig-height: 6

cumulative_exclusive <- nodelist_exclusive %>%
  dplyr::select(root, clade_name, queryItem) %>%
  unique() %>%
  arrange(desc(root)) %>%
  group_by(root, clade_name) %>%
  summarise(n = n(), .groups = "drop") %>%
  arrange(desc(root)) %>%
  mutate(cumulative = cumsum(n))

cumulative_exclusive$clade_name <- factor(cumulative_exclusive$clade_name, levels = clade_order)

# Cumulative line plot
p_cumul <- ggplot(cumulative_exclusive, aes(x = clade_name, y = cumulative)) +
  geom_line(group = 1, color = "#2E86AB", linewidth = 1.2) +
  geom_point(color = "#2E86AB", size = 2) +
  labs(x = "Evolutionary Root (Clade)",
       y = "Cumulative Gene Count",
       title = "Cumulative Rooting: Exclusive Genes",
       subtitle = paste0("Total exclusive genes: n=",
                         nrow(nodelist_exclusive %>% dplyr::select(queryItem) %>% unique()))) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
    panel.grid.major.y = element_line(linewidth = 0.25, linetype = "dotted", color = "#E0E0E0"),
    panel.grid.major.x = element_blank()
  )

print(p_cumul)

ggsave(file.path(plot_dir, "cumulative_exclusive_genes.svg"), plot = p_cumul, width = 12, height = 6)
ggsave(file.path(plot_dir, "cumulative_exclusive_genes.png"), plot = p_cumul, width = 12, height = 6, dpi = 300)
```

## Cumulative Rooting by Biological Process (Exclusive Genes Only)

Per-process cumulative rooting reveals how the exclusive gene complement
for each biological process accumulated over evolutionary time, enabling
comparison of acquisition timing across functional categories.

```{r cumulative-by-process}
#| fig-width: 14
#| fig-height: 8

cumulative_bp_exclusive <- nodelist_exclusive %>%
  pivot_longer(
    cols = all_of(biological_processes),
    names_to = "process",
    values_to = "presence"
  ) %>%
  filter(presence == 1) %>%
  dplyr::select(root, clade_name, queryItem, process) %>%
  unique() %>%
  arrange(desc(root)) %>%
  group_by(root, clade_name, process) %>%
  summarise(n = n(), .groups = "drop") %>%
  arrange(desc(root)) %>%
  group_by(process) %>%
  mutate(cumulative = cumsum(n))

cumulative_bp_exclusive$clade_name <- factor(cumulative_bp_exclusive$clade_name, levels = clade_order)

# Multi-line cumulative plot by process
p_cumul_bp <- ggplot(cumulative_bp_exclusive, aes(x = clade_name, y = cumulative,
                                     color = process, group = process)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  scale_color_manual(values = annotation_colors) +
  labs(x = "Evolutionary Root (Clade)",
       y = "Cumulative Gene Count",
       title = "Cumulative Rooting by Biological Process: Exclusive Genes Only",
       color = "Biological Process") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
    legend.position = "bottom",
    legend.text = element_text(size = 8),
    panel.grid.major.y = element_line(linewidth = 0.25, linetype = "dotted", color = "#E0E0E0"),
    panel.grid.major.x = element_blank()
  ) +
  guides(color = guide_legend(nrow = 2))

print(p_cumul_bp)

ggsave(file.path(plot_dir, "cumulative_bp_exclusive.svg"), plot = p_cumul_bp, width = 14, height = 8)
ggsave(file.path(plot_dir, "cumulative_bp_exclusive.png"), plot = p_cumul_bp, width = 14, height = 8, dpi = 300)
```

## Hypergeometric Enrichment Test for Exclusive Genes

A hypergeometric test assesses whether each biological process is
over-represented among exclusive genes within each evolutionary clade.
P-values are corrected for multiple comparisons using the
Benjamini-Hochberg method (FDR < 0.05).

```{r hypergeometric-test}
#| fig-width: 12
#| fig-height: 10

clades <- unique(nodelist_exclusive$clade_name)
processes <- biological_processes

result <- list()

for (process in processes) {
  for (clade in clades) {
    N <- length(unique(nodelist_exclusive$string_id))             # total unique genes
    M <- sum(nodelist_exclusive[, process])                       # genes in this process
    n <- length(which(nodelist_exclusive$clade_name == clade))    # genes in this clade
    k <- sum(filter(nodelist_exclusive, clade_name == clade)[[process]])  # clade genes in process

    p_valor <- 1 - phyper(k - 1, M, N - M, n)

    result[[paste(process, clade, sep = "_")]] <- c(process, clade, k, M, N, n, p_valor)
  }
}

hipergeometric_exclusive_df <- do.call(rbind, result) %>%
  as.data.frame() %>%
  setNames(c("Process", "clade_name", "k", "M", "N", "n", "p.val")) %>%
  arrange(p.val)
rownames(hipergeometric_exclusive_df) <- NULL

hipergeometric_exclusive_df <- hipergeometric_exclusive_df %>%
  mutate(across(all_of(c("k", "M", "N", "n", "p.val")), as.numeric))

hipergeometric_exclusive_df <- hipergeometric_exclusive_df %>%
  mutate(p.adj = p.adjust(p.val, method = "BH"))

# Assign significance color for plotting
hipergeometric_exclusive_df$p.adj <- as.numeric(hipergeometric_exclusive_df$p.adj)
hipergeometric_exclusive_df$p.val_color <- ifelse(hipergeometric_exclusive_df$p.adj > 0.05, "gray", "red")

# Merge with clade root order
i <- nodelist_exclusive %>%
  dplyr::select(clade_name, root) %>%
  unique()

hipergeometric_exclusive_df <- hipergeometric_exclusive_df %>%
  inner_join(i, by = "clade_name")

hipergeometric_exclusive_df$clade_name <- factor(
  hipergeometric_exclusive_df$clade_name,
  levels = unique(hipergeometric_exclusive_df$clade_name[order(hipergeometric_exclusive_df$root)])
)

hipergeometric_exclusive_df$Process <- factor(
  hipergeometric_exclusive_df$Process,
  levels = c("extracellular matrix organization",
             "epithelial to mesenchymal transition",
             "cell junction organization",
             "regulation of metallopeptidase activity",
             "cell adhesion",
             "cellular extravasation")
)

# Bubble plot: enrichment significance by clade and process
p_hyper <- ggplot(hipergeometric_exclusive_df, aes(x = Process,
                                                    y = clade_name,
                                                    size = as.numeric(n),
                                                    color = p.val_color)) +
  geom_point() +
  scale_color_manual(values = c("gray" = "gray", "red" = "red"),
                     labels = c("Not Significant", "Significant")) +
  labs(x = "Biological Process",
       y = "Clade",
       size = "Number of Genes",
       color = "Significance",
       title = "Hypergeometric Test: Exclusive Genes") +
  scale_size_continuous(breaks = c(5, 20, 50), labels = c("5 genes", "20 genes", "50 genes")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_blank()) +
  scale_y_discrete(labels = function(x) {
    x <- gsub("Choanoflagellata", "<b>Choanoflagellata</b>", x)
    x <- gsub("Ctenophora", "<b>Ctenophora</b>", x)
    x <- gsub("Actinopterygii", "<b>Actinopterygii</b>", x)
    return(x)
  }) +
  theme(axis.text.y = ggtext::element_markdown())

print(p_hyper)

ggsave(file.path(plot_dir, "hypergeometric_exclusive.svg"), plot = p_hyper, width = 12, height = 10)
ggsave(file.path(plot_dir, "hypergeometric_exclusive.png"), plot = p_hyper, width = 12, height = 10, dpi = 300)
```

## Export Results and Equivalence Checks

```{r save-outputs}
#| cache: false

# Save exclusive gene list
nodelist_exclusive %>%
  vroom::vroom_write(file = file.path(output_dir, "nodelist_exclusive.csv"), delim = ",")

# Save hypergeometric results
hipergeometric_exclusive_df %>%
  vroom::vroom_write(file = file.path(output_dir, "hypergeometric_exclusive.csv"), delim = ",")

# Save z-score results
zscore_exclusive %>%
  vroom::vroom_write(file = file.path(output_dir, "zscore_exclusive.csv"), delim = ",")

# --- Equivalence Checks (temporary) ---
trim_char_cols <- function(df) {
  df[] <- lapply(df, function(col) {
    if (is.character(col)) trimws(col) else col
  })
  df
}
