---
title: "Metastasis Suppressor Genes (MSG) Enrichment Analysis with TreeAndLeaf"
author: "Gleison M. Azevedo"
format: html
execute:
  cache: true
  echo: true
  message: false
  warning: false
---

## Package Management

```{r pkg-install}
#| cache: false

# Automatically detect and install any missing packages before loading.
# Includes both CRAN and Bioconductor packages.
required_cran <- c(
  "tidyverse", "vroom", "here", "RColorBrewer", "classInt", "UpSetR", "svglite", "igraph", "scales", "ggtext"
)

required_bioc <- c(
  "TreeAndLeaf", "RedeR", "clusterProfiler", "org.Hs.eg.db", "GOSemSim"
)

# Install CRAN packages
missing_cran <- required_cran[
  !vapply(required_cran, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_cran) > 0) {
  install.packages(missing_cran, repos = "https://cloud.r-project.org")
}

# Install Bioconductor packages
missing_bioc <- required_bioc[
  !vapply(required_bioc, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_bioc) > 0) {
  if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager", repos = "https://cloud.r-project.org")
  BiocManager::install(missing_bioc, update = FALSE)
}

library(TreeAndLeaf)
library(RedeR)
library(igraph)
library(tidyverse)
library(here)
library(vroom)
library(RColorBrewer)
library(clusterProfiler)
library(org.Hs.eg.db)
library(GOSemSim)
library(classInt)
library(UpSetR)
library(svglite)
```

## Data Integrity Checksum

```{r checksum}
#| cache: false

# Verify MD5 checksum of critical input files to ensure data integrity.
expected_checksums <- list(
  "assets/metastasis_supressors_nature.csv" = "6f1717e97a277fb0fed475e92647d214"
)

for (file_path in names(expected_checksums)) {
  full_path <- here(file_path)
  if (file.exists(full_path)) {
    actual_checksum <- tools::md5sum(full_path)
    if (actual_checksum != expected_checksums[[file_path]]) {
      warning(paste("Checksum mismatch for", file_path, 
                    "\nExpected:", expected_checksums[[file_path]], 
                    "\nActual:  ", actual_checksum))
    } else {
      message(paste("Checksum verified for", file_path))
    }
  } else {
    warning(paste("File not found:", file_path))
  }
}
```

## Helper Functions and Configuration

```{r functions}
# Define Evolutionary Group Colors
group_colors <- c(
  "Unicellular" = "#E69F00",                # Orange
  "Multicellular non-vertebrate" = "#0072B2", # Blue
  "Vertebrates" = "#009E73"                  # Green
)

# Visualization sizing parameters
sizeMultiplier <- 15
sizeBase <- 50

# --- Function: GO Enrichment ---
perform_go_enrichment <- function(genes, group_name) {
  cat(sprintf("  Processing %s (%d genes)...\n", group_name, length(genes)))
  
  if (length(genes) == 0) return(NULL)

  # Convert gene symbols to Entrez IDs
  gene_ids <- suppressMessages(bitr(
    genes,
    fromType = "SYMBOL",
    toType = "ENTREZID",
    OrgDb = org.Hs.eg.db
  ))

  if (nrow(gene_ids) == 0) {
    cat(sprintf("    No genes mapped for %s\n", group_name))
    return(NULL)
  }

  # Perform GO enrichment
  ego <- enrichGO(
    gene = gene_ids$ENTREZID,
    OrgDb = org.Hs.eg.db,
    ont = "BP",
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05,
    qvalueCutoff = 0.05,
    readable = TRUE
  )

  if (is.null(ego) || nrow(as.data.frame(ego)) == 0) {
    cat(sprintf("    No enriched terms for %s\n", group_name))
    return(NULL)
  }

  result <- as.data.frame(ego) %>%
    mutate(evolutionary_group = group_name)

  cat(sprintf("    Found %d enriched GO terms\n", nrow(result)))
  return(result)
}

# --- Function: KEGG Enrichment ---
perform_kegg_enrichment <- function(genes, group_name) {
  cat(sprintf("  Processing %s (%d genes)...\n", group_name, length(genes)))
  
  if (length(genes) == 0) return(NULL)

  # Convert gene symbols to Entrez IDs
  gene_ids <- suppressMessages(bitr(
    genes,
    fromType = "SYMBOL",
    toType = "ENTREZID",
    OrgDb = org.Hs.eg.db
  ))

  if (nrow(gene_ids) == 0) {
    cat(sprintf("    No genes mapped for %s\n", group_name))
    return(NULL)
  }

  # Perform KEGG enrichment
  ekegg <- enrichKEGG(
    gene = gene_ids$ENTREZID,
    organism = "hsa",
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05,
    qvalueCutoff = 0.05
  )

  if (is.null(ekegg) || nrow(as.data.frame(ekegg)) == 0) {
    cat(sprintf("    No enriched KEGG pathways for %s\n", group_name))
    return(NULL)
  }

  # Convert to readable gene symbols
  ekegg <- setReadable(ekegg, OrgDb = org.Hs.eg.db, keyType = "ENTREZID")

  result <- as.data.frame(ekegg) %>%
    mutate(evolutionary_group = group_name)

  cat(sprintf("    Found %d enriched KEGG pathways\n", nrow(result)))
  return(result)
}

# --- Function: Filter Human Diseases ---
filter_human_diseases <- function(df) {
  if (is.null(df) || nrow(df) == 0) return(df)
  df_filtered <- df %>%
    filter(!str_detect(ID, "^hsa05"))
  cat(sprintf(
    "  Removed %d Human Disease pathways, %d remaining\n",
    nrow(df) - nrow(df_filtered), nrow(df_filtered)
  ))
  if (nrow(df_filtered) == 0) return(NULL)
  return(df_filtered)
}
```

## Data Loading

```{r data-load}
# Nodelist with rooting information and biological process annotations
nodelist <- vroom::vroom(
  here("results/orthology_data/nodelist.csv"),
  show_col_types = FALSE
)

# Metastasis suppressor genes reference list
msg_reference <- vroom::vroom(
  here("assets/metastasis_supressors_nature.csv"),
  delim = "\t",
  show_col_types = FALSE
)

# Create output directories
output_base <- here("results/enrichment")
dir.create(file.path(output_base, "tables"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(output_base, "plots"), recursive = TRUE, showWarnings = FALSE)
```

## Filter and Classify Metastasis Suppressor Genes

Define evolutionary groups based on root values:
- **Unicellular**: Root >= 30 (before Choanoflagellata)
- **Multicellular non-vertebrate**: 21 <= Root < 30 (Choanoflagellata to Tunicata)
- **Vertebrates**: Root <= 20 (Actinopterygii and newer)

```{r filter-msg}
# Filter nodelist to keep only metastasis suppressor genes
msg_nodelist <- nodelist %>%
  filter(`metastasis suppressor` == 1) %>%
  dplyr::select(cog_id, root, clade_name, string_id, queryItem)

cat("Total metastasis suppressor genes in nodelist:", nrow(msg_nodelist), "\n")

# Assign Evolutionary Groups Based on Root Values
msg_nodelist <- msg_nodelist %>%
  mutate(
    evolutionary_group = case_when(
      root => 30 ~ "Unicellular",
      root >= 21 & root < 30 ~ "Multicellular non-vertebrate",
      root <= 20 ~ "Vertebrates"
    )
  ) %>%
  mutate(
    group_color = group_colors[evolutionary_group]
  )

# Summary of genes per group
cat("\n--- Gene Distribution by Evolutionary Group ---\n")
print(table(msg_nodelist$evolutionary_group))

# Extract unique genes per group
genes_unicellular <- unique(msg_nodelist$queryItem[msg_nodelist$evolutionary_group == "Unicellular"])
genes_multicellular <- unique(msg_nodelist$queryItem[msg_nodelist$evolutionary_group == "Multicellular non-vertebrate"])
genes_vertebrates <- unique(msg_nodelist$queryItem[msg_nodelist$evolutionary_group == "Vertebrates"])
```

## GO Enrichment Analysis

Perform Gene Ontology (Biological Process) enrichment analysis for each evolutionary group.

```{r go-enrichment}
#| cache: true

set.seed(42) # Ensure reproducibility

# Perform enrichment for each group
go_unicellular <- perform_go_enrichment(genes_unicellular, "Unicellular")
go_multicellular <- perform_go_enrichment(genes_multicellular, "Multicellular non-vertebrate")
go_vertebrates <- perform_go_enrichment(genes_vertebrates, "Vertebrates")

# Combine all results
go_combined <- bind_rows(
  go_unicellular,
  go_multicellular,
  go_vertebrates
) %>%
  filter(!is.na(ID))

if (nrow(go_combined) == 0) {
  stop("No enriched GO terms found in any group.")
}

cat(sprintf(
  "\nTotal enriched GO terms across all groups: %d\n",
  nrow(go_combined)
))

# Save enrichment tables
vroom::vroom_write(go_combined, file.path(output_base, "tables/MSG_GO_enrichment_combined.csv"), delim = ",")
if (!is.null(go_unicellular)) vroom::vroom_write(go_unicellular, file.path(output_base, "tables/MSG_GO_enrichment_unicellular.csv"), delim = ",")
if (!is.null(go_multicellular)) vroom::vroom_write(go_multicellular, file.path(output_base, "tables/MSG_GO_enrichment_multicellular.csv"), delim = ",")
if (!is.null(go_vertebrates)) vroom::vroom_write(go_vertebrates, file.path(output_base, "tables/MSG_GO_enrichment_vertebrates.csv"), delim = ",")
```

## GO Enrichment Visualization

```{r go-plots}
#| fig-width: 12
#| fig-height: 14

# Select top 15 GO terms per group by p.adjust
go_top <- go_combined %>%
  group_by(evolutionary_group) %>%
  slice_min(p.adjust, n = 15, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    neg_log10_padj = -log10(p.adjust),
    evolutionary_group = factor(
      evolutionary_group,
      levels = c("Vertebrates", "Multicellular non-vertebrate", "Unicellular")
    )
  )

# Order descriptions within each group for cleaner plotting
go_top <- go_top %>%
  arrange(evolutionary_group, neg_log10_padj) %>%
  mutate(Description = factor(Description, levels = unique(Description)))

# Plot: All Groups
p_go_all <- ggplot(go_top, aes(x = neg_log10_padj, y = Description, fill = evolutionary_group)) +
  geom_bar(stat = "identity", width = 0.7) +
  facet_grid(evolutionary_group ~ ., scales = "free_y", space = "free_y") +
  scale_fill_manual(values = group_colors, name = "Evolutionary Group") +
  labs(
    title = "GO Biological Process Enrichment — Metastasis Suppressor Genes",
    subtitle = "Top 15 terms per evolutionary group",
    x = expression(-log[10](p.adjust)),
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text.y = element_text(angle = 0, face = "bold", size = 11),
    legend.position = "none",
    panel.grid.major.y = element_blank(),
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(color = "gray40")
  )

print(p_go_all)
ggsave(file.path(output_base, "plots/MSG_GO_barplot_all_groups.svg"), p_go_all, width = 12, height = 14)
ggsave(file.path(output_base, "plots/MSG_GO_barplot_all_groups.png"), p_go_all, width = 12, height = 14, dpi = 300)

# Plot: Vertebrates Only
go_vert <- go_combined %>%
  filter(evolutionary_group == "Vertebrates") %>%
  slice_min(p.adjust, n = 20, with_ties = FALSE) %>%
  mutate(neg_log10_padj = -log10(p.adjust)) %>%
  arrange(neg_log10_padj) %>%
  mutate(Description = factor(Description, levels = unique(Description)))

if (nrow(go_vert) > 0) {
  p_go_vert <- ggplot(go_vert, aes(x = neg_log10_padj, y = Description)) +
    geom_bar(stat = "identity", fill = group_colors["Vertebrates"], width = 0.7) +
    labs(
      title = "GO Biological Process Enrichment — Vertebrates MSGs",
      subtitle = "Top 20 enriched terms",
      x = expression(-log[10](p.adjust)),
      y = NULL
    ) +
    theme_minimal(base_size = 12) +
    theme(
      panel.grid.major.y = element_blank(),
      plot.title = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(color = "gray40")
    )
  
  print(p_go_vert)
  ggsave(file.path(output_base, "plots/MSG_GO_barplot_vertebrates.svg"), p_go_vert, width = 10, height = 8)
  ggsave(file.path(output_base, "plots/MSG_GO_barplot_vertebrates.png"), p_go_vert, width = 10, height = 8, dpi = 300)
}
```

## GO Term Overlap (UpSet Plot)

```{r go-upset}
#| fig-width: 10
#| fig-height: 7

# Prepare list for UpSet
go_terms_list <- list(
  "Unicellular" = if (!is.null(go_unicellular)) go_unicellular$ID else character(0),
  "Multicellular" = if (!is.null(go_multicellular)) go_multicellular$ID else character(0),
  "Vertebrates" = if (!is.null(go_vertebrates)) go_vertebrates$ID else character(0)
)
go_terms_list <- go_terms_list[sapply(go_terms_list, length) > 0]

if (length(go_terms_list) >= 2) {
  svglite(file.path(output_base, "plots/MSG_GO_upset_plot.svg"), width = 10, height = 7)
  upset(
    fromList(go_terms_list),
    order.by = "freq",
    decreasing = TRUE,
    mb.ratio = c(0.6, 0.4),
    text.scale = 1.5,
    point.size = 3.5,
    line.size = 1.5,
    mainbar.y.label = "GO Terms Intersection",
    sets.x.label = "GO Terms per Group",
    sets.bar.color = c(
      group_colors["Unicellular"],
      group_colors["Multicellular non-vertebrate"],
      group_colors["Vertebrates"]
    )[names(go_terms_list)]
  )
  dev.off()
  
  # Also display inline
  upset(fromList(go_terms_list), order.by = "freq", sets.bar.color = "gray30")
}
```

## GO Semantic Similarity Clustering

Calculate semantic similarity between enriched GO terms to cluster them into functionally related groups.

```{r go-semantic}
#| cache: true

cat("--- Calculating GO Semantic Similarity ---\n")
go_terms <- unique(go_combined$ID)
cat(sprintf("Calculating similarity for %d unique GO terms...\n", length(go_terms)))

# Create semantic data object
semData <- godata(OrgDb = org.Hs.eg.db, ont = "BP")

# Calculate pairwise semantic similarity
mSim <- mgoSim(
  go_terms,
  go_terms,
  semData = semData,
  measure = "Wang",
  combine = NULL
)

# Handle NA values and convert to distance
mSim[is.na(mSim)] <- 0
dist_matrix <- as.dist(1 - mSim)
hc <- hclust(dist_matrix, method = "average")
```

## TreeAndLeaf Visualization Logic

Prepare the `TreeAndLeaf` object for visualization in RedeR. This structure maps GO terms onto the dendrogram leaves, scaling nodes by gene count and coloring by significance.

```{r treeandleaf-viz}
#| cache: true

# Pre-process GO data for mapping
go_combined <- go_combined %>%
  mutate(
    path_length = as.integer(sapply(strsplit(GeneRatio, "/"), "[", 2)),
    ratio = Count / path_length
  )

# Select best p-value per term (if term appears in multiple groups)
go_term_info <- go_combined %>%
  group_by(ID) %>%
  slice_min(pvalue, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  dplyr::select(ID, Description, Count, ratio, pvalue, p.adjust, evolutionary_group)

# Organize node sizes
size <- go_term_info$Count
aux <- sort(unique(size))
names(aux) <- as.character(1:length(aux))
sizeRanks <- as.factor(names(aux[match(size, aux)]))
sizeIntervals <- min(7, length(unique(size)))
go_term_info$size <- (sizeBase + (as.numeric(sizeRanks) * sizeMultiplier))
go_term_info$group_color <- group_colors[go_term_info$evolutionary_group]

# Create TreeAndLeaf Object
tal <- treeAndLeaf(hc)
tal <- att.mapv(g = tal, dat = as.data.frame(go_term_info), refcol = 1)

# Set visual attributes
pal <- brewer.pal(9, "OrRd")
tal <- att.setv(g = tal, from = "Description", to = "nodeAlias")
tal <- att.setv(g = tal, from = "ratio", to = "nodeColor", cols = pal, nquant = 5)
tal <- att.setv(g = tal, from = "size", to = "nodeSize",
                xlim = c((sizeBase + sizeMultiplier), (sizeBase + (sizeMultiplier * sizeIntervals)), sizeMultiplier),
                nquant = sizeIntervals)
tal <- att.addv(tal, "nodeFontSize", value = 15, index = V(tal)$isLeaf)
tal <- att.adde(tal, "edgeWidth", value = 3)
tal <- att.adde(tal, "edgeColor", value = "gray80")

# Set border color based on evolutionary group
leaf_indices <- which(V(tal)$isLeaf)
for (i in leaf_indices) {
  node_name <- V(tal)$name[i]
  node_group <- go_term_info %>% filter(ID == node_name) %>% pull(evolutionary_group)
  if (length(node_group) > 0) {
    V(tal)$nodeLineColor[i] <- group_colors[node_group]
  } else {
    V(tal)$nodeLineColor[i] <- "gray50"
  }
}
tal <- att.addv(tal, "nodeLineWidth", value = 4, index = V(tal)$isLeaf)

# Interactive RedeR Visualization (Run Manually)
# Note: Requires Java GUI. Set to eval=FALSE for non-interactive rendering.
# rdp <- RedPort()
# calld(rdp)
# resetd(rdp)
# addGraph(rdp, tal, layout = NULL)
# relax(rdp, p1 = 21, p2 = 70, p3 = 255, p4 = 100, p5 = 50)
```

## KEGG Pathway Enrichment Analysis

Perform KEGG pathway enrichment, excluding broad "Human Diseases" categories.

```{r kegg-enrichment}
#| cache: true

cat("--- Performing KEGG Enrichment Analysis ---\n")
kegg_unicellular <- perform_kegg_enrichment(genes_unicellular, "Unicellular")
kegg_multicellular <- perform_kegg_enrichment(genes_multicellular, "Multicellular non-vertebrate")
kegg_vertebrates <- perform_kegg_enrichment(genes_vertebrates, "Vertebrates")

# Filter Human Diseases
kegg_unicellular <- filter_human_diseases(kegg_unicellular)
kegg_multicellular <- filter_human_diseases(kegg_multicellular)
kegg_vertebrates <- filter_human_diseases(kegg_vertebrates)

# Combine results
kegg_combined <- bind_rows(
  kegg_unicellular,
  kegg_multicellular,
  kegg_vertebrates
) %>%
  filter(!is.na(ID))

# Save results
vroom::vroom_write(kegg_combined, file.path(output_base, "tables/MSG_KEGG_enrichment_combined.csv"), delim = ",")
if (!is.null(kegg_unicellular)) vroom::vroom_write(kegg_unicellular, file.path(output_base, "tables/MSG_KEGG_enrichment_unicellular.csv"), delim = ",")
if (!is.null(kegg_multicellular)) vroom::vroom_write(kegg_multicellular, file.path(output_base, "tables/MSG_KEGG_enrichment_multicellular.csv"), delim = ",")
if (!is.null(kegg_vertebrates)) vroom::vroom_write(kegg_vertebrates, file.path(output_base, "tables/MSG_KEGG_enrichment_vertebrates.csv"), delim = ",")
```

## KEGG Enrichment Visualization

```{r kegg-plots}
#| fig-width: 12
#| fig-height: 10

if (nrow(kegg_combined) > 0) {
  kegg_top <- kegg_combined %>%
    group_by(evolutionary_group) %>%
    slice_min(p.adjust, n = 15, with_ties = FALSE) %>%
    ungroup() %>%
    mutate(
      neg_log10_padj = -log10(p.adjust),
      evolutionary_group = factor(
        evolutionary_group,
        levels = c("Vertebrates", "Multicellular non-vertebrate", "Unicellular")
      )
    ) %>%
    arrange(evolutionary_group, neg_log10_padj) %>%
    mutate(Description = factor(Description, levels = unique(Description)))

  p_kegg_all <- ggplot(kegg_top, aes(x = neg_log10_padj, y = Description, fill = evolutionary_group)) +
    geom_bar(stat = "identity", width = 0.7) +
    facet_grid(evolutionary_group ~ ., scales = "free_y", space = "free_y") +
    scale_fill_manual(values = group_colors, name = "Evolutionary Group") +
    labs(
      title = "KEGG Pathway Enrichment — Metastasis Suppressor Genes",
      subtitle = "Top 15 pathways per evolutionary group (Human Diseases excluded)",
      x = expression(-log[10](p.adjust)),
      y = NULL
    ) +
    theme_minimal(base_size = 12) +
    theme(
      strip.text.y = element_text(angle = 0, face = "bold", size = 11),
      legend.position = "none",
      panel.grid.major.y = element_blank(),
      plot.title = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(color = "gray40")
    )

  print(p_kegg_all)
  ggsave(file.path(output_base, "plots/MSG_KEGG_barplot_all_groups.svg"), p_kegg_all, width = 12, height = 10)
  ggsave(file.path(output_base, "plots/MSG_KEGG_barplot_all_groups.png"), p_kegg_all, width = 12, height = 10, dpi = 300)
}
```

## KEGG Pathway Overlap (UpSet Plot)

```{r kegg-upset}
#| fig-width: 10
#| fig-height: 7

# Prepare list for UpSet
kegg_terms_list <- list(
  "Unicellular" = if (!is.null(kegg_unicellular)) kegg_unicellular$ID else character(0),
  "Multicellular" = if (!is.null(kegg_multicellular)) kegg_multicellular$ID else character(0),
  "Vertebrates" = if (!is.null(kegg_vertebrates)) kegg_vertebrates$ID else character(0)
)
kegg_terms_list <- kegg_terms_list[sapply(kegg_terms_list, length) > 0]

if (length(kegg_terms_list) >= 2) {
  svglite(file.path(output_base, "plots/MSG_KEGG_upset_plot.svg"), width = 10, height = 7)
  upset(
    fromList(kegg_terms_list),
    order.by = "freq",
    decreasing = TRUE,
    mb.ratio = c(0.6, 0.4),
    text.scale = 1.5,
    point.size = 3.5,
    line.size = 1.5,
    mainbar.y.label = "KEGG Pathways Intersection",
    sets.x.label = "Pathways per Group",
    sets.bar.color = c(
      group_colors["Unicellular"],
      group_colors["Multicellular non-vertebrate"],
      group_colors["Vertebrates"]
    )[names(kegg_terms_list)]
  )
  dev.off()
  
  # Also display inline
  upset(fromList(kegg_terms_list), order.by = "freq", sets.bar.color = "gray30")
}
```

## Save Outputs and Cleanup

```{r save-outputs}
#| cache: false

# Export consolidated nodelist with group assignments
msg_nodelist %>%
  vroom::vroom_write(file.path(output_base, "tables/MSG_nodelist_grouped.csv"), delim = ",")

# Temporary equivalence check example (run if originals exist)
if (file.exists(here("results/paper_data/MSG_GO_enrichment_combined.csv"))) {
  old_go <- vroom::vroom(here("results/paper_data/MSG_GO_enrichment_combined.csv"), show_col_types = FALSE)
  new_go <- vroom::vroom(file.path(output_base, "tables/MSG_GO_enrichment_combined.csv"), show_col_types = FALSE)
  cat("MSG_GO_enrichment_combined — old rows:", nrow(old_go), ", new rows:", nrow(new_go), "\n")
}

# Memory cleanup
rm(mSim, dist_matrix, hc, tal, go_combined, kegg_combined, nodelist)
gc()
```
