---
title: "MSG Enrichment Analysis with TreeAndLeaf"
author: "BioME"
format: html
editor: source
---

# Metastasis Suppressor Genes (MSG) Enrichment Analysis

This notebook performs GO and KEGG enrichment of metastasis suppressor genes across
3 evolutionary groups, clusters by semantic similarity, and visualizes using
TreeAndLeaf/RedeR. Horizontal bar plots are generated for both GO and KEGG results.

## Loading Libraries

```{r}
#| warning: false
#| message: false
library(TreeAndLeaf)
library(RedeR)
library(igraph)
library(tidyverse)
library(here)
library(vroom)
library(RColorBrewer)
library(clusterProfiler)
library(org.Hs.eg.db)
library(GOSemSim)
library(classInt)
library(UpSetR)
```

## Configuration

```{r}
# Define Evolutionary Group Colors
group_colors <- c(
  "Unicellular" = "#E69F00",                # Orange
  "Multicellular non-vertebrate" = "#0072B2", # Blue
  "Vertebrates" = "#009E73"                  # Green
)

# Visualization sizing parameters
sizeMultiplier <- 15
sizeBase <- 50
```

## Read Data

```{r}
#| message: false
# Nodelist with rooting information and biological process annotations
nodelist <- vroom::vroom(here("results/orthology_data/nodelist.csv"))

# Metastasis suppressor genes reference list
msg_reference <- vroom::vroom(
  here("assets/metastasis_supressors_nature.csv"),
  delim = "\t"
)
```

## Filter and Classify Metastasis Suppressor Genes

```{r}
# Filter nodelist to keep only metastasis suppressor genes
msg_nodelist <- nodelist %>%
  filter(`metastasis suppressor` == 1) %>%
  dplyr::select(cog_id, root, clade_name, string_id, queryItem)

cat("Total metastasis suppressor genes in nodelist:", nrow(msg_nodelist), "\n")

# Assign Evolutionary Groups Based on Root Values
# Group 1: Unicellular - before Choanoflagellata (exclusive), root > 30
# Group 2: Multicellular non-vertebrate - Choanoflagellata to Tunicata, 21 <= root <= 30
# Group 3: Vertebrates - from Actinopterygii (inclusive), root <= 20
msg_nodelist <- msg_nodelist %>%
  mutate(
    evolutionary_group = case_when(
      root > 30 ~ "Unicellular",
      root >= 21 & root <= 30 ~ "Multicellular non-vertebrate",
      root <= 20 ~ "Vertebrates"
    )
  ) %>%
  mutate(
    group_color = group_colors[evolutionary_group]
  )

# Summary of genes per group
cat("\n--- Gene Distribution by Evolutionary Group ---\n")
print(table(msg_nodelist$evolutionary_group))

# Get unique genes per group
genes_unicellular <- msg_nodelist %>%
  filter(evolutionary_group == "Unicellular") %>%
  pull(queryItem) %>%
  unique()

genes_multicellular <- msg_nodelist %>%
  filter(evolutionary_group == "Multicellular non-vertebrate") %>%
  pull(queryItem) %>%
  unique()

genes_vertebrates <- msg_nodelist %>%
  filter(evolutionary_group == "Vertebrates") %>%
  pull(queryItem) %>%
  unique()
```

# GO Enrichment Analysis

## GO Enrichment per Evolutionary Group

```{r}
#| warning: false
#| message: false

# Function to perform GO enrichment for a group
perform_go_enrichment <- function(genes, group_name) {
  cat(sprintf("  Processing %s (%d genes)...\n", group_name, length(genes)))

  # Convert gene symbols to Entrez IDs
  gene_ids <- bitr(
    genes,
    fromType = "SYMBOL",
    toType = "ENTREZID",
    OrgDb = org.Hs.eg.db
  )

  if (nrow(gene_ids) == 0) {
    cat(sprintf("    No genes mapped for %s\n", group_name))
    return(NULL)
  }

  # Perform GO enrichment
  ego <- enrichGO(
    gene = gene_ids$ENTREZID,
    OrgDb = org.Hs.eg.db,
    ont = "BP",
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05,
    qvalueCutoff = 0.05,
    readable = TRUE
  )

  if (is.null(ego) || nrow(as.data.frame(ego)) == 0) {
    cat(sprintf("    No enriched terms for %s\n", group_name))
    return(NULL)
  }

  result <- as.data.frame(ego) %>%
    mutate(evolutionary_group = group_name)

  cat(sprintf("    Found %d enriched GO terms\n", nrow(result)))
  return(result)
}

# Perform enrichment for each group
go_unicellular <- perform_go_enrichment(genes_unicellular, "Unicellular")
go_multicellular <- perform_go_enrichment(
  genes_multicellular,
  "Multicellular non-vertebrate"
)
go_vertebrates <- perform_go_enrichment(genes_vertebrates, "Vertebrates")

# Combine all results
go_combined <- bind_rows(
  go_unicellular,
  go_multicellular,
  go_vertebrates
) %>%
  filter(!is.na(ID))

if (nrow(go_combined) == 0) {
  stop("No enriched GO terms found in any group.")
}

cat(sprintf(
  "\nTotal enriched GO terms across all groups: %d\n",
  nrow(go_combined)
))
```

## Save GO Enrichment Results

```{r}
#| message: false
if (!is.null(go_unicellular) && nrow(go_unicellular) > 0) {
  vroom::vroom_write(
    go_unicellular,
    here("results/paper_data/MSG_GO_enrichment_unicellular.csv"),
    delim = ","
  )
}

if (!is.null(go_multicellular) && nrow(go_multicellular) > 0) {
  vroom::vroom_write(
    go_multicellular,
    here("results/paper_data/MSG_GO_enrichment_multicellular.csv"),
    delim = ","
  )
}

if (!is.null(go_vertebrates) && nrow(go_vertebrates) > 0) {
  vroom::vroom_write(
    go_vertebrates,
    here("results/paper_data/MSG_GO_enrichment_vertebrates.csv"),
    delim = ","
  )
}

vroom::vroom_write(
  go_combined,
  here("results/paper_data/MSG_GO_enrichment_combined.csv"),
  delim = ","
)

cat("Enrichment results saved to results/paper_data/\n")
```

## GO Enrichment Bar Plot — All Groups

```{r}
#| fig-width: 12
#| fig-height: 14
#| warning: false

# Select top 15 GO terms per group by p.adjust
go_top <- go_combined %>%
  group_by(evolutionary_group) %>%
  slice_min(p.adjust, n = 15, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    neg_log10_padj = -log10(p.adjust),
    # Set factor order: Unicellular on top, Multicellular middle, Vertebrates bottom
    evolutionary_group = factor(
      evolutionary_group,
      levels = c("Vertebrates", "Multicellular non-vertebrate", "Unicellular")
    )
  )

# Order descriptions within each group
go_top <- go_top %>%
  arrange(evolutionary_group, neg_log10_padj) %>%
  mutate(Description = factor(Description, levels = unique(Description)))

p_go_all <- ggplot(go_top, aes(x = neg_log10_padj, y = Description, fill = evolutionary_group)) +
  geom_bar(stat = "identity", width = 0.7) +
  facet_grid(evolutionary_group ~ ., scales = "free_y", space = "free_y") +
  scale_fill_manual(values = group_colors, name = "Evolutionary Group") +
  labs(
    title = "GO Biological Process Enrichment — Metastasis Suppressor Genes",
    subtitle = "Top 15 terms per evolutionary group",
    x = expression(-log[10](p.adjust)),
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text.y = element_text(angle = 0, face = "bold", size = 11),
    legend.position = "none",
    panel.grid.major.y = element_blank(),
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(color = "gray40")
  )

print(p_go_all)

# Save
ggsave(
  here("results/plots/MSG_GO_barplot_all_groups.pdf"),
  p_go_all, width = 12, height = 14
)
ggsave(
  here("results/plots/MSG_GO_barplot_all_groups.png"),
  p_go_all, width = 12, height = 14, dpi = 300
)
```

## GO Enrichment Bar Plot — Vertebrates Only

```{r}
#| fig-width: 10
#| fig-height: 8
#| warning: false

go_vert <- go_combined %>%
  filter(evolutionary_group == "Vertebrates") %>%
  slice_min(p.adjust, n = 20, with_ties = FALSE) %>%
  mutate(
    neg_log10_padj = -log10(p.adjust)
  ) %>%
  arrange(neg_log10_padj) %>%
  mutate(Description = factor(Description, levels = unique(Description)))

p_go_vert <- ggplot(go_vert, aes(x = neg_log10_padj, y = Description)) +
  geom_bar(stat = "identity", fill = group_colors["Vertebrates"], width = 0.7) +
  labs(
    title = "GO Biological Process Enrichment — Vertebrates MSGs",
    subtitle = "Top 20 enriched terms",
    x = expression(-log[10](p.adjust)),
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.y = element_blank(),
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(color = "gray40")
  )

print(p_go_vert)

ggsave(
  here("results/plots/MSG_GO_barplot_vertebrates.pdf"),
  p_go_vert, width = 10, height = 8
)
ggsave(
  here("results/plots/MSG_GO_barplot_vertebrates.png"),
  p_go_vert, width = 10, height = 8, dpi = 300
)
```

## UpSet Plot — GO Term Overlap Between Groups

```{r}
#| fig-width: 10
#| fig-height: 7
#| warning: false

# Create list of GO terms per group
go_terms_list <- list(
  "Unicellular" = if (!is.null(go_unicellular) && nrow(go_unicellular) > 0) {
    go_unicellular$ID
  } else {
    character(0)
  },
  "Multicellular" = if (
    !is.null(go_multicellular) && nrow(go_multicellular) > 0
  ) {
    go_multicellular$ID
  } else {
    character(0)
  },
  "Vertebrates" = if (!is.null(go_vertebrates) && nrow(go_vertebrates) > 0) {
    go_vertebrates$ID
  } else {
    character(0)
  }
)

# Remove empty groups
go_terms_list <- go_terms_list[sapply(go_terms_list, length) > 0]

if (length(go_terms_list) >= 2) {
  upset_plot <- upset(
    fromList(go_terms_list),
    order.by = "freq",
    decreasing = TRUE,
    mb.ratio = c(0.6, 0.4),
    number.angles = 0,
    text.scale = 1.5,
    point.size = 3.5,
    line.size = 1.5,
    mainbar.y.label = "GO Terms Intersection",
    sets.x.label = "GO Terms per Group",
    sets.bar.color = c(
      group_colors["Unicellular"],
      group_colors["Multicellular non-vertebrate"],
      group_colors["Vertebrates"]
    )[names(go_terms_list)]
  )

  print(upset_plot)

  # Save UpSet plot
  pdf(here("results/plots/MSG_GO_upset_plot.pdf"), width = 10, height = 7)
  print(upset_plot)
  dev.off()

  png(
    here("results/plots/MSG_GO_upset_plot.png"),
    width = 1000, height = 700, res = 100
  )
  print(upset_plot)
  dev.off()

  cat("UpSet plot saved to results/plots/MSG_GO_upset_plot.pdf and .png\n")
}
```

## GO Term Overlap Statistics

```{r}
if (length(go_terms_list) >= 2) {
  # Find terms in multiple groups
  all_terms <- unique(go_combined$ID)
  terms_in_groups <- sapply(all_terms, function(term) {
    groups <- go_combined %>%
      filter(ID == term) %>%
      pull(evolutionary_group) %>%
      unique()
    paste(groups, collapse = " & ")
  })

  overlap_summary <- table(terms_in_groups)
  cat("Distribution of GO terms across groups:\n")
  print(overlap_summary)

  # List terms that appear in multiple groups
  shared_terms <- go_combined %>%
    group_by(ID, Description) %>%
    summarise(
      groups = paste(unique(evolutionary_group), collapse = " & "),
      n_groups = n_distinct(evolutionary_group),
      .groups = "drop"
    ) %>%
    filter(n_groups > 1) %>%
    arrange(desc(n_groups))

  if (nrow(shared_terms) > 0) {
    cat(sprintf(
      "\nGO terms appearing in multiple groups: %d\n",
      nrow(shared_terms)
    ))
    print(as.data.frame(shared_terms), row.names = FALSE)

    vroom::vroom_write(
      shared_terms,
      here("results/paper_data/MSG_GO_shared_terms.csv"),
      delim = ","
    )
  } else {
    cat("\nNo GO terms appear in multiple groups.\n")
  }
}
```

## GO Semantic Similarity Clustering

```{r}
#| message: false
cat("--- Calculating GO Semantic Similarity ---\n")

# Get unique GO term IDs
go_terms <- unique(go_combined$ID)
cat(sprintf(
  "Calculating similarity for %d unique GO terms...\n",
  length(go_terms)
))

# Create semantic data object
semData <- godata(OrgDb = org.Hs.eg.db, ont = "BP")

# Calculate pairwise semantic similarity
mSim <- mgoSim(
  go_terms,
  go_terms,
  semData = semData,
  measure = "Wang",
  combine = NULL
)

# Handle NA values by replacing with 0
mSim[is.na(mSim)] <- 0

# Create distance matrix (convert similarity to distance)
dist_matrix <- as.dist(1 - mSim)

# Hierarchical clustering
hc <- hclust(dist_matrix, method = "average")

cat("Semantic similarity clustering complete.\n")
```

## GO TreeAndLeaf Visualization

```{r}
#| message: false
# Prepare Data for Visualization
go_combined <- go_combined %>%
  mutate(
    path_length = as.integer(sapply(strsplit(GeneRatio, "/"), "[", 2)),
    ratio = Count / path_length
  )

# For each GO term, determine the primary group
go_term_info <- go_combined %>%
  group_by(ID) %>%
  slice_min(pvalue, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  dplyr::select(
    ID, Description, Count, ratio, pvalue, p.adjust, evolutionary_group
  )

# Organize node sizes
size <- go_term_info$Count
aux <- sort(unique(size))
names(aux) <- as.character(1:length(aux))
sizeRanks <- as.factor(names(aux[match(size, aux)]))
sizeIntervals <- min(7, length(unique(size)))
go_term_info$size <- (sizeBase + (as.numeric(sizeRanks) * sizeMultiplier))

# Add group color
go_term_info$group_color <- group_colors[go_term_info$evolutionary_group]

# Create TreeAndLeaf Object
cat("--- Creating TreeAndLeaf Visualization ---\n")

tal <- treeAndLeaf(hc)

# Map attributes to the graph
tal <- att.mapv(g = tal, dat = as.data.frame(go_term_info), refcol = 1)

# Color palette for ratio
pal <- brewer.pal(9, "OrRd")

# Set node labels (GO term descriptions)
tal <- att.setv(g = tal, from = "Description", to = "nodeAlias")

# Set node colors based on ratio
tal <- att.setv(
  g = tal,
  from = "ratio",
  to = "nodeColor",
  cols = pal,
  nquant = 5
)

# Set node sizes based on gene count
tal <- att.setv(
  g = tal,
  from = "size",
  to = "nodeSize",
  xlim = c(
    (sizeBase + sizeMultiplier),
    (sizeBase + (sizeMultiplier * sizeIntervals)),
    sizeMultiplier
  ),
  nquant = sizeIntervals
)

# Set font size for labels
tal <- att.addv(tal, "nodeFontSize", value = 15, index = V(tal)$isLeaf)

# Set edge attributes
tal <- att.adde(tal, "edgeWidth", value = 3)
tal <- att.adde(tal, "edgeColor", value = "gray80")

# Set nodeLineColor Based on Evolutionary Group
leaf_indices <- which(V(tal)$isLeaf)

for (i in leaf_indices) {
  node_name <- V(tal)$name[i]
  node_group <- go_term_info %>%
    filter(ID == node_name) %>%
    pull(evolutionary_group)

  if (length(node_group) > 0 && !is.na(node_group)) {
    V(tal)$nodeLineColor[i] <- group_colors[node_group]
  } else {
    V(tal)$nodeLineColor[i] <- "gray50"
  }
}

# Set line width for the border
tal <- att.addv(tal, "nodeLineWidth", value = 4, index = V(tal)$isLeaf)

# For non-leaf nodes, set gray color
non_leaf_indices <- which(!V(tal)$isLeaf)
V(tal)$nodeLineColor[non_leaf_indices] <- "gray80"
V(tal)$nodeLineWidth[non_leaf_indices] <- 1
```

## Display GO TreeAndLeaf in RedeR

```{r}
#| eval: false
cat("--- Starting RedeR Visualization ---\n")
cat("Group Colors (nodeLineColor):\n")
for (group in names(group_colors)) {
  cat(sprintf("  %s: %s\n", group, group_colors[group]))
}

# Start RedeR
rdp <- RedPort()
calld(rdp)

# Reset and add graph
resetd(rdp)
addGraph(rdp, tal, layout = NULL)

# Add legends
addLegend.color(
  obj = rdp,
  tal,
  title = "Gene Ratio",
  position = "topright"
)

addLegend.size(
  obj = rdp,
  tal,
  title = "Gene Count",
  position = "bottomright"
)

# Relax the network layout
relax(rdp, p1 = 21, p2 = 70, p3 = 255, p4 = 100, p5 = 50)

cat("GO TreeAndLeaf visualization displayed in RedeR.\n")
```

## GO Summary Statistics

```{r}
cat("--- GO Analysis Summary ---\n")
cat(sprintf("Total MSG genes analyzed: %d\n", nrow(msg_nodelist)))
cat(sprintf(
  "  Unicellular: %d genes\n",
  sum(msg_nodelist$evolutionary_group == "Unicellular")
))
cat(sprintf(
  "  Multicellular non-vertebrate: %d genes\n",
  sum(msg_nodelist$evolutionary_group == "Multicellular non-vertebrate")
))
cat(sprintf(
  "  Vertebrates: %d genes\n",
  sum(msg_nodelist$evolutionary_group == "Vertebrates")
))

cat(sprintf("\nTotal enriched GO terms: %d\n", nrow(go_combined)))
cat("GO terms per group:\n")
print(table(go_term_info$evolutionary_group))

# Export MSG nodelist with group assignment
msg_nodelist %>%
  vroom::vroom_write(
    file = here("results/orthology_data/MSG_nodelist_grouped.csv"),
    delim = ","
  )
```

# KEGG Pathway Enrichment Analysis

## KEGG Enrichment per Evolutionary Group

```{r}
#| warning: false
#| message: false

cat("--- Performing KEGG Enrichment Analysis ---\n")

# Function to perform KEGG enrichment for a group
perform_kegg_enrichment <- function(genes, group_name) {
  cat(sprintf("  Processing %s (%d genes)...\n", group_name, length(genes)))

  # Convert gene symbols to Entrez IDs
  gene_ids <- bitr(
    genes,
    fromType = "SYMBOL",
    toType = "ENTREZID",
    OrgDb = org.Hs.eg.db
  )

  if (nrow(gene_ids) == 0) {
    cat(sprintf("    No genes mapped for %s\n", group_name))
    return(NULL)
  }

  # Perform KEGG enrichment
  ekegg <- enrichKEGG(
    gene = gene_ids$ENTREZID,
    organism = "hsa",
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05,
    qvalueCutoff = 0.05
  )

  if (is.null(ekegg) || nrow(as.data.frame(ekegg)) == 0) {
    cat(sprintf("    No enriched KEGG pathways for %s\n", group_name))
    return(NULL)
  }

  # Convert to readable gene symbols
  ekegg <- setReadable(ekegg, OrgDb = org.Hs.eg.db, keyType = "ENTREZID")

  result <- as.data.frame(ekegg) %>%
    mutate(evolutionary_group = group_name)

  cat(sprintf("    Found %d enriched KEGG pathways\n", nrow(result)))
  return(result)
}

# Perform KEGG enrichment for each group
kegg_unicellular <- perform_kegg_enrichment(genes_unicellular, "Unicellular")
kegg_multicellular <- perform_kegg_enrichment(
  genes_multicellular,
  "Multicellular non-vertebrate"
)
kegg_vertebrates <- perform_kegg_enrichment(genes_vertebrates, "Vertebrates")
```

## Filter KEGG Human Disease Pathways

KEGG pathways in the "Human Diseases" category have IDs starting with `hsa05`.
These are removed to focus on biological and signaling pathways.

```{r}
# Filter out Human Diseases (hsa05*) from each group
filter_human_diseases <- function(df) {
  if (is.null(df) || nrow(df) == 0) return(df)
  df_filtered <- df %>%
    filter(!str_detect(ID, "^hsa05"))
  cat(sprintf(
    "  Removed %d Human Disease pathways, %d remaining\n",
    nrow(df) - nrow(df_filtered), nrow(df_filtered)
  ))
  if (nrow(df_filtered) == 0) return(NULL)
  return(df_filtered)
}

cat("--- Filtering Human Disease Pathways ---\n")
kegg_unicellular <- filter_human_diseases(kegg_unicellular)
kegg_multicellular <- filter_human_diseases(kegg_multicellular)
kegg_vertebrates <- filter_human_diseases(kegg_vertebrates)

# Combine all KEGG results (filtered)
kegg_combined <- bind_rows(
  kegg_unicellular,
  kegg_multicellular,
  kegg_vertebrates
) %>%
  filter(!is.na(ID))

cat(sprintf(
  "\nTotal enriched KEGG pathways after filtering: %d\n",
  nrow(kegg_combined)
))
```

## Save KEGG Enrichment Results

```{r}
#| message: false
if (nrow(kegg_combined) > 0) {
  if (!is.null(kegg_unicellular) && nrow(kegg_unicellular) > 0) {
    vroom::vroom_write(
      kegg_unicellular,
      here("results/paper_data/MSG_KEGG_enrichment_unicellular.csv"),
      delim = ","
    )
  }

  if (!is.null(kegg_multicellular) && nrow(kegg_multicellular) > 0) {
    vroom::vroom_write(
      kegg_multicellular,
      here("results/paper_data/MSG_KEGG_enrichment_multicellular.csv"),
      delim = ","
    )
  }

  if (!is.null(kegg_vertebrates) && nrow(kegg_vertebrates) > 0) {
    vroom::vroom_write(
      kegg_vertebrates,
      here("results/paper_data/MSG_KEGG_enrichment_vertebrates.csv"),
      delim = ","
    )
  }

  vroom::vroom_write(
    kegg_combined,
    here("results/paper_data/MSG_KEGG_enrichment_combined.csv"),
    delim = ","
  )

  cat("KEGG enrichment results saved to results/paper_data/\n")
}
```

## KEGG Enrichment Bar Plot — All Groups

```{r}
#| fig-width: 12
#| fig-height: 10
#| warning: false

if (nrow(kegg_combined) > 0) {
  kegg_top <- kegg_combined %>%
    group_by(evolutionary_group) %>%
    slice_min(p.adjust, n = 15, with_ties = FALSE) %>%
    ungroup() %>%
    mutate(
      neg_log10_padj = -log10(p.adjust),
      evolutionary_group = factor(
        evolutionary_group,
        levels = c("Vertebrates", "Multicellular non-vertebrate", "Unicellular")
      )
    )

  kegg_top <- kegg_top %>%
    arrange(evolutionary_group, neg_log10_padj) %>%
    mutate(Description = factor(Description, levels = unique(Description)))

  p_kegg_all <- ggplot(
    kegg_top,
    aes(x = neg_log10_padj, y = Description, fill = evolutionary_group)
  ) +
    geom_bar(stat = "identity", width = 0.7) +
    facet_grid(evolutionary_group ~ ., scales = "free_y", space = "free_y") +
    scale_fill_manual(values = group_colors, name = "Evolutionary Group") +
    labs(
      title = "KEGG Pathway Enrichment — Metastasis Suppressor Genes",
      subtitle = "Top 15 pathways per evolutionary group (Human Diseases excluded)",
      x = expression(-log[10](p.adjust)),
      y = NULL
    ) +
    theme_minimal(base_size = 12) +
    theme(
      strip.text.y = element_text(angle = 0, face = "bold", size = 11),
      legend.position = "none",
      panel.grid.major.y = element_blank(),
      plot.title = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(color = "gray40")
    )

  print(p_kegg_all)

  ggsave(
    here("results/plots/MSG_KEGG_barplot_all_groups.pdf"),
    p_kegg_all, width = 12, height = 10
  )
  ggsave(
    here("results/plots/MSG_KEGG_barplot_all_groups.png"),
    p_kegg_all, width = 12, height = 10, dpi = 300
  )
}
```

## KEGG Enrichment Bar Plot — Vertebrates Only

```{r}
#| fig-width: 10
#| fig-height: 8
#| warning: false

if (!is.null(kegg_vertebrates) && nrow(kegg_vertebrates) > 0) {
  kegg_vert <- kegg_vertebrates %>%
    slice_min(p.adjust, n = 20, with_ties = FALSE) %>%
    mutate(
      neg_log10_padj = -log10(p.adjust)
    ) %>%
    arrange(neg_log10_padj) %>%
    mutate(Description = factor(Description, levels = unique(Description)))

  p_kegg_vert <- ggplot(
    kegg_vert,
    aes(x = neg_log10_padj, y = Description)
  ) +
    geom_bar(stat = "identity", fill = group_colors["Vertebrates"], width = 0.7) +
    labs(
      title = "KEGG Pathway Enrichment — Vertebrates MSGs",
      subtitle = "Top 20 enriched pathways (Human Diseases excluded)",
      x = expression(-log[10](p.adjust)),
      y = NULL
    ) +
    theme_minimal(base_size = 12) +
    theme(
      panel.grid.major.y = element_blank(),
      plot.title = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(color = "gray40")
    )

  print(p_kegg_vert)

  ggsave(
    here("results/plots/MSG_KEGG_barplot_vertebrates.pdf"),
    p_kegg_vert, width = 10, height = 8
  )
  ggsave(
    here("results/plots/MSG_KEGG_barplot_vertebrates.png"),
    p_kegg_vert, width = 10, height = 8, dpi = 300
  )
}
```

## KEGG UpSet Plot

```{r}
#| fig-width: 10
#| fig-height: 7
#| warning: false

if (nrow(kegg_combined) > 0) {
  kegg_terms_list <- list(
    "Unicellular" = if (!is.null(kegg_unicellular) && nrow(kegg_unicellular) > 0)
      kegg_unicellular$ID else character(0),
    "Multicellular" = if (!is.null(kegg_multicellular) && nrow(kegg_multicellular) > 0)
      kegg_multicellular$ID else character(0),
    "Vertebrates" = if (!is.null(kegg_vertebrates) && nrow(kegg_vertebrates) > 0)
      kegg_vertebrates$ID else character(0)
  )

  kegg_terms_list <- kegg_terms_list[sapply(kegg_terms_list, length) > 0]

  if (length(kegg_terms_list) >= 2) {
    kegg_upset <- upset(
      fromList(kegg_terms_list),
      order.by = "freq",
      decreasing = TRUE,
      mb.ratio = c(0.6, 0.4),
      number.angles = 0,
      text.scale = 1.5,
      point.size = 3.5,
      line.size = 1.5,
      mainbar.y.label = "KEGG Pathways Intersection",
      sets.x.label = "Pathways per Group"
    )

    print(kegg_upset)

    pdf(here("results/plots/MSG_KEGG_upset_plot.pdf"), width = 10, height = 7)
    print(kegg_upset)
    dev.off()

    png(
      here("results/plots/MSG_KEGG_upset_plot.png"),
      width = 1000, height = 700, res = 100
    )
    print(kegg_upset)
    dev.off()

    cat("KEGG UpSet plot saved to results/plots/\n")
  }
}
```

<!-- ## KEGG Clustering and TreeAndLeaf

```{r}
#| message: false
if (nrow(kegg_combined) > 0) {
  cat("--- Creating KEGG TreeAndLeaf Visualization ---\n")

  # Get unique KEGG pathway IDs
  kegg_ids <- unique(kegg_combined$ID)
  cat(sprintf("Creating clustering for %d unique KEGG pathways...\n", length(kegg_ids)))

  # Calculate distance based on gene overlap (Jaccard similarity)
  kegg_gene_matrix <- matrix(0, nrow = length(kegg_ids), ncol = length(kegg_ids))
  rownames(kegg_gene_matrix) <- kegg_ids
  colnames(kegg_gene_matrix) <- kegg_ids

  for (i in seq_along(kegg_ids)) {
    for (j in seq_along(kegg_ids)) {
      if (i <= j) {
        genes_i <- unlist(strsplit(
          kegg_combined$geneID[kegg_combined$ID == kegg_ids[i]][1], "/"
        ))
        genes_j <- unlist(strsplit(
          kegg_combined$geneID[kegg_combined$ID == kegg_ids[j]][1], "/"
        ))

        intersection <- length(intersect(genes_i, genes_j))
        union_len <- length(unique(c(genes_i, genes_j)))

        if (union_len > 0) {
          jaccard <- intersection / union_len
        } else {
          jaccard <- 0
        }

        kegg_gene_matrix[i, j] <- jaccard
        kegg_gene_matrix[j, i] <- jaccard
      }
    }
  }

  # Convert similarity to distance
  kegg_dist <- as.dist(1 - kegg_gene_matrix)

  # Handle zero variation case
  if (all(kegg_dist == 0)) {
    kegg_dist <- kegg_dist + runif(length(kegg_dist), 0, 0.001)
  }

  # Hierarchical clustering
  hc_kegg <- hclust(kegg_dist, method = "average")

  # Create TreeAndLeaf object
  tal_kegg <- treeAndLeaf(hc_kegg)

  # Prepare KEGG data for visualization
  kegg_combined <- kegg_combined %>%
    mutate(
      path_length = as.integer(sapply(strsplit(GeneRatio, "/"), "[", 2)),
      ratio = Count / path_length
    )

  kegg_term_info <- kegg_combined %>%
    group_by(ID) %>%
    slice_min(pvalue, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    dplyr::select(ID, Description, Count, ratio, pvalue, p.adjust, evolutionary_group)

  # Organize node sizes
  kegg_size <- kegg_term_info$Count
  kegg_aux <- sort(unique(kegg_size))
  names(kegg_aux) <- as.character(1:length(kegg_aux))
  kegg_sizeRanks <- as.factor(names(kegg_aux[match(kegg_size, kegg_aux)]))
  kegg_sizeIntervals <- min(7, length(unique(kegg_size)))
  kegg_term_info$size <- (sizeBase + (as.numeric(kegg_sizeRanks) * sizeMultiplier))

  kegg_term_info$group_color <- group_colors[kegg_term_info$evolutionary_group]

  # Map attributes to KEGG graph
  tal_kegg <- att.mapv(g = tal_kegg, dat = as.data.frame(kegg_term_info), refcol = 1)

  # Set node labels
  tal_kegg <- att.setv(g = tal_kegg, from = "Description", to = "nodeAlias")

  # Set node colors based on ratio
  pal_kegg <- brewer.pal(9, "YlOrRd")
  tal_kegg <- att.setv(
    g = tal_kegg,
    from = "ratio",
    to = "nodeColor",
    cols = pal_kegg,
    nquant = 5
  )

  # Set node sizes
  tal_kegg <- att.setv(
    g = tal_kegg,
    from = "size",
    to = "nodeSize",
    xlim = c(
      (sizeBase + sizeMultiplier),
      (sizeBase + (sizeMultiplier * kegg_sizeIntervals)),
      sizeMultiplier
    ),
    nquant = kegg_sizeIntervals
  )

  # Set font size
  tal_kegg <- att.addv(tal_kegg, "nodeFontSize", value = 15, index = V(tal_kegg)$isLeaf)

  # Set edge attributes
  tal_kegg <- att.adde(tal_kegg, "edgeWidth", value = 3)
  tal_kegg <- att.adde(tal_kegg, "edgeColor", value = "gray80")

  # Set nodeLineColor based on evolutionary group
  kegg_leaf_indices <- which(V(tal_kegg)$isLeaf)

  for (i in kegg_leaf_indices) {
    node_name <- V(tal_kegg)$name[i]
    node_group <- kegg_term_info %>%
      filter(ID == node_name) %>%
      pull(evolutionary_group)

    if (length(node_group) > 0 && !is.na(node_group)) {
      V(tal_kegg)$nodeLineColor[i] <- group_colors[node_group]
    } else {
      V(tal_kegg)$nodeLineColor[i] <- "gray50"
    }
  }

  tal_kegg <- att.addv(tal_kegg, "nodeLineWidth", value = 4, index = V(tal_kegg)$isLeaf)

  kegg_non_leaf_indices <- which(!V(tal_kegg)$isLeaf)
  V(tal_kegg)$nodeLineColor[kegg_non_leaf_indices] <- "gray80"
  V(tal_kegg)$nodeLineWidth[kegg_non_leaf_indices] <- 1

  cat("KEGG TreeAndLeaf object created.\n")
}
```

## Display KEGG TreeAndLeaf in RedeR

```{r}
#| eval: false
if (nrow(kegg_combined) > 0) {
  cat("--- Starting KEGG RedeR Visualization ---\n")
  resetRedeR()

  addGraph(rdp, tal_kegg, layout = NULL)

  addLegend.color(
    obj = rdp,
    tal_kegg,
    title = "KEGG Gene Ratio",
    position = "topleft"
  )

  relax(rdp, p1 = 21, p2 = 70, p3 = 255, p4 = 100, p5 = 50)

  cat("KEGG TreeAndLeaf visualization added to RedeR.\n")
}
``` -->

## KEGG Summary Statistics

```{r}
if (nrow(kegg_combined) > 0) {
  cat("--- KEGG Analysis Summary ---\n")
  cat(sprintf("Total enriched KEGG pathways: %d\n", nrow(kegg_combined)))
  cat("KEGG pathways per group:\n")
  print(table(kegg_term_info$evolutionary_group))
} else {
  cat("\nNo KEGG pathways were enriched in any group.\n")
}

cat("\n=== Analysis Complete ===\n")
```
