---
title: "Metastasis Suppressor Genes (MSG) Cumulative Enrichment Analysis"
author: "Gleison M. Azevedo"
format: html
execute:
  cache: true
  echo: true
  message: false
  warning: false
---

## Package Management

```{r pkg-install}
#| cache: false

# Automatically detect and install any missing packages before loading.
required_cran <- c(
  "tidyverse", "vroom", "here", "RColorBrewer", "classInt", "UpSetR", "svglite", "igraph", "scales", "ggtext"
)

required_bioc <- c(
  "TreeAndLeaf", "RedeR", "clusterProfiler", "org.Hs.eg.db", "GOSemSim"
)

# Install CRAN packages
missing_cran <- required_cran[
  !vapply(required_cran, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_cran) > 0) {
  install.packages(missing_cran, repos = "https://cloud.r-project.org")
}

# Install Bioconductor packages
missing_bioc <- required_bioc[
  !vapply(required_bioc, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_bioc) > 0) {
  if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager", repos = "https://cloud.r-project.org")
  BiocManager::install(missing_bioc, update = FALSE)
}

library(TreeAndLeaf)
library(RedeR)
library(igraph)
library(tidyverse)
library(here)
library(vroom)
library(RColorBrewer)
library(clusterProfiler)
library(org.Hs.eg.db)
library(GOSemSim)
library(classInt)
library(UpSetR)
library(svglite)
```

## Data Integrity Checksum

```{r checksum}
#| cache: false

# Verify MD5 checksum of critical input files.
expected_checksums <- list(
  "assets/metastasis_supressors_nature.csv" = "6f1717e97a277fb0fed475e92647d214"
)

for (file_path in names(expected_checksums)) {
  full_path <- here(file_path)
  if (file.exists(full_path)) {
    actual_checksum <- tools::md5sum(full_path)
    if (actual_checksum != expected_checksums[[file_path]]) {
      warning(paste("Checksum mismatch for", file_path, 
                    "\nExpected:", expected_checksums[[file_path]], 
                    "\nActual:  ", actual_checksum))
    } else {
      message(paste("Checksum verified for", file_path))
    }
  } else {
    warning(paste("File not found:", file_path))
  }
}
```

## Helper Functions and Configuration

```{r functions}
# Define Evolutionary Group Colors
group_colors <- c(
  "Unicellular" = "#E69F00",                # Orange
  "Multicellular non-vertebrate" = "#0072B2", # Blue
  "Vertebrates" = "#009E73"                  # Green
)

# Visualization sizing parameters
sizeMultiplier <- 15
sizeBase <- 50

# --- Function: GO Enrichment ---
perform_go_enrichment <- function(genes, group_name) {
  cat(sprintf("  Processing %s (%d genes)...\n", group_name, length(genes)))
  if (length(genes) == 0) return(NULL)

  gene_ids <- suppressMessages(bitr(genes, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db))
  if (nrow(gene_ids) == 0) {
    cat(sprintf("    No genes mapped for %s\n", group_name))
    return(NULL)
  }

  ego <- enrichGO(
    gene = gene_ids$ENTREZID,
    OrgDb = org.Hs.eg.db,
    ont = "BP",
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05,
    qvalueCutoff = 0.05,
    readable = TRUE
  )

  if (is.null(ego) || nrow(as.data.frame(ego)) == 0) return(NULL)

  result <- as.data.frame(ego) %>% mutate(evolutionary_group = group_name)
  cat(sprintf("    Found %d enriched GO terms\n", nrow(result)))
  return(result)
}

# --- Function: KEGG Enrichment ---
perform_kegg_enrichment <- function(genes, group_name) {
  cat(sprintf("  Processing %s (%d genes)...\n", group_name, length(genes)))
  if (length(genes) == 0) return(NULL)

  gene_ids <- suppressMessages(bitr(genes, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db))
  if (nrow(gene_ids) == 0) return(NULL)

  ekegg <- enrichKEGG(
    gene = gene_ids$ENTREZID,
    organism = "hsa",
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05,
    qvalueCutoff = 0.05
  )

  if (is.null(ekegg) || nrow(as.data.frame(ekegg)) == 0) return(NULL)

  ekegg <- setReadable(ekegg, OrgDb = org.Hs.eg.db, keyType = "ENTREZID")
  result <- as.data.frame(ekegg) %>% mutate(evolutionary_group = group_name)
  cat(sprintf("    Found %d enriched KEGG pathways\n", nrow(result)))
  return(result)
}

# --- Function: Filter Human Diseases ---
filter_human_diseases <- function(df) {
  if (is.null(df) || nrow(df) == 0) return(df)
  df_filtered <- df %>% filter(!str_detect(ID, "^hsa05"))
  cat(sprintf("  Removed %d Human Disease pathways, %d remaining\n", nrow(df) - nrow(df_filtered), nrow(df_filtered)))
  if (nrow(df_filtered) == 0) return(NULL)
  return(df_filtered)
}

# --- Function: Cumulative KEGG Clade Enrichment ---
perform_kegg_clade <- function(genes, clade_name) {
  gene_ids <- tryCatch(
    bitr(genes, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db),
    error = function(e) data.frame(SYMBOL = character(0), ENTREZID = character(0))
  )
  if (nrow(gene_ids) < 3) return(NULL)

  ekegg <- tryCatch(
    enrichKEGG(gene = gene_ids$ENTREZID, organism = "hsa", pAdjustMethod = "BH", pvalueCutoff = 0.05, qvalueCutoff = 0.05),
    error = function(e) NULL
  )

  if (is.null(ekegg) || nrow(as.data.frame(ekegg)) == 0) return(NULL)

  ekegg <- setReadable(ekegg, OrgDb = org.Hs.eg.db, keyType = "ENTREZID")
  result <- as.data.frame(ekegg) %>% filter(!str_detect(ID, "^hsa05")) %>% mutate(clade = clade_name)
  if (nrow(result) == 0) return(NULL)
  return(result)
}
```

## Data Loading

```{r data-load}
nodelist <- vroom::vroom(
  here("results/orthology_data/nodelist.csv"),
  show_col_types = FALSE
)

msg_reference <- vroom::vroom(
  here("assets/metastasis_supressors_nature.csv"),
  delim = "\t",
  show_col_types = FALSE
)

# Output Setup
output_base <- here("results/enrichment")
dir.create(file.path(output_base, "tables"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(output_base, "plots"), recursive = TRUE, showWarnings = FALSE)
```

## Filter and Classify Metastasis Suppressor Genes (Cumulative)

In the cumulative approach, each evolutionary group includes **all genes that existed up to that point**, not only those that first appeared (rooted) there.

```{r define-cumulative}
# Filter for MSG
msg_nodelist <- nodelist %>%
  filter(`metastasis suppressor` == 1) %>%
  dplyr::select(cog_id, root, clade_name, string_id, queryItem)

cat("Total metastasis suppressor genes in nodelist:", nrow(msg_nodelist), "\n")

# Assign root-based group for heatmap grouping later
msg_nodelist <- msg_nodelist %>%
  mutate(
    rooted_group = case_when(
      root > 30 ~ "Unicellular",
      root >= 21 & root <= 30 ~ "Multicellular non-vertebrate",
      root <= 20 ~ "Vertebrates"
    )
  )

# Build CUMULATIVE gene lists
genes_unicellular <- unique(msg_nodelist$queryItem[msg_nodelist$root > 30])
genes_multicellular <- unique(msg_nodelist$queryItem[msg_nodelist$root >= 21])
genes_vertebrates <- unique(msg_nodelist$queryItem)

cat("\n--- Cumulative Gene Counts ---\n")
cat(sprintf("  Unicellular (root > 30):  %d genes\n", length(genes_unicellular)))
cat(sprintf("  Multicellular (root >= 21): %d genes\n", length(genes_multicellular)))
cat(sprintf("  Vertebrates (all):          %d genes\n", length(genes_vertebrates)))
```

## GO Enrichment Analysis (Cumulative)

```{r go-enrichment}
#| cache: true

set.seed(42)

# Run enrichment
go_unicellular <- perform_go_enrichment(genes_unicellular, "Unicellular")
go_multicellular <- perform_go_enrichment(genes_multicellular, "Multicellular non-vertebrate")
go_vertebrates <- perform_go_enrichment(genes_vertebrates, "Vertebrates")

# Combine
go_combined <- bind_rows(go_unicellular, go_multicellular, go_vertebrates) %>% filter(!is.na(ID))

if (nrow(go_combined) == 0) stop("No enriched GO terms found.")

cat(sprintf("\nTotal enriched GO terms across all groups: %d\n", nrow(go_combined)))

# Save tables
vroom::vroom_write(go_combined, file.path(output_base, "tables/MSG_GO_cumulative_combined.csv"), delim = ",")
if (!is.null(go_unicellular)) vroom::vroom_write(go_unicellular, file.path(output_base, "tables/MSG_GO_cumulative_unicellular.csv"), delim = ",")
if (!is.null(go_multicellular)) vroom::vroom_write(go_multicellular, file.path(output_base, "tables/MSG_GO_cumulative_multicellular.csv"), delim = ",")
if (!is.null(go_vertebrates)) vroom::vroom_write(go_vertebrates, file.path(output_base, "tables/MSG_GO_cumulative_vertebrates.csv"), delim = ",")
```

## GO Visualization

```{r go-plots}
#| fig-width: 12
#| fig-height: 14

# Prepare top terms
go_top <- go_combined %>%
  group_by(evolutionary_group) %>%
  slice_min(p.adjust, n = 15, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(neg_log10_padj = -log10(p.adjust),
         evolutionary_group = factor(evolutionary_group, levels = c("Vertebrates", "Multicellular non-vertebrate", "Unicellular"))) %>%
  arrange(evolutionary_group, neg_log10_padj) %>%
  mutate(Description = factor(Description, levels = unique(Description)))

# Plot All Groups
p_go_all <- ggplot(go_top, aes(x = neg_log10_padj, y = Description, fill = evolutionary_group)) +
  geom_bar(stat = "identity", width = 0.7) +
  facet_grid(evolutionary_group ~ ., scales = "free_y", space = "free_y") +
  scale_fill_manual(values = group_colors, name = "Evolutionary Group") +
  labs(title = "GO Biological Process Enrichment — Cumulative Gene Sets",
       subtitle = "Top 15 terms per evolutionary group (all orthologs present up to each point)",
       x = expression(-log[10](p.adjust)), y = NULL) +
  theme_minimal(base_size = 12) +
  theme(strip.text.y = element_text(angle = 0, face = "bold", size = 11),
        legend.position = "none", panel.grid.major.y = element_blank())

print(p_go_all)
ggsave(file.path(output_base, "plots/MSG_GO_cumulative_barplot_all_groups.svg"), p_go_all, width = 12, height = 14)
ggsave(file.path(output_base, "plots/MSG_GO_cumulative_barplot_all_groups.png"), p_go_all, width = 12, height = 14, dpi = 300)

# Plot Vertebrates Only
go_vert <- go_combined %>%
  filter(evolutionary_group == "Vertebrates") %>%
  slice_min(p.adjust, n = 20, with_ties = FALSE) %>%
  mutate(neg_log10_padj = -log10(p.adjust)) %>%
  arrange(neg_log10_padj) %>%
  mutate(Description = factor(Description, levels = unique(Description)))

if (nrow(go_vert) > 0) {
  p_go_vert <- ggplot(go_vert, aes(x = neg_log10_padj, y = Description)) +
    geom_bar(stat = "identity", fill = group_colors["Vertebrates"], width = 0.7) +
    labs(title = "GO Biological Process Enrichment — Vertebrates (Cumulative)",
         subtitle = "Top 20 enriched terms (all MSG orthologs)",
         x = expression(-log[10](p.adjust)), y = NULL) +
    theme_minimal(base_size = 12) +
    theme(panel.grid.major.y = element_blank())
    
  print(p_go_vert)
  ggsave(file.path(output_base, "plots/MSG_GO_cumulative_barplot_vertebrates.svg"), p_go_vert, width = 10, height = 8)
  ggsave(file.path(output_base, "plots/MSG_GO_cumulative_barplot_vertebrates.png"), p_go_vert, width = 10, height = 8, dpi = 300)
}
```

## GO UpSet Plot

```{r go-upset}
#| fig-width: 10
#| fig-height: 7

go_terms_list <- list(
  "Unicellular" = if (!is.null(go_unicellular)) go_unicellular$ID else character(0),
  "Multicellular" = if (!is.null(go_multicellular)) go_multicellular$ID else character(0),
  "Vertebrates" = if (!is.null(go_vertebrates)) go_vertebrates$ID else character(0)
)
go_terms_list <- go_terms_list[sapply(go_terms_list, length) > 0]

if (length(go_terms_list) >= 2) {
  svglite(file.path(output_base, "plots/MSG_GO_cumulative_upset_plot.svg"), width = 10, height = 7)
  upset(fromList(go_terms_list), order.by = "freq", decreasing = TRUE, mb.ratio = c(0.6, 0.4),
        text.scale = 1.5, point.size = 3.5, line.size = 1.5,
        mainbar.y.label = "GO Terms Intersection", sets.x.label = "GO Terms per Group",
        sets.bar.color = c(group_colors["Unicellular"], group_colors["Multicellular non-vertebrate"], group_colors["Vertebrates"])[names(go_terms_list)])
  dev.off()
  
  upset(fromList(go_terms_list), order.by = "freq", sets.bar.color = "gray30")
}
```

## GO Semantic Similarity Clustering

```{r go-semantic}
#| cache: true

cat("--- Calculating GO Semantic Similarity ---\n")
go_terms <- unique(go_combined$ID)
semData <- godata(OrgDb = org.Hs.eg.db, ont = "BP")
mSim <- mgoSim(go_terms, go_terms, semData = semData, measure = "Wang", combine = NULL)
mSim[is.na(mSim)] <- 0
dist_matrix <- as.dist(1 - mSim)
hc <- hclust(dist_matrix, method = "average")
```

## TreeAndLeaf Visualization

```{r treeandleaf-viz}
#| cache: true

# Prepare Data
go_combined <- go_combined %>% mutate(path_length = as.integer(sapply(strsplit(GeneRatio, "/"), "[", 2)), ratio = Count / path_length)
go_term_info <- go_combined %>% group_by(ID) %>% slice_min(pvalue, n = 1, with_ties = FALSE) %>% ungroup() %>%
  dplyr::select(ID, Description, Count, ratio, pvalue, p.adjust, evolutionary_group)

# Node sizes
size <- go_term_info$Count
aux <- sort(unique(size))
names(aux) <- as.character(1:length(aux))
sizeRanks <- as.factor(names(aux[match(size, aux)]))
sizeIntervals <- min(7, length(unique(size)))
go_term_info$size <- (sizeBase + (as.numeric(sizeRanks) * sizeMultiplier))
go_term_info$group_color <- group_colors[go_term_info$evolutionary_group]

# Create TreeAndLeaf
tal <- treeAndLeaf(hc)
tal <- att.mapv(g = tal, dat = as.data.frame(go_term_info), refcol = 1)

# Visual Attributes
pal <- brewer.pal(9, "OrRd")
tal <- att.setv(g = tal, from = "Description", to = "nodeAlias")
tal <- att.setv(g = tal, from = "ratio", to = "nodeColor", cols = pal, nquant = 5)
tal <- att.setv(g = tal, from = "size", to = "nodeSize", xlim = c((sizeBase + sizeMultiplier), (sizeBase + (sizeMultiplier * sizeIntervals)), sizeMultiplier), nquant = sizeIntervals)
tal <- att.addv(tal, "nodeFontSize", value = 15, index = V(tal)$isLeaf)
tal <- att.adde(tal, "edgeWidth", value = 3)
tal <- att.adde(tal, "edgeColor", value = "gray80")

leaf_indices <- which(V(tal)$isLeaf)
for (i in leaf_indices) {
  node_name <- V(tal)$name[i]
  node_group <- go_term_info %>% filter(ID == node_name) %>% pull(evolutionary_group)
  V(tal)$nodeLineColor[i] <- ifelse(length(node_group) > 0, group_colors[node_group], "gray50")
}
tal <- att.addv(tal, "nodeLineWidth", value = 4, index = V(tal)$isLeaf)

# Interactive RedeR (run manually)
# rdp <- RedPort()
# calld(rdp)
# resetd(rdp)
# addGraph(rdp, tal, layout = NULL)
# relax(rdp, p1 = 21, p2 = 70, p3 = 255, p4 = 100, p5 = 50)
```

## KEGG Enrichment (Cumulative)

```{r kegg-enrichment}
#| cache: true

cat("--- Performing KEGG Enrichment (Cumulative) ---\n")
kegg_unicellular <- perform_kegg_enrichment(genes_unicellular, "Unicellular")
kegg_multicellular <- perform_kegg_enrichment(genes_multicellular, "Multicellular non-vertebrate")
kegg_vertebrates <- perform_kegg_enrichment(genes_vertebrates, "Vertebrates")

# Filter Human Diseases
kegg_unicellular <- filter_human_diseases(kegg_unicellular)
kegg_multicellular <- filter_human_diseases(kegg_multicellular)
kegg_vertebrates <- filter_human_diseases(kegg_vertebrates)

# Combine
kegg_combined <- bind_rows(kegg_unicellular, kegg_multicellular, kegg_vertebrates) %>% filter(!is.na(ID))

# Save
vroom::vroom_write(kegg_combined, file.path(output_base, "tables/MSG_KEGG_cumulative_combined.csv"), delim = ",")
if (!is.null(kegg_unicellular)) vroom::vroom_write(kegg_unicellular, file.path(output_base, "tables/MSG_KEGG_cumulative_unicellular.csv"), delim = ",")
if (!is.null(kegg_multicellular)) vroom::vroom_write(kegg_multicellular, file.path(output_base, "tables/MSG_KEGG_cumulative_multicellular.csv"), delim = ",")
if (!is.null(kegg_vertebrates)) vroom::vroom_write(kegg_vertebrates, file.path(output_base, "tables/MSG_KEGG_cumulative_vertebrates.csv"), delim = ",")
```

## KEGG Visualization

```{r kegg-plots}
#| fig-width: 12
#| fig-height: 10

if (nrow(kegg_combined) > 0) {
  kegg_top <- kegg_combined %>%
    group_by(evolutionary_group) %>%
    slice_min(p.adjust, n = 15, with_ties = FALSE) %>%
    ungroup() %>%
    mutate(neg_log10_padj = -log10(p.adjust),
           evolutionary_group = factor(evolutionary_group, levels = c("Vertebrates", "Multicellular non-vertebrate", "Unicellular"))) %>%
    arrange(evolutionary_group, neg_log10_padj) %>%
    mutate(Description = factor(Description, levels = unique(Description)))

  p_kegg_all <- ggplot(kegg_top, aes(x = neg_log10_padj, y = Description, fill = evolutionary_group)) +
    geom_bar(stat = "identity", width = 0.7) +
    facet_grid(evolutionary_group ~ ., scales = "free_y", space = "free_y") +
    scale_fill_manual(values = group_colors, name = "Evolutionary Group") +
    labs(title = "KEGG Pathway Enrichment — Cumulative Gene Sets",
         subtitle = "Top 15 pathways per group (Human Diseases excluded)",
         x = expression(-log[10](p.adjust)), y = NULL) +
    theme_minimal(base_size = 12) +
    theme(strip.text.y = element_text(angle = 0, face = "bold", size = 11),
          legend.position = "none", panel.grid.major.y = element_blank())

  print(p_kegg_all)
  ggsave(file.path(output_base, "plots/MSG_KEGG_cumulative_barplot_all_groups.svg"), p_kegg_all, width = 12, height = 10)
  ggsave(file.path(output_base, "plots/MSG_KEGG_cumulative_barplot_all_groups.png"), p_kegg_all, width = 12, height = 10, dpi = 300)
}
```

## KEGG UpSet Plot (Cumulative)

```{r kegg-upset}
#| fig-width: 10
#| fig-height: 7

kegg_terms_list <- list(
  "Unicellular" = if (!is.null(kegg_unicellular)) kegg_unicellular$ID else character(0),
  "Multicellular" = if (!is.null(kegg_multicellular)) kegg_multicellular$ID else character(0),
  "Vertebrates" = if (!is.null(kegg_vertebrates)) kegg_vertebrates$ID else character(0)
)
kegg_terms_list <- kegg_terms_list[sapply(kegg_terms_list, length) > 0]

if (length(kegg_terms_list) >= 2) {
  svglite(file.path(output_base, "plots/MSG_KEGG_cumulative_upset_plot.svg"), width = 10, height = 7)
  upset(fromList(kegg_terms_list), order.by = "freq", decreasing = TRUE, mb.ratio = c(0.6, 0.4),
        text.scale = 1.5, point.size = 3.5, line.size = 1.5,
        mainbar.y.label = "KEGG Pathways Intersection", sets.x.label = "Pathways per Group",
        sets.bar.color = c(group_colors["Unicellular"], group_colors["Multicellular non-vertebrate"], group_colors["Vertebrates"])[names(kegg_terms_list)])
  dev.off()
  
  upset(fromList(kegg_terms_list), order.by = "freq", sets.bar.color = "gray30")
}
```

## KEGG Clade-Level Heatmap (Cumulative)

```{r kegg-clade-loop}
#| cache: true

clade_info <- msg_nodelist %>%
  dplyr::select(clade_name, root) %>%
  distinct() %>%
  arrange(desc(root))
clade_order <- clade_info$clade_name

cat(sprintf("Performing KEGG enrichment for %d individual clades (cumulative)...\n", length(clade_order)))

kegg_per_clade <- list()
for (clade in clade_order) {
  clade_root <- clade_info$root[clade_info$clade_name == clade]
  cumulative_genes <- msg_nodelist %>% filter(root >= clade_root) %>% pull(queryItem) %>% unique()
  
  cat(sprintf("  %s (root=%d): %d cumulative genes... ", clade, clade_root, length(cumulative_genes)))
  result <- perform_kegg_clade(cumulative_genes, clade)
  
  if (!is.null(result)) {
    kegg_per_clade[[clade]] <- result
    cat(sprintf("%d pathways\n", nrow(result)))
  } else {
    cat("no enrichment\n")
  }
}
kegg_all_clades <- bind_rows(kegg_per_clade)
```

```{r kegg-heatmap}
#| fig-width: 16
#| fig-height: 20

if (nrow(kegg_all_clades) > 0) {
  all_pathways <- unique(kegg_all_clades$Description)
  clades_with_data <- clade_order[clade_order %in% unique(kegg_all_clades$clade)]
  
  heatmap_data <- expand.grid(Description = all_pathways, Clade = clades_with_data, stringsAsFactors = FALSE)
  enriched_pairs <- kegg_all_clades %>% dplyr::select("Description", "clade") %>% distinct() %>% mutate(enriched = TRUE) %>% dplyr::rename(Clade = clade)
  
  heatmap_data <- heatmap_data %>% left_join(enriched_pairs, by = c("Description", "Clade")) %>% mutate(enriched = ifelse(is.na(enriched), FALSE, enriched))
  
  clade_group <- clade_info %>% mutate(group = case_when(root > 30 ~ "Unicellular", root >= 21 & root <= 30 ~ "Multicellular non-vertebrate", root <= 20 ~ "Vertebrates"))
  heatmap_data <- heatmap_data %>% left_join(clade_group %>% dplyr::select("clade_name", "group"), by = c("Clade" = "clade_name"))
  
  heatmap_data <- heatmap_data %>% mutate(fill_color = case_when(!enriched ~ "Not enriched", group == "Unicellular" ~ "Unicellular", group == "Multicellular non-vertebrate" ~ "Multicellular non-vertebrate", group == "Vertebrates" ~ "Vertebrates"))
  
  heatmap_data$Clade <- factor(heatmap_data$Clade, levels = clades_with_data)
  heatmap_data$group <- factor(heatmap_data$group, levels = c("Unicellular", "Multicellular non-vertebrate", "Vertebrates"))
  
  pathway_matrix <- heatmap_data %>% mutate(val = as.integer(enriched)) %>% dplyr::select("Description", "Clade", "val") %>% pivot_wider(names_from = Clade, values_from = val, values_fill = 0)
  mat <- as.matrix(pathway_matrix[, -1])
  rownames(mat) <- pathway_matrix$Description
  
  if (nrow(mat) > 2) {
    hc_paths <- hclust(dist(mat, method = "binary"), method = "complete")
    pathway_order <- rownames(mat)[hc_paths$order]
  } else {
    pathway_order <- rownames(mat)
  }
  heatmap_data$Description <- factor(heatmap_data$Description, levels = pathway_order)
  
  fill_values <- c("Not enriched" = "white", "Unicellular" = group_colors["Unicellular"], "Multicellular non-vertebrate" = group_colors["Multicellular non-vertebrate"], "Vertebrates" = group_colors["Vertebrates"])
  
  p_heatmap <- ggplot(heatmap_data, aes(x = Clade, y = Description, fill = fill_color)) +
    geom_tile(color = "gray80", linewidth = 0.3) +
    facet_grid(. ~ group, scales = "free_x", space = "free_x") +
    scale_fill_manual(values = fill_values, name = "Evolutionary Group") +
    labs(title = "KEGG Pathway Enrichment Across Evolutionary Clades", subtitle = "Cumulative gene sets per clade | Human Diseases excluded", x = NULL, y = NULL) +
    theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 9), axis.text.y = element_text(size = 8), panel.grid = element_blank(), panel.spacing = unit(0.3, "lines"), strip.text.x = element_text(face = "bold", size = 10), legend.position = "none")
  
  print(p_heatmap)
  ggsave(file.path(output_base, "plots/MSG_KEGG_cumulative_clade_heatmap.svg"), p_heatmap, width = 16, height = 20)
  ggsave(file.path(output_base, "plots/MSG_KEGG_cumulative_clade_heatmap.png"), p_heatmap, width = 16, height = 20, dpi = 300)
}
```

## Save Outputs and Cleanup

```{r save-outputs}
#| cache: false

# Final check
if (file.exists(here("results/paper_data/MSG_GO_cumulative_combined.csv"))) {
  old_go <- vroom::vroom(here("results/paper_data/MSG_GO_cumulative_combined.csv"), show_col_types = FALSE)
  new_go <- vroom::vroom(file.path(output_base, "tables/MSG_GO_cumulative_combined.csv"), show_col_types = FALSE)
  cat("MSG_GO_cumulative_combined — old rows:", nrow(old_go), ", new rows:", nrow(new_go), "\n")
}

rm(mSim, dist_matrix, hc, tal, go_combined, kegg_combined, kegg_all_clades, layout, pathway_matrix, mat)
gc()
```
