---
title: "Cancer Gene Set Enrichment Analysis (TSG & Oncogene) with TreeAndLeaf"
author: "Gleison M. Azevedo"
format: html
execute:
  cache: true
  echo: true
  message: false
  warning: false
---

## Package Management

```{r pkg-install}
#| cache: false

# Automatically detect and install any missing packages before loading.
# Includes both CRAN and Bioconductor packages.
required_cran <- c(
  "tidyverse", "vroom", "here", "RColorBrewer", "classInt", "UpSetR", "svglite", "igraph", "scales", "ggtext"
)

required_bioc <- c(
  "TreeAndLeaf", "RedeR", "clusterProfiler", "org.Hs.eg.db", "GOSemSim"
)

# Install CRAN packages
missing_cran <- required_cran[
  !vapply(required_cran, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_cran) > 0) {
  install.packages(missing_cran, repos = "https://cloud.r-project.org")
}

# Install Bioconductor packages
missing_bioc <- required_bioc[
  !vapply(required_bioc, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_bioc) > 0) {
  if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager", repos = "https://cloud.r-project.org")
  BiocManager::install(missing_bioc, update = FALSE)
}

library(TreeAndLeaf)
library(RedeR)
library(igraph)
library(tidyverse)
library(here)
library(vroom)
library(RColorBrewer)
library(clusterProfiler)
library(org.Hs.eg.db)
library(GOSemSim)
library(classInt)
library(UpSetR)
library(svglite)
```

## Helper Functions and Configuration

```{r functions}
# Define Evolutionary Group Colors
group_colors <- c(
  "Unicellular" = "#E69F00",                # Orange
  "Multicellular non-vertebrate" = "#0072B2", # Blue
  "Vertebrates" = "#009E73"                  # Green
)

# Visualization sizing parameters
sizeMultiplier <- 15
sizeBase <- 50

# --- Function: GO Enrichment ---
perform_go_enrichment <- function(genes, group_name) {
  cat(sprintf("  Processing %s (%d genes)...\n", group_name, length(genes)))
  
  if (length(genes) == 0) return(NULL)

  # Convert gene symbols to Entrez IDs
  gene_ids <- suppressMessages(bitr(
    genes,
    fromType = "SYMBOL",
    toType = "ENTREZID",
    OrgDb = org.Hs.eg.db
  ))

  if (nrow(gene_ids) == 0) {
    cat(sprintf("    No genes mapped for %s\n", group_name))
    return(NULL)
  }

  # Perform GO enrichment
  ego <- enrichGO(
    gene = gene_ids$ENTREZID,
    OrgDb = org.Hs.eg.db,
    ont = "BP",
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05,
    qvalueCutoff = 0.05,
    readable = TRUE
  )

  if (is.null(ego) || nrow(as.data.frame(ego)) == 0) {
    cat(sprintf("    No enriched terms for %s\n", group_name))
    return(NULL)
  }

  result <- as.data.frame(ego) %>%
    mutate(evolutionary_group = group_name)

  cat(sprintf("    Found %d enriched GO terms\n", nrow(result)))
  return(result)
}

# --- Function: KEGG Enrichment ---
perform_kegg_enrichment <- function(genes, group_name) {
  cat(sprintf("  Processing %s (%d genes)...\n", group_name, length(genes)))
  
  if (length(genes) == 0) return(NULL)

  # Convert gene symbols to Entrez IDs
  gene_ids <- suppressMessages(bitr(
    genes,
    fromType = "SYMBOL",
    toType = "ENTREZID",
    OrgDb = org.Hs.eg.db
  ))

  if (nrow(gene_ids) == 0) {
    cat(sprintf("    No genes mapped for %s\n", group_name))
    return(NULL)
  }

  # Perform KEGG enrichment
  ekegg <- enrichKEGG(
    gene = gene_ids$ENTREZID,
    organism = "hsa",
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05,
    qvalueCutoff = 0.05
  )

  if (is.null(ekegg) || nrow(as.data.frame(ekegg)) == 0) {
    cat(sprintf("    No enriched KEGG pathways for %s\n", group_name))
    return(NULL)
  }

  # Convert to readable gene symbols
  ekegg <- setReadable(ekegg, OrgDb = org.Hs.eg.db, keyType = "ENTREZID")

  result <- as.data.frame(ekegg) %>%
    mutate(evolutionary_group = group_name)

  cat(sprintf("    Found %d enriched KEGG pathways\n", nrow(result)))
  return(result)
}

# --- Function: Filter Human Diseases ---
filter_human_diseases <- function(df) {
  if (is.null(df) || nrow(df) == 0) return(df)
  df_filtered <- df %>%
    filter(!str_detect(ID, "^hsa05"))
  cat(sprintf(
    "  Removed %d Human Disease pathways, %d remaining\n",
    nrow(df) - nrow(df_filtered), nrow(df_filtered)
  ))
  if (nrow(df_filtered) == 0) return(NULL)
  return(df_filtered)
}

# --- Function: Run full enrichment pipeline for a gene set ---
run_enrichment_pipeline <- function(gene_set_nodelist, set_name, set_prefix, output_base) {
  
  tables_dir <- file.path(output_base, "tables")
  plots_dir <- file.path(output_base, "plots")
  
  cat(sprintf("\n========================================\n"))
  cat(sprintf("  Running enrichment for: %s\n", set_name))
  cat(sprintf("  Total genes in nodelist: %d\n", nrow(gene_set_nodelist)))
  cat(sprintf("========================================\n"))
  
  # Assign Evolutionary Groups Based on Root Values
  gene_set_nodelist <- gene_set_nodelist %>%
    mutate(
      evolutionary_group = case_when(
        root >= 30 ~ "Unicellular",
        root >= 21 & root < 30 ~ "Multicellular non-vertebrate",
        root <= 20 ~ "Vertebrates"
      )
    ) %>%
    mutate(
      group_color = group_colors[evolutionary_group]
    )
  
  # Summary of genes per group
  cat("\n--- Gene Distribution by Evolutionary Group ---\n")
  print(table(gene_set_nodelist$evolutionary_group))
  
  # Extract unique genes per group
  genes_unicellular <- unique(gene_set_nodelist$queryItem[gene_set_nodelist$evolutionary_group == "Unicellular"])
  genes_multicellular <- unique(gene_set_nodelist$queryItem[gene_set_nodelist$evolutionary_group == "Multicellular non-vertebrate"])
  genes_vertebrates <- unique(gene_set_nodelist$queryItem[gene_set_nodelist$evolutionary_group == "Vertebrates"])
  
  ## ---- GO Enrichment Analysis ----
  cat("\n--- GO Enrichment Analysis ---\n")
  set.seed(42)
  
  go_unicellular <- perform_go_enrichment(genes_unicellular, "Unicellular")
  go_multicellular <- perform_go_enrichment(genes_multicellular, "Multicellular non-vertebrate")
  go_vertebrates <- perform_go_enrichment(genes_vertebrates, "Vertebrates")
  
  go_combined <- bind_rows(go_unicellular, go_multicellular, go_vertebrates) %>%
    filter(!is.na(ID))
  
  if (nrow(go_combined) == 0) {
    cat(sprintf("No enriched GO terms found for %s. Skipping GO analysis.\n", set_name))
  } else {
    cat(sprintf("\nTotal enriched GO terms across all groups: %d\n", nrow(go_combined)))
    
    # Save enrichment tables
    vroom::vroom_write(go_combined, file.path(tables_dir, paste0(set_prefix, "_GO_enrichment_combined.csv")), delim = ",")
    if (!is.null(go_unicellular)) vroom::vroom_write(go_unicellular, file.path(tables_dir, paste0(set_prefix, "_GO_enrichment_unicellular.csv")), delim = ",")
    if (!is.null(go_multicellular)) vroom::vroom_write(go_multicellular, file.path(tables_dir, paste0(set_prefix, "_GO_enrichment_multicellular.csv")), delim = ",")
    if (!is.null(go_vertebrates)) vroom::vroom_write(go_vertebrates, file.path(tables_dir, paste0(set_prefix, "_GO_enrichment_vertebrates.csv")), delim = ",")
  }
  
  ## ---- KEGG Enrichment Analysis ----
  cat("\n--- KEGG Enrichment Analysis ---\n")
  kegg_unicellular <- perform_kegg_enrichment(genes_unicellular, "Unicellular")
  kegg_multicellular <- perform_kegg_enrichment(genes_multicellular, "Multicellular non-vertebrate")
  kegg_vertebrates <- perform_kegg_enrichment(genes_vertebrates, "Vertebrates")
  
  kegg_unicellular <- filter_human_diseases(kegg_unicellular)
  kegg_multicellular <- filter_human_diseases(kegg_multicellular)
  kegg_vertebrates <- filter_human_diseases(kegg_vertebrates)
  
  kegg_combined <- bind_rows(kegg_unicellular, kegg_multicellular, kegg_vertebrates) %>%
    filter(!is.na(ID))
  
  # Save KEGG results
  vroom::vroom_write(kegg_combined, file.path(tables_dir, paste0(set_prefix, "_KEGG_enrichment_combined.csv")), delim = ",")
  if (!is.null(kegg_unicellular)) vroom::vroom_write(kegg_unicellular, file.path(tables_dir, paste0(set_prefix, "_KEGG_enrichment_unicellular.csv")), delim = ",")
  if (!is.null(kegg_multicellular)) vroom::vroom_write(kegg_multicellular, file.path(tables_dir, paste0(set_prefix, "_KEGG_enrichment_multicellular.csv")), delim = ",")
  if (!is.null(kegg_vertebrates)) vroom::vroom_write(kegg_vertebrates, file.path(tables_dir, paste0(set_prefix, "_KEGG_enrichment_vertebrates.csv")), delim = ",")
  
  # Export nodelist with group assignments
  gene_set_nodelist %>%
    vroom::vroom_write(file.path(tables_dir, paste0(set_prefix, "_nodelist_grouped.csv")), delim = ",")
  
  # Return all results for plotting
  list(
    nodelist = gene_set_nodelist,
    go_combined = go_combined,
    go_unicellular = go_unicellular,
    go_multicellular = go_multicellular,
    go_vertebrates = go_vertebrates,
    kegg_combined = kegg_combined,
    kegg_unicellular = kegg_unicellular,
    kegg_multicellular = kegg_multicellular,
    kegg_vertebrates = kegg_vertebrates
  )
}
```

## Data Loading

```{r data-load}
# Nodelist with rooting information and biological process annotations
nodelist <- vroom::vroom(
  here("results/orthology_data/nodelist.csv"),
  show_col_types = FALSE
)

# Cancer gene list
cancer_genes <- vroom::vroom(
  here("assets/cancerGeneList.tsv"),
  delim = "\t",
  show_col_types = FALSE
)

cat("Cancer gene list dimensions:", nrow(cancer_genes), "genes x", ncol(cancer_genes), "columns\n")
cat("\nGene Type distribution:\n")
print(table(cancer_genes$`Gene Type`, useNA = "ifany"))

# Create output directories
output_base <- here("results/enrichment_cancerGenes")
dir.create(file.path(output_base, "tables"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(output_base, "plots"), recursive = TRUE, showWarnings = FALSE)
```

## Build Cancer Gene Sets

Define two gene sets from the cancer gene list:

- **TSG gene set**: genes with `Gene Type` = `TSG` or `ONCOGENE_AND_TSG`
- **Oncogene gene set**: genes with `Gene Type` = `ONCOGENE` or `ONCOGENE_AND_TSG`

Note that `ONCOGENE_AND_TSG` genes are included in **both** sets.

```{r build-gene-sets}
# Build gene sets
tsg_genes <- cancer_genes %>%
  filter(`Gene Type` %in% c("TSG", "ONCOGENE_AND_TSG")) %>%
  pull(`Hugo Symbol`) %>%
  unique()

onc_genes <- cancer_genes %>%
  filter(`Gene Type` %in% c("ONCOGENE", "ONCOGENE_AND_TSG")) %>%
  pull(`Hugo Symbol`) %>%
  unique()

cat(sprintf("TSG gene set: %d genes (TSG + ONCOGENE_AND_TSG)\n", length(tsg_genes)))
cat(sprintf("Oncogene gene set: %d genes (ONCOGENE + ONCOGENE_AND_TSG)\n", length(onc_genes)))
cat(sprintf("Overlap (ONCOGENE_AND_TSG): %d genes\n", length(intersect(tsg_genes, onc_genes))))

# Filter nodelist to keep only genes in each cancer gene set
tsg_nodelist <- nodelist %>%
  filter(queryItem %in% tsg_genes) %>%
  dplyr::select(cog_id, root, clade_name, string_id, queryItem)

onc_nodelist <- nodelist %>%
  filter(queryItem %in% onc_genes) %>%
  dplyr::select(cog_id, root, clade_name, string_id, queryItem)

cat(sprintf("\nTSG genes found in nodelist: %d (unique: %d)\n", nrow(tsg_nodelist), length(unique(tsg_nodelist$queryItem))))
cat(sprintf("Oncogene genes found in nodelist: %d (unique: %d)\n", nrow(onc_nodelist), length(unique(onc_nodelist$queryItem))))
```

## TSG Enrichment Analysis

```{r tsg-enrichment}
#| cache: true

tsg_results <- run_enrichment_pipeline(tsg_nodelist, "Tumor Suppressor Genes (TSG)", "TSG", output_base)
```

## TSG GO Enrichment Visualization

```{r tsg-go-plots}
#| fig-width: 12
#| fig-height: 14

if (nrow(tsg_results$go_combined) > 0) {
  # Select top 15 GO terms per group by p.adjust
  go_top <- tsg_results$go_combined %>%
    group_by(evolutionary_group) %>%
    slice_min(p.adjust, n = 15, with_ties = FALSE) %>%
    ungroup() %>%
    mutate(
      neg_log10_padj = -log10(p.adjust),
      evolutionary_group = factor(
        evolutionary_group,
        levels = c("Vertebrates", "Multicellular non-vertebrate", "Unicellular")
      )
    ) %>%
    arrange(evolutionary_group, neg_log10_padj) %>%
    mutate(Description = factor(Description, levels = unique(Description)))
  
  p_go_all <- ggplot(go_top, aes(x = neg_log10_padj, y = Description, fill = evolutionary_group)) +
    geom_bar(stat = "identity", width = 0.7) +
    facet_grid(evolutionary_group ~ ., scales = "free_y", space = "free_y") +
    scale_fill_manual(values = group_colors, name = "Evolutionary Group") +
    labs(
      title = "GO Biological Process Enrichment — Tumor Suppressor Genes (TSG)",
      subtitle = "Top 15 terms per evolutionary group",
      x = expression(-log[10](p.adjust)),
      y = NULL
    ) +
    theme_minimal(base_size = 12) +
    theme(
      strip.text.y = element_text(angle = 0, face = "bold", size = 11),
      legend.position = "none",
      panel.grid.major.y = element_blank(),
      plot.title = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(color = "gray40")
    )
  
  print(p_go_all)
  ggsave(file.path(output_base, "plots/TSG_GO_barplot_all_groups.svg"), p_go_all, width = 12, height = 14)
  ggsave(file.path(output_base, "plots/TSG_GO_barplot_all_groups.png"), p_go_all, width = 12, height = 14, dpi = 300)
  
  # Plot: Vertebrates Only
  go_vert <- tsg_results$go_combined %>%
    filter(evolutionary_group == "Vertebrates") %>%
    slice_min(p.adjust, n = 20, with_ties = FALSE) %>%
    mutate(neg_log10_padj = -log10(p.adjust)) %>%
    arrange(neg_log10_padj) %>%
    mutate(Description = factor(Description, levels = unique(Description)))
  
  if (nrow(go_vert) > 0) {
    p_go_vert <- ggplot(go_vert, aes(x = neg_log10_padj, y = Description)) +
      geom_bar(stat = "identity", fill = group_colors["Vertebrates"], width = 0.7) +
      labs(
        title = "GO Biological Process Enrichment — Vertebrates TSG",
        subtitle = "Top 20 enriched terms",
        x = expression(-log[10](p.adjust)),
        y = NULL
      ) +
      theme_minimal(base_size = 12) +
      theme(
        panel.grid.major.y = element_blank(),
        plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(color = "gray40")
      )
    
    print(p_go_vert)
    ggsave(file.path(output_base, "plots/TSG_GO_barplot_vertebrates.svg"), p_go_vert, width = 10, height = 8)
    ggsave(file.path(output_base, "plots/TSG_GO_barplot_vertebrates.png"), p_go_vert, width = 10, height = 8, dpi = 300)
  }
}
```

## TSG GO Term Overlap (UpSet Plot)

```{r tsg-go-upset}
#| fig-width: 10
#| fig-height: 7

go_terms_list <- list(
  "Unicellular" = if (!is.null(tsg_results$go_unicellular)) tsg_results$go_unicellular$ID else character(0),
  "Multicellular" = if (!is.null(tsg_results$go_multicellular)) tsg_results$go_multicellular$ID else character(0),
  "Vertebrates" = if (!is.null(tsg_results$go_vertebrates)) tsg_results$go_vertebrates$ID else character(0)
)
go_terms_list <- go_terms_list[sapply(go_terms_list, length) > 0]

if (length(go_terms_list) >= 2) {
  svglite(file.path(output_base, "plots/TSG_GO_upset_plot.svg"), width = 10, height = 7)
  upset(
    fromList(go_terms_list),
    order.by = "freq",
    decreasing = TRUE,
    mb.ratio = c(0.6, 0.4),
    text.scale = 1.5,
    point.size = 3.5,
    line.size = 1.5,
    mainbar.y.label = "GO Terms Intersection",
    sets.x.label = "GO Terms per Group",
    sets.bar.color = c(
      group_colors["Unicellular"],
      group_colors["Multicellular non-vertebrate"],
      group_colors["Vertebrates"]
    )[names(go_terms_list)]
  )
  dev.off()
  
  upset(fromList(go_terms_list), order.by = "freq", sets.bar.color = "gray30")
}
```

## TSG GO Semantic Similarity Clustering

```{r tsg-go-semantic}
#| cache: true

if (nrow(tsg_results$go_combined) > 0) {
  cat("--- Calculating GO Semantic Similarity (TSG) ---\n")
  go_terms <- unique(tsg_results$go_combined$ID)
  cat(sprintf("Calculating similarity for %d unique GO terms...\n", length(go_terms)))
  
  semData <- godata(OrgDb = org.Hs.eg.db, ont = "BP")
  
  mSim <- mgoSim(go_terms, go_terms, semData = semData, measure = "Wang", combine = NULL)
  mSim[is.na(mSim)] <- 0
  dist_matrix <- as.dist(1 - mSim)
  hc_tsg <- hclust(dist_matrix, method = "average")
}
```

## TSG TreeAndLeaf Visualization

```{r tsg-treeandleaf}
#| cache: true

if (nrow(tsg_results$go_combined) > 0) {
  go_combined_tsg <- tsg_results$go_combined %>%
    mutate(
      path_length = as.integer(sapply(strsplit(GeneRatio, "/"), "[", 2)),
      ratio = Count / path_length
    )
  
  go_term_info <- go_combined_tsg %>%
    group_by(ID) %>%
    slice_min(pvalue, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    dplyr::select(ID, Description, Count, ratio, pvalue, p.adjust, evolutionary_group)
  
  size <- go_term_info$Count
  aux <- sort(unique(size))
  names(aux) <- as.character(1:length(aux))
  sizeRanks <- as.factor(names(aux[match(size, aux)]))
  sizeIntervals <- min(7, length(unique(size)))
  go_term_info$size <- (sizeBase + (as.numeric(sizeRanks) * sizeMultiplier))
  go_term_info$group_color <- group_colors[go_term_info$evolutionary_group]
  
  tal <- treeAndLeaf(hc_tsg)
  tal <- att.mapv(g = tal, dat = as.data.frame(go_term_info), refcol = 1)
  
  pal <- brewer.pal(9, "OrRd")
  tal <- att.setv(g = tal, from = "Description", to = "nodeAlias")
  tal <- att.setv(g = tal, from = "ratio", to = "nodeColor", cols = pal, nquant = 5)
  tal <- att.setv(g = tal, from = "size", to = "nodeSize",
                  xlim = c((sizeBase + sizeMultiplier), (sizeBase + (sizeMultiplier * sizeIntervals)), sizeMultiplier),
                  nquant = sizeIntervals)
  tal <- att.addv(tal, "nodeFontSize", value = 15, index = V(tal)$isLeaf)
  tal <- att.adde(tal, "edgeWidth", value = 3)
  tal <- att.adde(tal, "edgeColor", value = "gray80")
  
  leaf_indices <- which(V(tal)$isLeaf)
  for (i in leaf_indices) {
    node_name <- V(tal)$name[i]
    node_group <- go_term_info %>% filter(ID == node_name) %>% pull(evolutionary_group)
    if (length(node_group) > 0) {
      V(tal)$nodeLineColor[i] <- group_colors[node_group]
    } else {
      V(tal)$nodeLineColor[i] <- "gray50"
    }
  }
  tal <- att.addv(tal, "nodeLineWidth", value = 4, index = V(tal)$isLeaf)
  
  # Interactive RedeR Visualization (Run Manually)
  # rdp <- RedPort()
  # calld(rdp)
  # resetd(rdp)
  # addGraph(rdp, tal, layout = NULL)
  # relax(rdp, p1 = 21, p2 = 70, p3 = 255, p4 = 100, p5 = 50)
  
  rm(mSim, dist_matrix, hc_tsg, tal, go_combined_tsg)
}
```

## TSG KEGG Enrichment Visualization

```{r tsg-kegg-plots}
#| fig-width: 12
#| fig-height: 10

if (nrow(tsg_results$kegg_combined) > 0) {
  kegg_top <- tsg_results$kegg_combined %>%
    group_by(evolutionary_group) %>%
    slice_min(p.adjust, n = 15, with_ties = FALSE) %>%
    ungroup() %>%
    mutate(
      neg_log10_padj = -log10(p.adjust),
      evolutionary_group = factor(
        evolutionary_group,
        levels = c("Vertebrates", "Multicellular non-vertebrate", "Unicellular")
      )
    ) %>%
    arrange(evolutionary_group, neg_log10_padj) %>%
    mutate(Description = factor(Description, levels = unique(Description)))
  
  p_kegg_all <- ggplot(kegg_top, aes(x = neg_log10_padj, y = Description, fill = evolutionary_group)) +
    geom_bar(stat = "identity", width = 0.7) +
    facet_grid(evolutionary_group ~ ., scales = "free_y", space = "free_y") +
    scale_fill_manual(values = group_colors, name = "Evolutionary Group") +
    labs(
      title = "KEGG Pathway Enrichment — Tumor Suppressor Genes (TSG)",
      subtitle = "Top 15 pathways per evolutionary group (Human Diseases excluded)",
      x = expression(-log[10](p.adjust)),
      y = NULL
    ) +
    theme_minimal(base_size = 12) +
    theme(
      strip.text.y = element_text(angle = 0, face = "bold", size = 11),
      legend.position = "none",
      panel.grid.major.y = element_blank(),
      plot.title = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(color = "gray40")
    )
  
  print(p_kegg_all)
  ggsave(file.path(output_base, "plots/TSG_KEGG_barplot_all_groups.svg"), p_kegg_all, width = 12, height = 10)
  ggsave(file.path(output_base, "plots/TSG_KEGG_barplot_all_groups.png"), p_kegg_all, width = 12, height = 10, dpi = 300)
}
```

## TSG KEGG Pathway Overlap (UpSet Plot)

```{r tsg-kegg-upset}
#| fig-width: 10
#| fig-height: 7

kegg_terms_list <- list(
  "Unicellular" = if (!is.null(tsg_results$kegg_unicellular)) tsg_results$kegg_unicellular$ID else character(0),
  "Multicellular" = if (!is.null(tsg_results$kegg_multicellular)) tsg_results$kegg_multicellular$ID else character(0),
  "Vertebrates" = if (!is.null(tsg_results$kegg_vertebrates)) tsg_results$kegg_vertebrates$ID else character(0)
)
kegg_terms_list <- kegg_terms_list[sapply(kegg_terms_list, length) > 0]

if (length(kegg_terms_list) >= 2) {
  svglite(file.path(output_base, "plots/TSG_KEGG_upset_plot.svg"), width = 10, height = 7)
  upset(
    fromList(kegg_terms_list),
    order.by = "freq",
    decreasing = TRUE,
    mb.ratio = c(0.6, 0.4),
    text.scale = 1.5,
    point.size = 3.5,
    line.size = 1.5,
    mainbar.y.label = "KEGG Pathways Intersection",
    sets.x.label = "Pathways per Group",
    sets.bar.color = c(
      group_colors["Unicellular"],
      group_colors["Multicellular non-vertebrate"],
      group_colors["Vertebrates"]
    )[names(kegg_terms_list)]
  )
  dev.off()
  
  upset(fromList(kegg_terms_list), order.by = "freq", sets.bar.color = "gray30")
}
```

---

# Oncogene Enrichment Analysis

---

## Oncogene Enrichment Analysis

```{r onc-enrichment}
#| cache: true

onc_results <- run_enrichment_pipeline(onc_nodelist, "Oncogenes (ONC)", "ONC", output_base)
```

## Oncogene GO Enrichment Visualization

```{r onc-go-plots}
#| fig-width: 12
#| fig-height: 14

if (nrow(onc_results$go_combined) > 0) {
  go_top <- onc_results$go_combined %>%
    group_by(evolutionary_group) %>%
    slice_min(p.adjust, n = 15, with_ties = FALSE) %>%
    ungroup() %>%
    mutate(
      neg_log10_padj = -log10(p.adjust),
      evolutionary_group = factor(
        evolutionary_group,
        levels = c("Vertebrates", "Multicellular non-vertebrate", "Unicellular")
      )
    ) %>%
    arrange(evolutionary_group, neg_log10_padj) %>%
    mutate(Description = factor(Description, levels = unique(Description)))
  
  p_go_all <- ggplot(go_top, aes(x = neg_log10_padj, y = Description, fill = evolutionary_group)) +
    geom_bar(stat = "identity", width = 0.7) +
    facet_grid(evolutionary_group ~ ., scales = "free_y", space = "free_y") +
    scale_fill_manual(values = group_colors, name = "Evolutionary Group") +
    labs(
      title = "GO Biological Process Enrichment — Oncogenes (ONC)",
      subtitle = "Top 15 terms per evolutionary group",
      x = expression(-log[10](p.adjust)),
      y = NULL
    ) +
    theme_minimal(base_size = 12) +
    theme(
      strip.text.y = element_text(angle = 0, face = "bold", size = 11),
      legend.position = "none",
      panel.grid.major.y = element_blank(),
      plot.title = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(color = "gray40")
    )
  
  print(p_go_all)
  ggsave(file.path(output_base, "plots/ONC_GO_barplot_all_groups.svg"), p_go_all, width = 12, height = 14)
  ggsave(file.path(output_base, "plots/ONC_GO_barplot_all_groups.png"), p_go_all, width = 12, height = 14, dpi = 300)
  
  # Plot: Vertebrates Only
  go_vert <- onc_results$go_combined %>%
    filter(evolutionary_group == "Vertebrates") %>%
    slice_min(p.adjust, n = 20, with_ties = FALSE) %>%
    mutate(neg_log10_padj = -log10(p.adjust)) %>%
    arrange(neg_log10_padj) %>%
    mutate(Description = factor(Description, levels = unique(Description)))
  
  if (nrow(go_vert) > 0) {
    p_go_vert <- ggplot(go_vert, aes(x = neg_log10_padj, y = Description)) +
      geom_bar(stat = "identity", fill = group_colors["Vertebrates"], width = 0.7) +
      labs(
        title = "GO Biological Process Enrichment — Vertebrates Oncogenes",
        subtitle = "Top 20 enriched terms",
        x = expression(-log[10](p.adjust)),
        y = NULL
      ) +
      theme_minimal(base_size = 12) +
      theme(
        panel.grid.major.y = element_blank(),
        plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(color = "gray40")
      )
    
    print(p_go_vert)
    ggsave(file.path(output_base, "plots/ONC_GO_barplot_vertebrates.svg"), p_go_vert, width = 10, height = 8)
    ggsave(file.path(output_base, "plots/ONC_GO_barplot_vertebrates.png"), p_go_vert, width = 10, height = 8, dpi = 300)
  }
}
```

## Oncogene GO Term Overlap (UpSet Plot)

```{r onc-go-upset}
#| fig-width: 10
#| fig-height: 7

go_terms_list <- list(
  "Unicellular" = if (!is.null(onc_results$go_unicellular)) onc_results$go_unicellular$ID else character(0),
  "Multicellular" = if (!is.null(onc_results$go_multicellular)) onc_results$go_multicellular$ID else character(0),
  "Vertebrates" = if (!is.null(onc_results$go_vertebrates)) onc_results$go_vertebrates$ID else character(0)
)
go_terms_list <- go_terms_list[sapply(go_terms_list, length) > 0]

if (length(go_terms_list) >= 2) {
  svglite(file.path(output_base, "plots/ONC_GO_upset_plot.svg"), width = 10, height = 7)
  upset(
    fromList(go_terms_list),
    order.by = "freq",
    decreasing = TRUE,
    mb.ratio = c(0.6, 0.4),
    text.scale = 1.5,
    point.size = 3.5,
    line.size = 1.5,
    mainbar.y.label = "GO Terms Intersection",
    sets.x.label = "GO Terms per Group",
    sets.bar.color = c(
      group_colors["Unicellular"],
      group_colors["Multicellular non-vertebrate"],
      group_colors["Vertebrates"]
    )[names(go_terms_list)]
  )
  dev.off()
  
  upset(fromList(go_terms_list), order.by = "freq", sets.bar.color = "gray30")
}
```

## Oncogene GO Semantic Similarity Clustering

```{r onc-go-semantic}
#| cache: true

if (nrow(onc_results$go_combined) > 0) {
  cat("--- Calculating GO Semantic Similarity (Oncogenes) ---\n")
  go_terms <- unique(onc_results$go_combined$ID)
  cat(sprintf("Calculating similarity for %d unique GO terms...\n", length(go_terms)))
  
  semData <- godata(OrgDb = org.Hs.eg.db, ont = "BP")
  
  mSim <- mgoSim(go_terms, go_terms, semData = semData, measure = "Wang", combine = NULL)
  mSim[is.na(mSim)] <- 0
  dist_matrix <- as.dist(1 - mSim)
  hc_onc <- hclust(dist_matrix, method = "average")
}
```

## Oncogene TreeAndLeaf Visualization

```{r onc-treeandleaf}
#| cache: true

if (nrow(onc_results$go_combined) > 0) {
  go_combined_onc <- onc_results$go_combined %>%
    mutate(
      path_length = as.integer(sapply(strsplit(GeneRatio, "/"), "[", 2)),
      ratio = Count / path_length
    )
  
  go_term_info <- go_combined_onc %>%
    group_by(ID) %>%
    slice_min(pvalue, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    dplyr::select(ID, Description, Count, ratio, pvalue, p.adjust, evolutionary_group)
  
  size <- go_term_info$Count
  aux <- sort(unique(size))
  names(aux) <- as.character(1:length(aux))
  sizeRanks <- as.factor(names(aux[match(size, aux)]))
  sizeIntervals <- min(7, length(unique(size)))
  go_term_info$size <- (sizeBase + (as.numeric(sizeRanks) * sizeMultiplier))
  go_term_info$group_color <- group_colors[go_term_info$evolutionary_group]
  
  tal <- treeAndLeaf(hc_onc)
  tal <- att.mapv(g = tal, dat = as.data.frame(go_term_info), refcol = 1)
  
  pal <- brewer.pal(9, "OrRd")
  tal <- att.setv(g = tal, from = "Description", to = "nodeAlias")
  tal <- att.setv(g = tal, from = "ratio", to = "nodeColor", cols = pal, nquant = 5)
  tal <- att.setv(g = tal, from = "size", to = "nodeSize",
                  xlim = c((sizeBase + sizeMultiplier), (sizeBase + (sizeMultiplier * sizeIntervals)), sizeMultiplier),
                  nquant = sizeIntervals)
  tal <- att.addv(tal, "nodeFontSize", value = 15, index = V(tal)$isLeaf)
  tal <- att.adde(tal, "edgeWidth", value = 3)
  tal <- att.adde(tal, "edgeColor", value = "gray80")
  
  leaf_indices <- which(V(tal)$isLeaf)
  for (i in leaf_indices) {
    node_name <- V(tal)$name[i]
    node_group <- go_term_info %>% filter(ID == node_name) %>% pull(evolutionary_group)
    if (length(node_group) > 0) {
      V(tal)$nodeLineColor[i] <- group_colors[node_group]
    } else {
      V(tal)$nodeLineColor[i] <- "gray50"
    }
  }
  tal <- att.addv(tal, "nodeLineWidth", value = 4, index = V(tal)$isLeaf)
  
  # Interactive RedeR Visualization (Run Manually)
  # rdp <- RedPort()
  # calld(rdp)
  # resetd(rdp)
  # addGraph(rdp, tal, layout = NULL)
  # relax(rdp, p1 = 21, p2 = 70, p3 = 255, p4 = 100, p5 = 50)
  
  rm(mSim, dist_matrix, hc_onc, tal, go_combined_onc)
}
```

## Oncogene KEGG Enrichment Visualization

```{r onc-kegg-plots}
#| fig-width: 12
#| fig-height: 10

if (nrow(onc_results$kegg_combined) > 0) {
  kegg_top <- onc_results$kegg_combined %>%
    group_by(evolutionary_group) %>%
    slice_min(p.adjust, n = 15, with_ties = FALSE) %>%
    ungroup() %>%
    mutate(
      neg_log10_padj = -log10(p.adjust),
      evolutionary_group = factor(
        evolutionary_group,
        levels = c("Vertebrates", "Multicellular non-vertebrate", "Unicellular")
      )
    ) %>%
    arrange(evolutionary_group, neg_log10_padj) %>%
    mutate(Description = factor(Description, levels = unique(Description)))
  
  p_kegg_all <- ggplot(kegg_top, aes(x = neg_log10_padj, y = Description, fill = evolutionary_group)) +
    geom_bar(stat = "identity", width = 0.7) +
    facet_grid(evolutionary_group ~ ., scales = "free_y", space = "free_y") +
    scale_fill_manual(values = group_colors, name = "Evolutionary Group") +
    labs(
      title = "KEGG Pathway Enrichment — Oncogenes (ONC)",
      subtitle = "Top 15 pathways per evolutionary group (Human Diseases excluded)",
      x = expression(-log[10](p.adjust)),
      y = NULL
    ) +
    theme_minimal(base_size = 12) +
    theme(
      strip.text.y = element_text(angle = 0, face = "bold", size = 11),
      legend.position = "none",
      panel.grid.major.y = element_blank(),
      plot.title = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(color = "gray40")
    )
  
  print(p_kegg_all)
  ggsave(file.path(output_base, "plots/ONC_KEGG_barplot_all_groups.svg"), p_kegg_all, width = 12, height = 10)
  ggsave(file.path(output_base, "plots/ONC_KEGG_barplot_all_groups.png"), p_kegg_all, width = 12, height = 10, dpi = 300)
}
```

## Oncogene KEGG Pathway Overlap (UpSet Plot)

```{r onc-kegg-upset}
#| fig-width: 10
#| fig-height: 7

kegg_terms_list <- list(
  "Unicellular" = if (!is.null(onc_results$kegg_unicellular)) onc_results$kegg_unicellular$ID else character(0),
  "Multicellular" = if (!is.null(onc_results$kegg_multicellular)) onc_results$kegg_multicellular$ID else character(0),
  "Vertebrates" = if (!is.null(onc_results$kegg_vertebrates)) onc_results$kegg_vertebrates$ID else character(0)
)
kegg_terms_list <- kegg_terms_list[sapply(kegg_terms_list, length) > 0]

if (length(kegg_terms_list) >= 2) {
  svglite(file.path(output_base, "plots/ONC_KEGG_upset_plot.svg"), width = 10, height = 7)
  upset(
    fromList(kegg_terms_list),
    order.by = "freq",
    decreasing = TRUE,
    mb.ratio = c(0.6, 0.4),
    text.scale = 1.5,
    point.size = 3.5,
    line.size = 1.5,
    mainbar.y.label = "KEGG Pathways Intersection",
    sets.x.label = "Pathways per Group",
    sets.bar.color = c(
      group_colors["Unicellular"],
      group_colors["Multicellular non-vertebrate"],
      group_colors["Vertebrates"]
    )[names(kegg_terms_list)]
  )
  dev.off()
  
  upset(fromList(kegg_terms_list), order.by = "freq", sets.bar.color = "gray30")
}
```

## Save Outputs and Cleanup

```{r save-outputs}
#| cache: false

rm(tsg_results, onc_results, nodelist, cancer_genes, tsg_nodelist, onc_nodelist)
gc()
```
